### 2017-07-07 2017 JORMiCS/articleRamics.tex
fixed a mistake in chp 6.2

\documentclass{svproc}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\begin{document}

\title{Ampersand uses Relation Algebra as Programming Language}
\toctitle{Ampersand uses Relation Algebra as Programming Language}
%\titlerunning{Ampersand} % (for running head)
\author{Stef Joosten\inst{1,}\inst{2} (0000-0001-8308-0189)}
%orcid.org/0000-0001-8308-0189
\authorrunning{Stef Joosten} % (for running head)
\tocauthor{Stef Joosten} % for table of contents
\institute{Open Universiteit Nederland,\\  Postbus 2960,   6401 DL Heerlen, the Netherlands\\
\and Ordina NV,  Nieuwegein, the Netherlands\\
\email{stef.joosten@ou.nl} % optional, same for URL of homepage
} % end of address field
\maketitle

\abstract{Relation Algebra can be used as a programming language for building information systems.
	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
	As a case study, we have developed a database-application for legal reasoning.
	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
	Beside being declarative, relation algebra comes with attractive promises for developing big software.
	The compiler that was used for this case study, Ampersand, is the result of an open source project.
	Ampersand has been tried and tested in practice and is available as free open source software\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{Joosten2017}.
\keywords{relation algebra, software development, legal reasoning, information systems design, Ampersand, MirrorMe, big software}
	}

\section{Introduction}
\label{sct:Introduction}
	This paper investigates how relation algebra can be used as a programming language for information systems.
	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database-application.
	Ampersand is a syntactically sugared version of heterogeneous relation algebra \cite{Schmidt1997}.
	We present a case study to demonstrate programming in relation algebra and its impact on the software development process.
	The case study takes the reader by the hand in the thinking process of a software developer who programs with relation algebra.

	The use of relation algebra as a programming language is not new.
	It stands in the tradition of relation algebra~\cite{Maddux06}, logic programming~\cite{Lloyd1984},
	database application programming~\cite{Codd70}, and formal specification of software.
	Ampersand uses these ideas to compile relation algebra to working software, an idea which is also found in RelView~\cite{Berghammer2005} by Berghammer (Univ. of Kiel).
	Some differences with earlier programming languages are discussed in section~\ref{sct:Reflection}.

	The user may regard Ampersand as a programming language, especially suited for designing the back-end of information systems.
	The axioms of Tarski can be used to manipulate expressions in a way that preserves meaning~\cite{vdWoude2011}.
	This makes Ampersand a declarative language.

	Our case study shows an argument assistant for legal professionals, which was built as innovation project at Ordina.
	The purpose of this argument assistant is to support legal professionals in constructing a legal brief%
\footnote{A \define{brief} is a document that is meant to summarize a lawsuit for the judge and counterparty.
	It provides legal reasons for claims in a lawsuit based on regulations, precedents, and other legally acceptable sources.
	It shows how the reasoning applies to facts from the case.}.
	To the programmer, this offers the challenge to represent the problem of legal reasoning in relation algebra.
	This challenge is by no means simple, but outside the scope of this paper.
	In the Ampersand project at the Open University of the Netherlands,
	we are studying practical and didactical issues of software development in relation algebra~\cite{Michels2015}.

	The contribution of this paper is twofold.
	The case study contributes to the idea that programming can be done in relation algebra.
	The theory in section~\ref{sct:Code Fragments} contributes a theory to assist in preserving invariants.
	That theory has been consolidated in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/}~\cite{Joosten-AFP17}.
	All proofs upon which this paper is based have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.

	Section~\ref{sct:Ampersand} introduces Ampersand and its semantics.
	Section~\ref{sct:Conceptual analysis} introduces part of a theory of legal reasoning,
	which was developed for argument assistance.
	Section~\ref{sct:Programming in RA} discusses the programming mechanism in the application,
	and section~\ref{sct:Programming} visualizes that mechanism.
	Section~\ref{sct:Code Fragments} presents one way (of infinitely many) to derive code for arbitrary rules in relation algebra.
	Section~\ref{sct:Reflection} reflects on software development in Ampersand.
	It also provides an overview of the use of Ampersand in practice and an outlook to its further development.

\section{Ampersand}
\label{sct:Ampersand}
	In this section we explain the basics of Ampersand.
	The reader is expected to have sufficient background in relation algebra,
	in order to understand the remainder of this paper.

	The core of an Ampersand-script is a triple $\la\rules,\rels,\concepts\ra$,
	which consists of a set of rules $\rules$, relations $\rels$, and concepts $\concepts$.
	Ampersand-scripts are interpreted by the compiler as an information system.
	The rules constitute a theory in heterogeneous relation algebra.
	They constrain a body of data that resides in a database.

	The Ampersand-compiler generates a database from relations in the script.
	A database-application%
\footnote{Ampersand generates an application that consists of a relational database and interface components.
	Currently this application runs server-side on a PHP/MySQL platform and on a web-browser on the client-side.}
	assists users to keep rules satisfied throughout the lifetime of the database. It is also generated by Ampersand.

	A rule is an equality between two terms.
	Terms are built from relations.
	Ampersand interprets every relation as a finite set of pairs, which are stored in the database.
	The phase in which Ampersand takes a script, and turns it into a database, is what we will refer to as \define{compile-time}.
	The phase in which a user interacts with the database, is what we will refer to as \define{run-time}.
	At run-time, Ampersand can decide which rules are satisfied by querying the database.
	The compiler generates all software needed to keep rules satisfied at run-time.
	If a rule is not satisfied as a result of data that has changed, that change is reverted (rolled back) to
	preserve a state in which all rules are satisfied.
	Changes to the database are not specified by the software developer, but generated by the compiler.
	Rules in Ampersand are kept satisfied rather than executed directly.

	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of all atoms is called $\atoms$.
        Each atom is an instance of a \emph{concept}.

	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a person, and \atom{074238991} as a telephone number.
        We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a \in A$ means that atom $a$ is an instance of concept $A$.
	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
	in an Ampersand-script states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``An orange is a fruit that ....''.
	Specialization is not used the remainder of this paper.

	Relations (from set $\rels$) are used in information systems to store data.
%	A \define{fact} is a statement that is true in a business context.
%	Facts are stored and kept as data in a computer.
	As data changes over time, so do the contents of these relations.
	In this paper relations are represented by variables $r$, $s$, and $d$.
	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
	in which \id{nm} is a name and $A$ and $B$ are concepts.
	We call $A$ the source concept and $B$ the target concept of the relation.
	The pair $\pair{A}{B}$ is called the type of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.

    The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
	It maps each relation to a set of facts.
	Furthermore, it is a run-time requirement that the pairs in $r$ are contained in its type:
\begin{equation}
	\pair{a}{b}\in\ti{\declare{nm}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
\end{equation}

	Terms are used to combine relations using operators.
	The set of terms is called $\terms$. It is defined by:
\begin{definition}[terms]
\label{def:terms}
\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
\begin{eqnarray}
	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
\end{eqnarray}
\end{definition}
	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
	The \define{type} of a term $r$ is a pair of concepts, given by $\tf{r}$.
	$\mathfrak{T}$ is a partial function that maps terms to types.
	If a term has one type, it is called \define{type correct}.
	The Ampersand compiler requires all terms to be type correct.
	If no type can be assigned, the compiler gives an error message.
	If multiple types are assignable to a term, the ambiguity is signalled by the compiler.
	In that case it will also emit an error message and prompt the programmer to disambiguate the term by adding type information.
	In practice, the derivation of types can disambiguate most terms.
	So, the programmer experiences some freedom to denote distinct relations by the same name without the obligation to specify their types at all occurrences.
	The compiler does not generate any code if there is a term that is not type correct.
	The type function and the restrictions it suffers are discussed in~\cite{Joosten2015}.
	However, for the remainder of this paper this is irrelevant.

 	The meaning of terms in Ampersand is given by an interpretation function $\mathfrak{I}$.
	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
\begin{definition}[interpretation of terms]
\label{interpretation of terms}
\item   For every $A,B\in\concepts$ and $r,s\in\terms$
\begin{eqnarray}
	\ti{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
	\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\notin\ti{s}\}	\\
	\ti{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{b}{c}\in\ti{s}\}	\\
	\ti{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\ti{r}\ \text{implies}\ \pair{a}{c}\in\ti{s}\}	\\
	\ti{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\ti{s}\ \text{implies}\ \pair{a}{c}\in\ti{r}\}	\\
	\ti{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\ti{r}    \}	\\
	\ti{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
%	\ti{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
%	\ti{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
	\ti{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
	\ti{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
\end{eqnarray}
\end{definition}
%	The Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$) have been implemented partially in the current Ampersand implementation.

	A \define{rule} is a pair of terms $r,s\in\terms$ with $\tf{r}=\tf{s}$, which is syntactically recognizable as a rule.
\[\text{RULE}\ r = s\]
	This means \(\ti{r} = \ti{s}\). In practice, many rules are written as:
\[\text{RULE}\ r\subs s\]
	This is a shorthand for 
\[\text{RULE}\ r\cap s = r\]
	We have enhanced the type function $\mathfrak{T}$ and the interpretation function $\mathfrak{I}$ to cover rules as well.
	If $\tf{r}=\tf{s}$ and $\tf{s}=\pair{A}{B}$:
\begin{eqnarray}
	\tf{\text{RULE}\ r = s}   &=&\pair{A}{B}\\
	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}\\
	\ti{\text{RULE}\ r = s}   &=&\ti{\fullt{A}{B}-((s-r)\cup(r-s))}\\
	\ti{\text{RULE}\ r\subs s}&=&\ti{\fullt{A}{B}-(r-s)}
\end{eqnarray}

	We call rule $r$ \define{satisfied} when $\ti{\text{RULE}\ r = s}\ =\ \ti{\fullt{A}{B}}$.
	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
	We call a rule \define{violated} if it is not satisfied.
	The set $\ti{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.
	Consequently, a rule is sometimes called \define{invariant}.}.
	Each pair in the violation set of a rule is called a violation of that rule.

	The software developer must define how to resolve violations when they occur.
	She does so by inserting and/or deleting pairs in appropriately chosen relations.
	Whatever choice she makes, she must ensure that her code yields data that satisfies the rules.
	When we say: ``rule $r$ specifies this action'' we mean that satisfaction of rule $r$ is a postcondition of any action specified by rule $r$.

\section{Software for Legal Reasoning}
\label{sct:Conceptual analysis}
	As a case study, an argument assistance system, MirrorMe, was built.
	We have chosen to implement the ideas of Toulmin \cite{Toulmin1958},
	because his book ``The uses of Arguments'' is still one of the most influential works in the area of legal argument management%
\footnote{Thanks to Elleke Baecke for pointing us towards this source.}.
	Toulmin is regarded as the first scholar in modern history to come up with a usable theory of argumentation.
	Recent work typically draws on Toulmin, so his doctrine offers a good starting point.
	Toulmin's ideas have been implemented before, for example in a tool called ArguMed \cite{Verheij1999}.

	Software systems that support legal arguments have been around for many decades.
	Legal reasoning differs from logic reasoning because of the human judgments that are involved.
	Verheij~\cite{Verheij2003} distinguishes between argument assistance systems and automated reasoning systems.
	An automated reasoning system (ARS) mimics human reasoning in an attempt to automate a decision.
	An argument assisance system (AAS) helps its user to construct a decision.
	The literature on legal reasoning~\cite{Lind2007} makes it quite clear why mathematical logic alone does not suffice.
	Legal reasoning puts a human decision at the root of its notion of truth,
	whereas mathematical logic eradicates any human influence.
	Legal reasoning places human judgment in the core of the very notion of truth.
	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
	Argument assistance systems~\cite{Verheij2005} have been more successful,
	because they respect the professional freedom of legal professionals to construct their own line of argumentation.
	AASs offer help in many different ways.
	They can help by looking up legal references, jurisdiction from the past, scholarly works etc.
	They can also help to construct and validate arguments by keeping arguments and evidence organized.
	They can store, disclose and share legal evidence.

	In software development, logic predominates.
	We have used it to build an AAS.
	The argumentation principles of Toulmin are \emph{implemented in} logic rather than \emph{replaced by} logic.
	The structure of MirrorMe was designed by conceptually analysing ideas of Toulmin,
	such as claim, warrant, argument, and rebuttal. They appear in MirrorMe as relations.
	The Ampersand compiler generates a conceptual data model
	to help the software developer to oversee all relations.
	Even though our case study yields a model that is a bit too large for this paper,
	figure~\ref{fig:conceptual model} gives a good impression of what it looks like.
\begin{figure}[htb]
\begin{center}
  \includegraphics[scale=.23]{LogicalDataModel.pdf}
\end{center}
\caption{Conceptual data model}
\label{fig:conceptual model}
\end{figure}

	To convey the flavour of software development in relation algebra, it suffices to discuss a tiny part of the whole program.
	Therefore, we shall discuss the part of the conceptual model that is used in the sequel.
	The context in which a user creates arguments and reasons about them is a legal case.
	For this reason, validity, falsehood and truth of statements are related to the case.
	A \define{statement} is a phrase in natural language that can be either true or false.
	An example is \stmtText{The employee, John Brown, is entitled to 50 Euros}%
\footnote{This type of statement is typical for cases. It is valid only in the case where this particular John Brown is known.
	In other cases, where John Brown is unknown, this statement is meaningless}.
	In MirrorMe, a user can define a template, such as \stmtText{The employee, [emp], is entitled to [increase]}.
	The strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
	The reason for using placeholders is that legal rules are stated in general terms, e.g.
	\stmtText{Every employee is entitled to an increase in salary}.
	The user of MirrorMe will pick a legal text (from a source he trusts) and substitute parts of that text by placeholders.
	When the facts are known, the argument can be completed by substituting placeholders by actual phrases.

	It is precisely this substitution process that we have chosen to describe in this paper as a case study in
	programming with relation algebra.

\section{Programming in Relation Algebra}
\label{sct:Programming in RA}
	At this point we have reached the core of this case study.
	We focus on the substitution of placeholders when their values change.
	By focusing on this tiny detail, we can discuss the mechanics ``under the hood'' of the application generated by Ampersand.

	First we show how the computer solves the issue by looking at an exerpt of a log file (figure~\ref{fig:log file}).
	It shows an alternating sequence of the computer (ExecEngine) mentioning a rule,
	followed by insert or delete actions to satisfy that rule.
	Then we zoom in further, one rule at a time, to explain precisely what each rule looks like and how programming is done.
	We then discuss the same flow of events by means of a graph (figure~\ref{fig:event flow}) in which these rules and actions are nodes.
	This graph serves as an event flow diagram to illustrate the process behind figure~\ref{fig:log file}.

\begin{figure}[htb]
\begin{verbatim}
 ExecEngine run started
 ExecEngine satisfying rule 'signal phrase update'
 InsPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'flush substitutions'
 DelPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'reset statement text'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, [emp], is entitled to [increase].)
 ExecEngine satisfying rule 'done initializing'
 DelPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'substitute'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, James, is entitled to [increase].)
 InsPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'fill shownPhrase'
 InsPair(shownPhrase,Binding,Bind625,Phrase,James)
 ExecEngine run completed
\end{verbatim}
\caption{Log file of a substitution}
\label{fig:log file}
\end{figure}
	The log file of figure~\ref{fig:log file} has been taken from a computer that carries out the procedure to satisfy all rules.
	The machine will only act on rules that are violated.
	The first rule to be violated is ``signal phrase update'', in which the computer signals that something or someone has made a change
	in relation \id{phrase}.
	The procedure ends by doing the necessary substitutions,
	ensuring that all statements have the actual phrase of a placeholder in their text.

	Let us now study the actual rules to see how these rules cause the right actions to take place in the correct order.
	We will follow the log file from figure~\ref{fig:log file} in reverse direction,
	reasoning backwards from the result.

	Whether a placeholder has been edited can be observed by comparing its new phrase to the shown phrase%
\footnote{Note that we use the notion ``the phrase of a placeholder'' to indicate a pair from $\flip{\id{phrase}};\id{placeholder}$.}.
	The phrase of a placeholder is kept in the relation \id{phrase}.
	The shown phrase is kept in the relation \id{shownPhrase}.
	The sole purpose for having the relation \id{shownPhrase} is to detect a change in \id{phrase}.
	Let us introduce \id{differB} to represent the bindings with an updated phrase:
\[\id{differB} = \ident{Binding}\cap\id{shownPhrase};\cmpl{\ident{Phrase}};\flip{\id{phrase}}\]
	When the phrase of a placeholder changes,
	that phrase must be updated in every statement in which the placeholder was used.
	That update action is specified in section~\ref{rule: substitute}.
	Section~\ref{rule: fill shownPhrase} specifies how \id{shownPhrase} is made equal to \id{phrase},
	after all necessary substitutions are done.

\subsection{rule: fill shownPhrase}
\label{rule: fill shownPhrase}

\[\text{RULE}\ (\ident{Binding}\cap\id{substituted}/\id{inStatement} );\id{phrase}\ \subs\ \id{shownPhrase} \]
        This rule says that for each binding that has been substituted in every statement it is used in,
	the \id{phrase} must be equal to the \id{shownPhrase}.
	When violated, it is satisfied by inserting all violations into \id{shownPhrase}.

\subsection{rule: substitute}
\label{rule: substitute}
	Let us now look into the process of substituting placeholders by phrases.
	The relation \id{tmplParsedText} contains the original text, provided by a user.
	Placeholders are specified by enclosing them in brackets,
	e.g. \stmtText{The employee, [emp], is entitled to [increase]}.
	The text in which a placeholder has been substituted by a phrase,
	e.g. \stmtText{The employee, John Brown, is entitled to 50 Euros},
	is kept in relation \id{stmtShowText}.
	Each substitution that has been done in a statement corresponds to a binding-statement pair in the relation \id{substituted}.
	This relation keeps track of all substitutions.
	After a placeholder has been substituted, it no longer occurs in \id{stmtShowText}.
	This poses a problem if we want to substitute the new phrase in \id{stmtShowText}.
    For the placeholder that defines the place in the text where to substitute, is no longer in that text.
	Therefore, substitutions must be done in the original text of the statement.
	All placeholders in that text must then be substituted again.
	So, the text in \id{stmtShowText} must first be reset to the original text from \id{tmplParsedText}.
	To keep track of substitutions correctly, all corresponding binding-statement pairs must be removed from the relation \id{substituted}.
	Only after resetting is done, the substitutions can be put back in place with the new phrases filled in.

	We define a relation \id{resetS} to register the statements that are being reset.
    In statements that are not being reset, $\ident{Statement}-\id{resetS}$, substitutions can take place.
	All placeholders that have a binding with a phrase can be substituted. 
	The following rule specifies the action of substituting placeholders.
\begin{displaymath}
\begin{array}{rl}
\multicolumn{2}{l}{\text{RULE}}\\
&(\ident{Binding}\cap \id{phrase} ;\flip{\id{phrase}});\id{inStatement} ;(\ident{Statement} - \id{resetS})\\
\subs\\
&\id{substituted}
\end{array}
\end{displaymath}
	Violations of this rule are binding-statement pairs, of which the binding has a phrase and the statement is not being reset.
	Hence, this rule can be satisfied by inserting every violation into the relation substituted.

\subsection{rule: done initializing}
\label{rule: done initializing}
	Resetting a statement is done when two conditions are met.
	First, every statement that is (still) being reset may have no bindings in the relation \id{substituted}.
	Second, the text in \id{stmtShowText} corresponds to the text in \id{tmplParsedText}.
	So the rule that specifies the action is:

\begin{displaymath}
\begin{array}{rl}
\multicolumn{2}{l}{\text{RULE}}\\
&(\id{resetS} - \flip{\id{inStatement}};\id{substituted})\ \cap\\
&\id{template} ;\id{tmplParsedText} ;\flip{\id{stmtShowText}}\\
\subs\\
&\overline{\id{resetS}}
\end{array}
\end{displaymath}
	Violations of this rule are statements that are no longer being reset, but are still in the relation \id{resetS}.
	The appropriate action is to remove them from \id{resetS}.

\subsection{rule: reset statement text}
\label{rule: reset statement text}
	To satisfy one condition from section~\ref{rule: done initializing},
	the text in \id{stmtShowText} must be made equal to the original text in the template.
	The action is specified by the following rule:
\[\text{RULE}\ \id{resetS} ;\id{template} ;\id{descriptor}\ \subs\ \id{stmtShowText} \]
	Violations of this rule are descriptors of templates that belong to statements that are being reset.
	These violations can be resolved by inserting them in \id{stmtShowText}.

\subsection{rule: flush substitutions}
\label{rule: flush substitutions}
	To satisfy the other condition from section~\ref{rule: done initializing},
	the following rule specifies the action to be taken:
\[\text{RULE}\ \fullt{Binding}{Binding};\id{inStatement} ;\id{resetS}\ \subs\ \overline{\id{substituted} }\]
	Every binding in a statement that is being reset needs to be removed from the relation \id{substituted}.
	The software developer can implement this by deleting all violations of this rule from the relation \id{substituted}.

\subsection{rule: signal phrase update}
\label{rule: signal phrase update}
	When the phrase in a binding is edited and the new phrase differs from the shown phrase,
	this signals that substitutions must be flushed (see~\ref{rule: flush substitutions}),
	that the statement text must be reset to the original text (see~\ref{rule: reset statement text}),
	that the reset-state must be revoked (see~\ref{rule: done initializing}),
	that substitution must take place (see~\ref{rule: substitute}),
	and finally that the phrase detection is switched off again (see~\ref{rule: fill shownPhrase}).
	The initial condition occurs if a binding has been used (substituted) in a statement,
	and the binding satisfies \id{differB}.
	The following rule specifies the action that resetting a statment can start:
\[\text{RULE}\ \ident{Statement}\cap \flip{\id{inStatement}};\id{{\it differB}};\id{substituted}\ \subs\ \id{resetS} \]
	Violations of this rule are statements of which a substitution must be re-done.
	The software developer can have these violations added to \id{resetS} to satisfy this rule.
	In doing so, the chain of events is triggered that ends when all rules are satisfied.

\section{Programming in the Small}
\label{sct:Programming}
	Let us take a closer look at the programming process.
	Working in relation algebra, a software developer must think about satisfying constraints.
	She considers how violations arise from changing content of relations.
	And she thinks about insert and delete actions to resolve these violations.
	In our case study, we have reasoned with the event types: Ins $\la\id{relation}\ra$ and Del $\la\id{relation}\ra$%
\footnote{Ins and Del are called each others duals.}.
\begin{table}[htb]
{\small\begin{tabular}{|l|p{3.3in}|}\hline
rule&event types\\ \hline
signal phrase update&\small Ins \id{inStatement}, Ins \id{differB}, Ins \id{substituted}, Del \id{resetS}\\
flush substitutions&\small Ins \id{inStatement}, Ins \id{resetS}, Del \id{substituted}\\
reset statement text&\small Ins \id{resetS}, Ins \id{template}, Ins \id{descriptor}, Del \id{stmtShowText}\\
done initializing&\small Ins \id{resetS}, Del \id{inStatement}, Del \id{substituted}, Ins \id{template}, Ins \id{tmplParsedText}, Ins \id{stmtShowText}\\
substitute&\small Ins \id{phrase}, Ins \id{inStatement}, Del \id{resetS}, Del \id{substituted}\\
fill shownPhrase&\small Ins \id{substituted}, Del \id{inStatement}, Del \id{shownPhrase}, Ins \id{phrase}\\ \hline
\end{tabular}}
\caption{By which type of events can rules be violated?}
\label{fig:violation of rules}
\end{table}
	Table~\ref{fig:violation of rules} shows which event types may violate which rules.
	Recall section~\ref{sct:Programming in RA}, where the process of substitution started by updating the value of a placeholder.
	This meant doing an insert after a delete on the relation \id{phrase}, causing the updated phrase to appear in \id{differB}.
	That was signaled by the rule ``signal phrase update''.
\begin{figure}[bht]
\begin{center}
  \includegraphics[scale=.55]{eventFlow.png}
\end{center}
\caption{event flow}
\label{fig:event flow}
\end{figure}
	In general, the software developer must decide how to resolve violations that occur as a result of some event.
	This can be done by choosing among the duals of the event types from table~\ref{fig:violation of rules}.
	By systematically swapping Ins for Del and vice-versa,
	the table shows by which type of events violations can be resolved.
	To ensure progress, the software developer will pick a different relation.
	For resolving the violations in the same relation would resolve the violations by undoing the change.
        Typically, preserving invariance of a rule is done by changing (one or more) different relations than the one that caused the violations.
	The software developer can draw a graph that contains all information from table~\ref{fig:violation of rules}
	and the dual information.
	Figure~\ref{fig:event flow} shows part of that graph.
	A circle represents a rule, a rectangle represents an event type, and the arrows connect them.
	The software developer can work her way through the graph, to write code to resolve violations for every event type that might occur.
	Figure~\ref{fig:event flow} shows just that part of the graph that corresponds with the case study in this article.

\section{Code Fragments}
\label{sct:Code Fragments}
	After the previous section, one question remains open: which code can preserve invariance and can we produce that code for arbitrary rules?
	We shall provide an answer to this question not just for the case study at hand, but for all conceivable cases.
	This section derives such code for every term in relation algebra.
	The problem of generating such code suffers from arbitrariness: there are many ways to generate correct code.
	We have not explored all possibilities, because there are too many.
	We do give one implementation for every conceivable term, pointing out where choices are to be made.

	The purpose of the code is to preserve invariance of rules by insertion and deletion into and from relations.
	Inserting (a set of) pairs $\Delta$ into a relation $r$ yields $r\cup\Delta$.
	Likewise, deleting pairs $\Delta$ from a relation $r$ yields $r-\Delta$.
	In this section we derive code for computers,
	which is built around two basic operations: insert and delete.
	The insert operation can be represented in terms of pre- and postconditions\footnote{this idea
	was introduced by Floyd and Hoare~\cite{Floyd1967,Hoare1969}.
	It is well known for making assertions about computer programs and reasoning about them.}.
\[\begin{array}{r@{~}l}
&\{\text{pre: }\code{P}=p\}\\
\code{INSERT}&\Delta\code{ IN P}\\
&\{\text{post: }\code{P}=p\cup\Delta\}\\
\end{array}\]

	The delete operation is introduced similarly:
\[\begin{array}{r@{~}l}
&\{\text{pre: }\code{P}=p\}\\
\code{DELETE}&\Delta\code{ FROM P}\\
&\{\text{post: }\code{P}=p-\Delta\}\\
\end{array}\]

	Let the following example introduce the idea of breaking and restoring invariance.
	Suppose the rule $\code{P}\subs\code{S}\cup\code{T}$ is to be kept satisfied by a computer,
	and \code{P}, \code{S}, and \code{T} are variables (memory locations) in that computer,
	which represent relations.
	We assume that before executing any code, we have $\code{P}=p$, $\code{S}=s$, and $\code{T}=t$.
	The following example shows that inserting a set of pairs, $\Delta$, in \code{P}
	may violate the rule.
	Inserting the same $\Delta$ in \code{S} restores invariance.
	In terms of code, we write:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
\code{INSERT}&\Delta\code{ IN P}\\
&\{\text{$\code{P}=p\cup\Delta$, $\code{S}=s$, and $\code{T}=t$, and invariance is broken: $\code{P}\not\subs\code{S}\cup\code{T}$}\}\\
\code{INSERT}&\Delta\code{ IN S}\\
&\{\text{postcondition: $\code{P}=p\cup\Delta$, $\code{S}=s\cup\Delta$, $\code{T}=t$, and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
\end{array}\]

	To prove that rule $\code{P}\subs\code{S}\cup\code{T}$ is kept satisfied,
	we must show that its value afterwards is equal to $\full$,
	under the condition that its initial value equals $\full$.
	The initial value is $\cmpl{p}\cup s\cup t$, which we may assume.
	The final value is $(p\cup\Delta)\subs(s\cup\Delta)\cup t$, which must be proven equal to $\full$:
\[\begin{array}{cl}
&(p\cup\Delta)\subs(s\cup\Delta)\cup t\\
=&\\
&\cmpl{p\cup\Delta}\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
=&\hspace{1in}\{\hbox{De Morgan's law}\}\\
&(\cmpl{p}\cap\cmpl{\Delta})\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
=&\hspace{1in}\{\hbox{Distribute $\cup$ over $\cap$}\}\\
&(\cmpl{p}\cup s\cup\Delta\cup t)\ \cap\ (\cmpl{\Delta}\cup s\cup\Delta\cup t)\\
=&\hspace{1in}\{\cmpl{\Delta}\cup\Delta=\full\}\\
&\cmpl{p}\cup s\cup\Delta\cup t\\
=&\hspace{1in}\{\full\ =\ \cmpl{p}\cup s\cup t\}\\
&\full
\end{array}\]
	From this derivation we may conclude that invariance of $\code{P}\subs\code{S}\cup\code{T}$ is restored.
	From this example, we can see that $\code{INSERT }\Delta\code{ IN T}$ would have worked just as well.
	In general, there are alternative ways of preserving invariance.
	This calls for a systematic analysis.
	The following sections analyze the various terms in Relation Algebra and provides some code to restore invariance of rules containing such terms.
	We start with an analysis of rules, then we proceed with boolean operators, and finally we arrive at relational operators.

\subsection{Implication Rules}
\label{ssct:Implication Rules}
	We will start by studying insertions and deletions in the simplest of rules:
\begin{equation}
\code{RULE }r\subs s\label{rule:r subs s}
\end{equation}
	Suppose an event occurs, e.g. $\Delta_r$ is inserted in $r$,
	and a programmer must write code that preserves invariance of rule~\ref{rule:r subs s}.
	This requires choosing a value for $\Delta_s$, such that equation~\ref{eqn:r subsRS s} holds:
\begin{eqnarray}
r\subs s&\Rightarrow&r\cup\Delta_r\subs s\cup \Delta_s\label{eqn:r subsRS s}
\end{eqnarray}
	The programmer can choose $\Delta_s=\Delta_r$, and call it just $\Delta$,
	satisfying equation~\ref{eqn:r subs s} trivially by monotonicity of $\subs$:
\begin{eqnarray}
r\subs s&\Rightarrow&r\cup\Delta\subs s\cup \Delta\label{eqn:r subs s}
\end{eqnarray}
	A programmer can take the antecedent of equation~\ref{eqn:r subs s}, $r\subs s$, as precondition.
	The consequent, $r\cup\Delta\subs s\cup \Delta$, is the postcondition.
	She then writes the following code fragment:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
\code{INSERT}&\Delta\code{ INTO R};\\
&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
\code{INSERT}&\Delta\code{ INTO S}\\
&\{\text{postcondition: $\code{R}=r\cup\Delta$, $\code{S}=s\cup\Delta$, and $\code{R}\subs\code{S}$}\}
\end{array}\]
	The first step, $\code{INSERT }\Delta\code{ INTO R}$, represents the event that may violate invariance of $\code{R}\subs\code{S}$.
	The second step, $\code{INSERT }\Delta\code{ INTO S}$, represents the reaction.

	A similar line of reasoning can be used for deletion as well:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
\code{DELETE}&\Delta\code{ FROM S};\\
&\{\text{$\code{R}=r$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
\code{DELETE}&\Delta\code{ FROM R}\\
&\{\text{postcondition: $\code{R}=r-\Delta$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$}\}
\end{array}\]
	The proof is noteworthy similar and equally trivial:
\begin{eqnarray}
r\subs s&\Rightarrow&r-\Delta\subs s-\Delta\label{eqn:r subs s by deletion}
\end{eqnarray}
	If the event is a deletion of $\Delta$ from $r$ or an insertion into $s$, no reaction is required.
	Such events cannot violate the invariance of rule~\ref{rule:r subs s},
	because equations~\ref{eqn:r subs s delfrom r} and~\ref{eqn:r subs s insinto s} hold for any $\Delta$:
\begin{eqnarray}
r\subs s&\Rightarrow&r-\Delta\subs s\label{eqn:r subs s delfrom r}\\
r\subs s&\Rightarrow&r\subs s\cup\Delta\label{eqn:r subs s insinto s}
\end{eqnarray}

Summarizing, some code for preserving invariance of rules of form $\code{R}\subs\code{S}$ is:
\[\begin{array}{|l|l|l|l|}
\hline
\text{invariant}&\text{action}&\text{reaction}&\text{eqn.}\\ \hline
\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{INSERT }\Delta\code{ INTO S}&\ref{eqn:r subs s}\\
\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM S}&\code{DELETE }\Delta\code{ FROM R}&\ref{eqn:r subs s by deletion}\\
\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\ref{eqn:r subs s delfrom r}\\
\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO S}&\code{Nop}&\ref{eqn:r subs s insinto s}\\ \hline
\end{array}\]
Preserving an invariant of the form $\code{R}=\code{S}$ is equivalent to preserving both $\code{R}\subs\code{S}$ and $\code{S}\subs\code{R}$ simultaneously.

\subsection{Rules with Boolean Operators}
	To write code for boolean expressions,
	we revert to the following lemma's:
\begin{eqnarray}
r=\cmpl{s}&\Rightarrow&r\cup\Delta=\cmpl{s-\Delta}\label{lbl:Preserve r = -s with Ins Delta into r}\\
r=\cmpl{s}&\Rightarrow&r-\Delta=\cmpl{s\cup\Delta}\label{lbl:Preserve r = -s with Del Delta from r}\\
t=r\cap s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cap(s\cup\Delta)\label{lbl:Preserve t=r inter s with Ins Delta into t}\\
t=r\cup s&\Rightarrow&t-\Delta=(r-\Delta)\cup(s-\Delta)\label{lbl:Preserve t=r union s with Del Delta from t}\\
t=r\cap s&\Rightarrow&t-\Delta=(r-\Delta)\cap s\label{lbl:Preserve t=r inter s with Del Delta from t}\\
t=r\cup s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cup s\label{lbl:Preserve t=r union s with Ins Delta into t}
\end{eqnarray}
	In proof assistant Isabelle, the proof of each of these lemmas is made in zero steps, so we omit these proofs here.

	Equations~\ref{lbl:Preserve r = -s with Ins Delta into r} and~\ref{lbl:Preserve r = -s with Del Delta from r}
	show that insertion in a term means deletion from its complement and vice versa.
	So the programmer can write the following code:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
\code{INSERT}&\Delta\code{ INTO R};\\
&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
\code{DELETE}&\Delta\code{ FROM S}\\
&\{\text{$\code{R}=r\cup\Delta$ and $\code{S}=s-\Delta$, so by eqn.~\ref{lbl:Preserve r = -s with Ins Delta into r} we have $\code{R}=\cmpl{\code{S}}$}\}
\end{array}\]
	Likewise, the reaction on a deletion will be an insertion:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
\code{DELETE}&\Delta\code{ FROM R};\\
&\{\text{$\code{R}=r-\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
\code{INSERT}&\Delta\code{ INTO S}\\
&\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
\end{array}\]

	Now consider equations~\ref{lbl:Preserve t=r inter s with Ins Delta into t} and~\ref{lbl:Preserve t=r union s with Del Delta from t}.
	If \code{T} changes, both \code{R} and \code{S} must be changed accordingly.
	Since $\cup$ and $\cap$ are associative operators, the arguments can be evaluated with no interaction between them.
	This allows us to define a primitive, \code{ALL}, without specifying in which order it processes its arguments\footnote{... leaving room to take advantage of parallelism.}.
	We define \code{ALL} to perform all actions from a given list:
\[\begin{array}{r@{~}l}
&\{\text{Assume }\forall\code{A}\in\code{AS}: (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
\code{ALL}&\ \code{AS}\\
&\{\forall\code{A}: \id{post}_{\code{\tiny A}}\}
\end{array}\]
	With this primitive, we can write code for inserting pairs in \code{T},
	using equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t}:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
\code{INSERT}&\Delta\code{ INTO T};\\
&\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
\code{ALL}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
&\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
\end{array}\]
	Similarly, we can derive code for equation~\ref{lbl:Preserve t=r union s with Del Delta from t}:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
\code{DELETE}&\Delta\code{ FROM T};\\
&\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
\code{ALL}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
&\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
\end{array}\]

	In equations~\ref{lbl:Preserve t=r inter s with Del Delta from t} and~\ref{lbl:Preserve t=r union s with Ins Delta into t},
	a choice must be made between $r$ and $s$. For that purpose we need a primitive \code{CHC} that chooses one action from a list $\code{AS}$, say $\code{A}$:
\[\begin{array}{r@{~}l}
&\{ (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
\code{CHC}&\ \code{AS}\\
&\{\id{post}_{\code{\tiny A}}\}
\end{array}\]
	Note that this definition leaves open which term is selected.
	The following code shows how this works out for equation~\ref{lbl:Preserve t=r inter s with Del Delta from t}:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
\code{DELETE}&\Delta\code{ FROM T};\\
&\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
\code{CHC}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
&\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
\end{array}\]

	The following code shows how this works out for equation~\ref{lbl:Preserve t=r union s with Ins Delta into t}:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
\code{INSERT}&\Delta\code{ INTO T};\\
&\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
\code{CHC}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
&\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
\end{array}\]

	Section~\ref{ssct:Implication Rules} treated rules of the form $\code{R}\subs\code{S}$.
	Note that this rule can be rewritten to $\cmpl{\code{R}}\cup\code{S}=\full$.
	Consequently, rules did not have to be treated separately.
	We could have treated them as a special case of equations with boolean operators only.
	In fact, every rule with boolean operators $\cup$, $\cap$, and $-$ can be rewritten as a number of rules in disjunctive normal form\footnote{%
	Normalizing a boolean algebra is existing knowledge, and is considered outside the scope of this paper.} (DNF~\cite{davey1990}).
	We can represent a rule in DNF informally by:
\begin{eqnarray}
\code{RULE }\cmpl{r_0}\cup\ldots\cup\cmpl{r_n}\cup s_0\cup\ldots\cup s_m
\end{eqnarray}
	or more formally by
\begin{eqnarray}
\full=\cmpl{\bigcap\id{antecedents}}\cup\bigcup\id{consequents}
\end{eqnarray}
	in which \id{antecedents} are the complemented (negative) terms and \id{consequents} are the non-complemented (positive) terms.

	Summarizing, we have obtained code to preserve invariants in a boolean algebra:
\[\begin{array}{|l|l|l|l|}
\hline
\text{invariant}&\text{action}&\text{reaction}&\text{eqn.}\\ \hline
\code{R}=\cmpl{\code{S}}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE }\Delta\code{ FROM S}&\ref{lbl:Preserve r = -s with Ins Delta into r}\\
\code{R}=\cmpl{\code{S}}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT }\Delta\code{ INTO S}&\ref{lbl:Preserve r = -s with Del Delta from r}\\
\code{T}=\bigcap\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{ALL }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Ins Delta into t}\\
\code{T}=\bigcup\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{ALL }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Del Delta from t}\\
\code{T}=\bigcap\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{CHC }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Del Delta from t}\\
\code{T}=\bigcup\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{CHC }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Ins Delta into t}\\ \hline
\end{array}\]
	Having treated the boolean operators, let us consider code that is needed when relational operators are involved.

\subsection{Rules with relational operators}
	This section studies how to insert pairs into a term $r;s$.
	To formulate this precisely, we introduce an auxiliary relation $t$ to represent the term $r;s$.
	First we will insert pairs in $t$ while preserving invariance of $t=r;s$
	Then we will insert pairs in $r$ preserving the same rule.

	To insert pairs $\Delta$ into $t$ involves finding relations $\Delta_r$ and $\Delta_s$
	that satisfy equation~\ref{eqn:t subs r;s}.
\begin{equation}
t=r;s\ \Rightarrow\ t\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:t subs r;s}
\end{equation}
	Let us define a software primitive to compute both $\Delta_r$ and $\Delta_s$ and call it \code{NEWPAIRS}.
	It computes $\Delta_r$ and $\Delta_s$ from $r$, $s$, and $\Delta$.
	We can derive its postconditions from equation~\ref{eqn:t subs r;s} by means of the following derivation:
\[\begin{array}{r@{~}l}
&t\cup\Delta\\
=&\hspace{1in}\{\text{since }t\cup\Delta=t\cup(\Delta-t)\}\\
&t\cup(\Delta-t)\\
=&\hspace{1in}\{t=r;s\text{ and let \code{NEWPAIRS} ensure that }\Delta_r;\Delta_s=\Delta-t\}\\
&r;s\cup\Delta_r;\Delta_s\\
=&\hspace{1in}\{\text{let \code{NEWPAIRS} ensure that }r;\Delta_s\subs t\cup\Delta\text{ and }\Delta_r;s\subs t\cup\Delta\}\\
&r;s\cup r;\Delta_s\cup \Delta_r;s\cup \Delta_r;\Delta_s\\
=&\hspace{1in}\{\text{distributivity}\}\\
&(r\cup \Delta_r);(s\cup \Delta_s)
\end{array}\]
	The conditions in this derivation must be satisfied by \code{NEWPAIRS}. 
	This yields the following specification of \code{NEWPAIRS}:
\[\begin{array}{r@{~}l}
&\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{DELTA}=\Delta\}\\
\code{NEWPAIRS}&(\code{R}, \code{S}, \code{DELTA}): (\code{DELTAR}, \code{DELTAS})\\
&\{\code{DELTAR};\code{DELTAS}=\Delta\text{ and }\code{R};\code{DELTAS}\subs r;s\cup\Delta\text{ and }\code{DELTAR};\code{S}\subs r;s\cup\Delta\}
\end{array}\]
	The pre- and postconditions of \code{NEWPAIRS} are sufficient to satisfy equation~\ref{eqn:t subs r;s}.
	Since $t$ does not occur in the pre- and postconditions, \code{NEWPAIRS} implements equation~\ref{eqn:ins delta r;s}:
\begin{equation}
r;s\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:ins delta r;s}
\end{equation}
	Note that the specification of \code{NEWPAIRS} does not determine its implementation uniquely.
	There are multiple ways to implement \code{NEWPAIRS}.
	If \code{NEWPAIRS} creates one new pair for every pair in $\Delta-t$,
	we get $r;\Delta_s=\emptyset$ and $\Delta_r;s=\emptyset$.
	The weaker condition from the derivation, i.e. $r;\Delta_s\subs t\cup\Delta$ and $\Delta_r;s\subs t\cup\Delta$ allows some reuse of existing pairs.
	However, any implementation that preserves invariance of equation~\ref{eqn:ins delta r;s} will do.

	At this point we have the ingredients for code that preserves invariance of $\code{T}=\code{R};\code{S}$
\[\begin{array}{l}
\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\text{ and }\code{T}=\code{R};\code{S}\}\\
\code{INSERT }\Delta\code{ INTO T};\\
\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}\neq\code{R};\code{S}\}\\
(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R}, \code{S}, \Delta-\code{T});\\
\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{ and }\\
\hspace{0.575in}\code{DELTAR};\code{DELTAS}=\Delta-\code{T}\text{ and }\\
\hspace{0.575in}\code{R};\code{DELTAS}\subs\code{T}\text{ and }\code{DELTAR};\code{S}\subs\code{T}\text{, so equation~\ref{eqn:ins delta r;s} is satisfied}\}\\
\code{INSERT DELTAR INTO R};\\
\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\}\\
\code{INSERT DELTAS INTO S}\\
\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\cup\code{DELTAS}\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}=\code{R};\code{S}\}
\end{array}\]
    This shows how to satisfy invariant $\code{T}=\code{R};\code{S}$ when pairs $\Delta$ are inserted into \code{T}.

	Let us now study how to keep $r;s$ constant when $r$ changes.
	Inserting pairs $\Delta$ into $r$ involves deleting pairs from $s$ such that:
\begin{equation}
(r\cup \Delta);(s-\Delta_s)=r;s\label{eqn:ins r del s}
\end{equation}
	In order to solve equation~\ref{eqn:ins r del s} towards $\Delta_s$,
	we derive\footnote{with thanks to Peter Jipsen, who pointed out this line of reasoning.}:
\begin{eqnarray}
&&(r\cup \Delta);(s-\Delta_s)\ \subs\ r;s\label{eqn:rs subs t by ins r del s}\\
&\Leftrightarrow&\nonumber\\
&&s-\Delta_s\ \subs\ (r\cup \Delta)\backslash(r;s)\nonumber\\
&\Leftrightarrow&\hspace{1in}\{\text{apply }x-y=z\ \Leftrightarrow\ x=z\cup y\text{ twice.}\}\nonumber\\
&&s-(r\cup \Delta)\backslash(r;s)\ \subs\ \Delta_s\nonumber
\end{eqnarray}
	This derivation shows that every solution $\Delta_s$ contains $s-(r\cup \Delta)\backslash(r;s)$.
	So the latter, $s-(r\cup \Delta)\backslash(r;s)$, is the smallest solution.

	By using the absorption property
\begin{equation}
s-(s\cap x)=s-x\label{absorb x}
\end{equation}
	it is obvious that $(r\cup \Delta)\backslash(r;s)$ is a solution too.
	Even though this solution may contain more pairs, it is equivalent to the minimal solution.

	The smallest solution might not contain all pairs from $r;s$.
	The missing pairs are $r;s-(r\cup\Delta);(s-(r\cup \Delta)\backslash(r;s))$.
	Therefore it may be necessary to reintroduce these missing pairs afterwards.
	Fortunately, we have just shown how to do just that with \code{NEWPAIRS}.
	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
\[\begin{array}{l}
\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
\code{INSERT }\Delta\code{ INTO R};\\
\hspace{0.5in}\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}\neq\code{T}\}\\
\code{DELETE R}\backslash\code{T FROM S};\\
\hspace{0.5in}\{\text{Since }\code{R}=r\cup\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r\cup \Delta)\backslash(r;s)\\
\hspace{0.575in}\text{So }\code{S}=s-(r\cup \Delta)\backslash(r;s)\text{ and therefore}\\
\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}\\
\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
\end{array}\]

	Let us now study how to delete pairs $\Delta$ from $r$ while keeping $r;s$ constant.
	This involves inserting pairs into $s$ such that:
\begin{equation}
(r-\Delta);(s\cup\Delta_s)=r;s\label{eqn:del r ins s}
\end{equation}
	The largest solution of equation~\ref{eqn:del r ins s} follows from the law $x;(x\backslash y)\subs y$:
\begin{eqnarray}
&&x;(x\backslash y)\ \subs\ y\nonumber\\
&\Leftrightarrow&\hspace{1in}\{\text{Let }x=r-\Delta\text{ and }y=r;s)\}\nonumber\\
&&(r-\Delta);((r-\Delta)\backslash(r;s))\ \subs\ r;s\nonumber\\
&\Leftrightarrow&\hspace{1in}\{\text{Let }\Delta_s\ =\ (r-\Delta)\backslash(r;s)\}\nonumber\\
&&(r-\Delta);\Delta_s\ \subs\ r;s\nonumber\\
&\Leftrightarrow&\hspace{1in}\{p;x\subs p;q\ \Leftrightarrow\ p;(q\cup x)\subs p;q\}\nonumber\\
&&(r-\Delta);(s\cup\Delta_s)\ \subs\ r;s\nonumber
\end{eqnarray}
	This derivation shows that $(r-\Delta)\backslash(r;s)$, is the largest solution.

	Once again, we might not have all pairs from $r;s$,
	so eqn.~\ref{eqn:del r ins s} needs to be completed with the missing pairs.
	The missing pairs are $r;s-(r-\Delta);(s\cup(r-\Delta)\backslash(r;s))$.
	In this case too, these missing pairs can be reintroduced afterwards.
	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
\[\begin{array}{l}
\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
\code{DELETE }\Delta\code{ FROM R};\\
\hspace{0.5in}\{\code{R}=r-\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so possibly }\code{R};\code{S}\neq\code{T}\}\\
\code{INSERT R}\backslash\code{T INTO S};\\
\hspace{0.5in}\{\text{Since }\code{R}=r-\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r-\Delta)\backslash(r;s)\\
\hspace{0.575in}\text{So }\code{S}=s\cup(r-\Delta)\backslash(r;s)\text{ and therefore}\\
\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}\\
\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
\end{array}\]

	Summarizing, we have obtained some more code to preserve invariants in a relation algebra:
\[\begin{array}{|l|l|l|l|}
\hline
\text{invariants}&\text{action}&\text{reaction}&\text{eqn.}\\ \hline
\begin{array}{l}
\code{R};\code{S}\subs\code{T}\\
\code{S}\subs\code{R}\backslash\code{T}\\
\code{R}\subs\code{T}\slash\code{S}
\end{array}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE R}\backslash\code{T FROM S}&\ref{eqn:rs subs t by ins r del s}\\ \hline
\begin{array}{l}
\code{R};\code{S}\subs\code{T}\\
\code{S}\subs\code{R}\backslash\code{T}\\
\code{R}\subs\code{T}\slash\code{S}
\end{array}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\\ \hline
\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO T}&
\begin{array}{l}
(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \Delta-\code{T});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}
\end{array}&\ref{eqn:ins delta r;s}\\ \hline
\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM T}&\code{Nop}&\\ \hline
\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{Nop}&\\ \hline
\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT R}\backslash\code{T INTO S}&\\ \hline
\code{R};\code{S}=\code{T}&\code{INSERT}\ \Delta\code{ INTO R}&
\begin{array}{l}
\code{DELETE R}\backslash\code{T FROM S};\\
(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}
\end{array}&\ref{eqn:ins r del s}\\ \hline
\code{R};\code{S}=\code{T}&\code{DELETE}\ \Delta\code{ FROM R}&
\begin{array}{l}
\code{INSERT R}\backslash\code{T INTO S};\\
(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}
\end{array}&\ref{eqn:del r ins s}\\ \hline
\end{array}\]
	The solutions given here are far from being exhaustive.
	We illustrate this point by giving an example.
	Suppose we have $\code{T}\subs\code{R};\code{S}$ and action $\code{DELETE}\ \Delta\code{ FROM R}$.
	So the question is which pairs must be added to $s$ after pairs $\Delta$ have been removed from $r$.
	In normal circumstances, we might just look it up in the table above, picking $\code{INSERT R}\backslash\code{T INTO S}$ as a reaction.
	In the circumstance that initially $\code{T}=r;s$, we might reason differently.
	We must still find a smallest value for $\Delta_s$ that satisfies:
\begin{equation}
r;s\ \subs\ (r-\Delta);(s\cup\Delta_s)\label{lbl:nonshrinking r;s when Del from r}
\end{equation}
	The following derivation shows that $\flip{(r-\Delta)};(r\cap\Delta);s$ satisfies this requirement,
	but on one condition: for every pair in $r;s$ there must be a pair remaining behind in $r-\Delta$.
	Here is the derivation:
\begin{eqnarray}
&&\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ \Delta_s\nonumber\\
&\Rightarrow\nonumber\\
&&(r-\Delta);\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Rightarrow&\hspace{1in}\{r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}\}\nonumber\\
&&(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Leftrightarrow\nonumber\\
&&(r-(r-\Delta));s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Rightarrow\nonumber\\
&&(r;s)-((r-\Delta);s)\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Leftrightarrow\nonumber\\
&&r;s\ \subs\ (r-\Delta);s \cup (r-\Delta);\Delta_s\nonumber\\
&\Leftrightarrow\nonumber\\
&&r;s\ \subs\ (r-\Delta);(s \cup \Delta_s)\nonumber
\end{eqnarray}
	This derivation justifies the reaction $\code{INSERT }\flip{(r-\Delta)};(r\cap\Delta);s\code{ INTO S}$,
	but only if $r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}$.

\section{Reflection}
\label{sct:Reflection}
	Ampersand is built on the belief that software development should be automated.
	It uses rules to represent requirements, in the firm belief that consistent requirements are essential to the software development process~\cite{Boehm1981}.
	Being fully aware of the fate of formal methods in computer science, Ampersand is founded on the belief that software development must be done more formally.
	So Ampersand is building further on the foundations laid by formal specification methods
	such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
	In contrast however with such methods, Ampersand is equipped with a software generator that generates an information system.
	Thus, specifying in Ampersand is developing software at the same time.
	Ampersand complements the RelView approach~\cite{Berghammer2005}, which also generates software but is stronger in specifying complex computational problems.

	In retrospect we see that programming in relation algebra yields an unconventional programming experience.
	This experience consists of inventing rules and choosing event types for resolving violations, as illustrated by the case study.
	To relate this experience to programming as we know it, section~\ref{sct:Comparison} makes a comparison with established programming languages (prior art).
	Section~\ref{sct:Contribution} summarizes the contributions to software development claimed by Ampersand.
	Section~\ref{sct:Ampersand in practice} summarizes the use of Ampersand in practice and
	section~\ref{sct:Further Research} gives an outlook on research that is required in the near future.

\subsection{Comparison}
\label{sct:Comparison}
	This section compares Ampersand with existing programming paradigms by mentioning the most important differences and similarities.

	The first to compare with is the imperative programming paradigm, known from popular languages such as Java~\cite{Java} and C++~\cite{Stroustrup97a}.
	Our case study cannot be called imperative, because the notion of control flow in imperative languages is very different.
	In imperative languages, the control flow is defined by the software developer.
	In this case study, the control flow emerges as a result of changes in relation content as illustrated
	by figure~\ref{fig:event flow}.

	The case study also differs from the logic programming paradigm of Prolog~\cite{Lloyd1984} and of all rule engines that can be considered to be an offspring of Prolog.
	A difference lies in the way rules are interpreted.
	In logic programming, a program consists of Horn-clauses and a resolution-proof is constructed on run-time.
	The software developer works with notions such as backtracking (backward chaining) and unification, which are absent in Ampersand.
	Ampersand is not restricted to Horn-clauses; any relation-algebraic equation over relations can be used as a rule.

	The case study also differs from functional programming, of which Haskell and Scala are prominent representants.
	A core idea in functional programming is to evaluate a program as a function from input to output~\cite{Backus1978}.
	A functional program consists of function definitions, that are evaluated by a term- or graph-rewriter, using various strategies such as lazy or eager evaluation.
	In contrast, one might interpret our case study as a relaxation of the constraint that everything is a function.
	In relation algebra, everything is a relation and a function is a restricted form of a relation.
	A similarity to functional programming is the declarative style, because substitution of equal terms without changing the semantics is a property we see in both worlds.

	A difference with database programming is found in the type of algebra that is used.
	Relational databases are founded on relational algebra~\cite{Codd70}. They are typically programmed in SQL.
	In contrast with database programming, Ampersand implements heterogeneous relation algebra~\cite{Schmidt1997}.
	A software developer working with relational algebra sees n-ary tables, while Ampersand is restricted to binary relations.
	The comparison between relation{\it al} algebra and relation algebra might relate to comparing relational databases with graph databases~\cite{Vicknair2010},
	although no literature was found to corroborate this.

	In the tradition of formal specification, there are many relational approaches, such as Z~\cite{Z}, CSP~\cite{CSP}, LOTOS~\cite{LOTOS}, VDM~\cite{VDM}.
	Where formal specification techniques typically analyse and diagnose specifications,
	Ampersand actually synthesizes (generates) information systems.

	If Ampersand represents a programming style at all, we might call it a relation-algebraic style of programming or maybe constraint programming\footnote{https://en.wikipedia.org/wiki/Constraint\_programming}.
	Ampersand's style would be characterized by a programmer who specifies constraints and a computer trying to satisfy these constraints by resolving violations.

\subsection{Contribution}
\label{sct:Contribution}
	Contributions of Ampersand to the software development process are:
\begin{itemize}
\item	Ampersand has the usual benefits of a declarative language:
	This means that terms can be manipulated by Tarski's axioms without changing their semantics~\cite{vdWoude2011}.
	It also means that the order in which rules are written has no consequence for their semantics.
\item	Heterogeneous relation algebra has a straightforward interpretation in natural language~\cite{RBD}.
	We have used that to formalize business requirements without exposing business stakeholders to any formal notation.
\item	Heterogeneous relation algebra in Ampersand is statically typed~\cite{Joosten2015}.
	There is much evidence for significantly lower software maintenance cost due to static typing as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
\item	Heterogeneous relation algebra is well studied.
	As a consequence, many tools that are readily available in the public domain can be put to good use.
	For executives of large organizations it can be reassuring that the formalism is free of childhood diseases.
\item	Relation algebra facilitates composing software from reusable components, because a program consists of rules.
	Since the union of sets of rules is a set of rules, compositionality comes from the union operator.
	In practice, when components are brought together in larger assemblies, hardly any adjustments have to be made%
\footnote{This is (unsubstantiated) experience collected from projects we have done with Ampersand.}.
\end{itemize}

	Ampersand also has disadvantages. It appears to be difficult to learn for large groups of software professionals.
	Research~\cite{Michels2015} shows that this is largely due to deficits in prerequisite knowledge, especially skills in discrete mathematics.
%	For this reason, we have made an addendum to the course we teach at the Open University of the Netherlands.
	Also, programming appears to be difficult in practice.

\subsection{Ampersand in Practice}
\label{sct:Ampersand in practice}
	Ampersand has been used in practice both in education (Open University of the Netherlands)
	and in industry (Ordina and TNO).
	For example, Ordina designed a proof-of-concept in 2007 of the INDIGO-system,
	which is still in use today as the core information system of the Dutch immigration authority, IND.
	The design of INDIGO was made with Ampersand, to obtain correct, detailed results in the least amount of time.
	More recently, Ampersand was used to design an information system called DTV for the Dutch food authority, NVWA.
	A prototype of DTV was built by Ampersand and was used as a model to build the actual system.
	TNO, a major Dutch industrial research laboratory, is using Ampersand for research purposes.
	For example, TNO did a study of international standardizations efforts such as
	RBAC (Role Based Access Control) in 2003 and architecture (IEEE 1471-2000)~\cite{IEEE1471} in 2004.
	Several inconsistencies were found in the last (draft) RBAC standard~\cite{RBAC}.
	TNO has also used the technique in conceiving several patents%
\footnote{e.g. patents DE60218042D, WO2006126875, EP1727327, WO2004046848, EP1563361, NL1023394C, EP1420323, WO03007571, and NL1013450C.}.
	At the Open University of the Netherlands, Ampersand is being taught in a course called Rule Based Design~\cite{RBD}.
	In this course, students use a platform called RAP, which has been built in Ampersand~\cite{Michels2015}.
	RAP has been the first Ampersand-application that has run in production.

\subsection{Further Research}
\label{sct:Further Research}
	Further research on this topic is required to bring relation algebra still closer to the community of practitioners.
	Further use of relation algebra can be made by incorporating a model checker, such as the Alloy analyser~\cite{Alloy2006}, to detect inconsistent rules.
	An exciting new development is Amperspiegel~\cite{Amperspiegel}, which brings notational flexibility at the fingertips of the user.
	Developments in the Ampersand-compiler are going towards a rule-repository (written in Ampersand itself).
	This will make collaborative information systems development in Ampersand easier, because the repository can assist in automating
	the software development process further.
	Other research is needed towards a comprehensive theory of information systems.
	Currently, there is no theory (in the mathematical meaning of the word) for information systems.
	In the Ampersand project, a sub-project called ``Formal Ampersand'' is being conducted to achieve this goal.
\bibliographystyle{spmpsci}
\bibliography{doc}

\end{document}

### 2017-07-11 2017 JLAMP/articleRamics.tex
process some remarks from Wolfram

--- 2017-07-07 2017 JORMiCS/articleRamics.tex
@@
 \begin{equation}
-s-(s\cap x)=s-x\label{absorb x}
+s-(s-x)=s\cap x\label{absorb x}
 \end{equation}
-	it is obvious that $(r\cup \Delta)\backslash(r;s)$ is a solution too.
-	Even though this solution may contain more pairs, it is equivalent to the minimal solution.
+	it is obvious that $(r\cup \Delta);(s\cap(r\cup \Delta)\backslash(r;s))\ \subs\ r;s$ is equivalent to equation~\ref{eqn:ins r del s}. Therefore, it yields an equivalent solution.
 
@@
 	In the Ampersand project, a sub-project called ``Formal Ampersand'' is being conducted to achieve this goal.
+\subsection{Acknowledgements}
+\label{sct:Acknowledgements}
+	Firstly, I wish to thank Sebastiaan Joosten for introducing me to Isabelle and for inspiring me with his challenging critiques.
+	I owe many thanks to Wolfram Kahl, Michiel Stornebrink, Han Joosten, and the anonymous reviewers of RAMiCS for their useful feedback.
+	And I am grateful to the Open University and Ordina for funding this research.
 \bibliographystyle{spmpsci}

### 2017-07-11 2017 JLAMP/articleRamics.tex
Adapt to Elsevier style (elsarticle.cls) for JLAMP

@@
-\documentclass{svproc}
+\documentclass{elsarticle}
 \usepackage{graphicx}
@@
 \title{Ampersand uses Relation Algebra as Programming Language}
-\toctitle{Ampersand uses Relation Algebra as Programming Language}
-%\titlerunning{Ampersand} % (for running head)
-\author{Stef Joosten\inst{1,}\inst{2} (0000-0001-8308-0189)}
+\author[ou,ordina]{Stef Joosten (0000-0001-8308-0189)}
 %orcid.org/0000-0001-8308-0189
-\authorrunning{Stef Joosten} % (for running head)
-\tocauthor{Stef Joosten} % for table of contents
-\institute{Open Universiteit Nederland,\\  Postbus 2960,   6401 DL Heerlen, the Netherlands\\
-\and Ordina NV,  Nieuwegein, the Netherlands\\
-\email{stef.joosten@ou.nl} % optional, same for URL of homepage
-} % end of address field
+\ead{stef.joosten@ou.nl}
+\address[ou]{Open Universiteit Nederland, Postbus 2960, 6401 DL Heerlen, the Netherlands}
+\address[ordina]{Ordina NV,  Nieuwegein, the Netherlands}
 \maketitle
 
-\abstract{Relation Algebra can be used as a programming language for building information systems.
+\begin{abstract}
+	Relation Algebra can be used as a programming language for building information systems.
 	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
@@
 	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{Joosten2017}.
-\keywords{relation algebra, software development, legal reasoning, information systems design, Ampersand, MirrorMe, big software}
-	}
+%\keywords{relation algebra, software development, legal reasoning, information systems design, Ampersand, MirrorMe, big software}
+\end{abstract}
 
@@
 	I owe many thanks to Wolfram Kahl, Michiel Stornebrink, Han Joosten, and the anonymous reviewers of RAMiCS for their useful feedback.
-	And I am grateful to the Open University and Ordina for funding this research.
-\bibliographystyle{spmpsci}
+	And I am very grateful to the Open University and Ordina for funding this research.
+
+\section{Bibliography}
+\bibliographystyle{elsarticle-harv}
 \bibliography{doc}

### 2017-07-13 2017 JLAMP/articleRamics.tex
changes for Elsevier format

@@
 \title{Ampersand uses Relation Algebra as Programming Language}
-\author[ou,ordina]{Stef Joosten (0000-0001-8308-0189)}
-%orcid.org/0000-0001-8308-0189
+\author[ou,ordina]{Stef Joosten\fnref{fn1}}
 \ead{stef.joosten@ou.nl}
@@
 \address[ordina]{Ordina NV,  Nieuwegein, the Netherlands}
-\maketitle
+\fntext[fn1]{ORCID 0000-0001-8308-0189}
 
@@
 	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{Joosten2017}.
-%\keywords{relation algebra, software development, legal reasoning, information systems design, Ampersand, MirrorMe, big software}
 \end{abstract}
+
+\begin{keyword}
+relation algebra\sep software development\sep legal reasoning\sep information systems design\sep Ampersand\sep MirrorMe\sep big software
+\end{keyword}
+\maketitle
 
@@
 	It maps each relation to a set of facts.
-	Furthermore, it is a run-time requirement that the pairs in $r$ are contained in its type:
+	The type system of Ampersand guarantees that each pair in $r$ is always contained in its type:
 \begin{equation}
@@
 \end{equation}
+	The type system is strong, which means that every atom has a type.
+	The type system is static, which means that type checking is entirely done at compile-time.
 
@@
 	The compiler does not generate any code if there is a term that is not type correct.
-	The type function and the restrictions it suffers are discussed in~\cite{Joosten2015}.
-	However, for the remainder of this paper this is irrelevant.
+	The type function and its restrictions are discussed in~\cite{Joosten2015}
+	and have no consequences for the remainder of this paper.
 

### 2017-07-20 2017 JLAMP/articleRamics.tex
commit for @sjcjoosten, to look at issue #1

@@
 	Ampersand uses these ideas to compile relation algebra to working software, an idea which is also found in RelView~\cite{Berghammer2005} by Berghammer (Univ. of Kiel).
-	Some differences with earlier programming languages are discussed in section~\ref{sct:Reflection}.
+ 	Ampersand lets the user specify constraints only and generates an information system that respects these constraints.
+	RelView also generates software but uses relational algebra as operators in an otherwise imperative programming language.
+	This makes RelView stronger in specifying complex computational problems.
+	Some more differences with earlier programming languages are discussed in section~\ref{sct:Comparison}.
+
+	The author shares the belief that a provably correct implementation of requirements are essential to the software development process~\cite{Boehm1981}.
+	Being fully aware of the fate of formal methods in computer science,
+        Ampersand is nevertheless founded on relation algebra.
+	It builds further on foundations laid by formal specification methods
+	such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
+	In contrast however with such methods, Ampersand is equipped with a software generator that generates an information system.
+	Thus, specifying in Ampersand is developing software at the same time.
 
@@
 
-	By using the absorption property
-\begin{equation}
-s-(s-x)=s\cap x\label{absorb x}
-\end{equation}
-	it is obvious that $(r\cup \Delta);(s\cap(r\cup \Delta)\backslash(r;s))\ \subs\ r;s$ is equivalent to equation~\ref{eqn:ins r del s}. Therefore, it yields an equivalent solution.
-
 	The smallest solution might not contain all pairs from $r;s$.
@@
 \label{sct:Reflection}
-	Ampersand is built on the belief that software development should be automated.
-	It uses rules to represent requirements, in the firm belief that consistent requirements are essential to the software development process~\cite{Boehm1981}.
-	Being fully aware of the fate of formal methods in computer science, Ampersand is founded on the belief that software development must be done more formally.
-	So Ampersand is building further on the foundations laid by formal specification methods
-	such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
-	In contrast however with such methods, Ampersand is equipped with a software generator that generates an information system.
-	Thus, specifying in Ampersand is developing software at the same time.
-	Ampersand complements the RelView approach~\cite{Berghammer2005}, which also generates software but is stronger in specifying complex computational problems.
-
-	In retrospect we see that programming in relation algebra yields an unconventional programming experience.
+	Ampersand is being built on the belief that software development should be automated.
+	The holy grail is to derive deterministic ways to preserve invariants, to allow for programming by constraints.
+	This is motivated by the benefits of describing a result in terms of constraints,
+	rather than having to invent all the steps that lead to that result.
+	For example, describing the situation that any financial transaction must preserve the total amount of money is simpler
+	than having to describe all the steps involved in such a transaction.
+	The attraction is not only for the programmer, but even more so for auditors and users for who the constraints are easier to grasp and validate.
+	This paper shows how a computer can derive certain steps,
+	helping the programmer to invent code an providing mathematical assurance of correctness (compliance-by-design).
+
+	The analysis in this paper shows once more that inventing algorithms is far from trivial.
+	Algorithms cannot be derived from constraints alone. The choices that remain must be resolved first.
+	This paper does show, however, that these choices occur in predictable ways.
+	That gives opportunities, for instance to make these choices explicit to a user, to optimize the number of choices,
+	or to use external knowledge to resolve the remaining choices on runtime.
+
+	In practice we experience that programming in relation algebra yields an unconventional way of programming.
 	This experience consists of inventing rules and choosing event types for resolving violations, as illustrated by the case study.

### 2017-07-20 2017 JLAMP/articleRamics.tex
Fixed antiquotations, so this file produces valid LaTeX.

@@
 	Ampersand has been tried and tested in practice and is available as free open source software\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
-	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{Joosten2017}.
+	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}.
 \end{abstract}
@@
 
+	The similarity with the reactive programming paradigm~\cite{Bainomugisha2013} is obvious from figure~\ref{fig:event flow}.
+	This graph is part of the program graph as conceived in reactive programs.
+	The code fragments described in section~\ref{sct:Code Fragments}
+	might be generated towards a mature reactive programming language (e.g. ELM)
+	and executed directly.
+	Ampersand lets you think in constraints and (reactive) code fragments to preserve invariants can be generated with the theory in this paper.
+	Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
+	but generation to a reactive programming language (e.g. ELM) is definitely on the wishlist.
+
 	If Ampersand represents a programming style at all, we might call it a relation-algebraic style of programming or maybe constraint programming\footnote{https://en.wikipedia.org/wiki/Constraint\_programming}.

### 2018-01-20 2017 JLAMP/articleRamics.tex
new changes before review

@@
 	The theory in section~\ref{sct:Code Fragments} contributes a theory to assist in preserving invariants.
-	That theory has been consolidated in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/}~\cite{Joosten-AFP17}.
-	All proofs upon which this paper is based have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
+	All lemmas in this paper have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
 
@@
 \end{eqnarray}
-	In proof assistant Isabelle, the proof of each of these lemmas is made in zero steps, so we omit these proofs here.
+	In proof assistant Isabelle, these lemmas are proven in zero steps, so we omit these proofs here.
 
@@
 	The attraction is not only for the programmer, but even more so for auditors and users for who the constraints are easier to grasp and validate.
-	This paper shows how a computer can derive certain steps,
-	helping the programmer to invent code an providing mathematical assurance of correctness (compliance-by-design).
-
-	The analysis in this paper shows once more that inventing algorithms is far from trivial.
+	In this paper we have shown how to derive specific steps,
+	helping the programmer to invent code and providing mathematical assurance of correctness (compliance-by-design).
+
+	The analysis in this paper shows once more that inventing algorithms for preserving invariants is far from trivial.
 	Algorithms cannot be derived from constraints alone. The choices that remain must be resolved first.
@@
 
-	In practice we experience that programming in relation algebra yields an unconventional way of programming.
+	In practice we have experienced that programming by constraints in relation algebra yields an unconventional way of programming.
 	This experience consists of inventing rules and choosing event types for resolving violations, as illustrated by the case study.
@@
 	The code fragments described in section~\ref{sct:Code Fragments}
-	might be generated towards a mature reactive programming language (e.g. ELM)
-	and executed directly.
+	might be generated towards a mature reactive programming language and executed directly.
 	Ampersand lets you think in constraints and (reactive) code fragments to preserve invariants can be generated with the theory in this paper.
 	Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
-	but generation to a reactive programming language (e.g. ELM) is definitely on the wishlist.
+	but generation to a reactive programming language is definitely on the wishlist.
 
@@
 	We have used that to formalize business requirements without exposing business stakeholders to any formal notation.
-\item	Heterogeneous relation algebra in Ampersand is statically typed~\cite{Joosten2015}.
-	There is much evidence for significantly lower software maintenance cost due to static typing as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
+\item	Ampersand is statically typed~\cite{Joosten2015}, which is an obvious consequence of being heterogeneous.
+	There is much evidence that static typing yields significantly lower software maintenance cost as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
 \item	Heterogeneous relation algebra is well studied.
@@
 
-	Ampersand also has disadvantages. It appears to be difficult to learn for large groups of software professionals.
+	Ampersand also has a disadvantage. It appears to be difficult to learn for large groups of software professionals.
 	Research~\cite{Michels2015} shows that this is largely due to deficits in prerequisite knowledge, especially skills in discrete mathematics.
@@
 	Also, programming appears to be difficult in practice.
+	Ongoing research is being conducted to find better ways for learning,
+	which is necessary for large scale use.
 

### 2018-01-20 2017 JLAMP/articleRamics.tex
Review of reviewer #1 processed

@@
 \include{preamramics}
+\hyphenation{ExecEngine}
 \begin{document}
@@
 	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
-	As a case study, we have developed a database-application for legal reasoning.
+	As a case study, we have developed a database application for legal reasoning.
 	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
@@
 	This paper investigates how relation algebra can be used as a programming language for information systems.
-	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database-application.
+	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database application.
 	Ampersand is a syntactically sugared version of heterogeneous relation algebra \cite{Schmidt1997}.
@@
 
-	The use of relation algebra as a programming language is not new.
-	It stands in the tradition of relation algebra~\cite{Maddux06}, logic programming~\cite{Lloyd1984},
+	The use of relation algebra as a programming language is rooted in relation algebra~\cite{Maddux06}, logic programming~\cite{Lloyd1984},
 	database application programming~\cite{Codd70}, and formal specification of software.
 	Ampersand uses these ideas to compile relation algebra to working software, an idea which is also found in RelView~\cite{Berghammer2005} by Berghammer (Univ. of Kiel).
- 	Ampersand lets the user specify constraints only and generates an information system that respects these constraints.
-	RelView also generates software but uses relational algebra as operators in an otherwise imperative programming language.
+ 	In Ampersand the user specifies constraints. The compiler then generates an information system that respects these constraints.
+	RelView also generates software but uses relational algebra operators in an otherwise imperative programming language.
 	This makes RelView stronger in specifying complex computational problems.
@@
 
-	The author shares the belief that a provably correct implementation of requirements are essential to the software development process~\cite{Boehm1981}.
-	Being fully aware of the fate of formal methods in computer science,
-        Ampersand is nevertheless founded on relation algebra.
-	It builds further on foundations laid by formal specification methods
-	such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
-	In contrast however with such methods, Ampersand is equipped with a software generator that generates an information system.
+	The author shares the conviction that a provably correct implementation of requirements is essential to the software development process~\cite{Boehm1981}.
+	Even though formal methods are not widely used in computer science,
+        Ampersand is built on formal foundations laid by specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
+	In contrast with such methods,
+	Ampersand is equipped with a software generator that generates information systems that consist of a database with user interfaces.
 	Thus, specifying in Ampersand is developing software at the same time.
@@
 	In the Ampersand project at the Open University of the Netherlands,
-	we are studying practical and didactical issues of software development in relation algebra~\cite{Michels2015}.
+	we are studying practical and didactic issues of software development in relation algebra~\cite{Michels2015}.
 
@@
 	The theory in section~\ref{sct:Code Fragments} contributes a theory to assist in preserving invariants.
-	All lemmas in this paper have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
+	Section~\ref{sct:Code Fragments} is the extension of the paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}
+	It provides derivations of program code,
+	all of which have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
 
@@
 	In this section we explain the basics of Ampersand.
-	The reader is expected to have sufficient background in relation algebra,
-	in order to understand the remainder of this paper.
-
-	The core of an Ampersand-script is a triple $\la\rules,\rels,\concepts\ra$,
+	The reader is expected to have sufficient background in relation algebra to understand the remainder of this paper.
+
+	The core of an Ampersand script is a triple $\la\rules,\rels,\concepts\ra$,
 	which consists of a set of rules $\rules$, relations $\rels$, and concepts $\concepts$.
-	Ampersand-scripts are interpreted by the compiler as an information system.
+	An Ampersand script is interpreted by the compiler as an information system.
 	The rules constitute a theory in heterogeneous relation algebra.
@@
 
-	The Ampersand-compiler generates a database from relations in the script.
-	A database-application%
+	The Ampersand compiler generates a database from relations in the script.
+	A database application%
 \footnote{Ampersand generates an application that consists of a relational database and interface components.
@@
 	The compiler generates all software needed to keep rules satisfied at run-time.
-	If a rule is not satisfied as a result of data that has changed, that change is reverted (rolled back) to
-	preserve a state in which all rules are satisfied.
-	Changes to the database are not specified by the software developer, but generated by the compiler.
-	Rules in Ampersand are kept satisfied rather than executed directly.
+	If one of the rules is not satisfied as a result of data that has changed,
+	that change is reverted (rolled back) to the original state.
+	Either way, Ampersand ensures that all rules remain satisfied.
+	Rules in Ampersand are kept satisfied rather than being executed directly.
 
@@
 	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
-	The expression $a \in A$ means that atom $a$ is an instance of concept $A$.
+	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
 	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
-	in an Ampersand-script states that any instance of $A$ is an instance of $B$ as well.
+	in an Ampersand script states that any instance of $A$ is an instance of $B$ as well.
 	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
@@
 	The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
-	It maps each relation to a set of facts.
-	The type system of Ampersand guarantees that each pair in $r$ is always contained in its type:
+	It maps each relation to a set of pairs.
+	The type system of Ampersand guarantees that each pair satisfies:
 \begin{equation}
@@
 	Terms are used to combine relations using operators.
-	The set of terms is called $\terms$. It is defined by:
+	The set of terms is called $\terms$.
 \begin{definition}[terms]
@@
 	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}\\
-	\ti{\text{RULE}\ r = s}   &=&\ti{\fullt{A}{B}-((s-r)\cup(r-s))}\\
-	\ti{\text{RULE}\ r\subs s}&=&\ti{\fullt{A}{B}-(r-s)}
-\end{eqnarray}
-
-	We call rule $r$ \define{satisfied} when $\ti{\text{RULE}\ r = s}\ =\ \ti{\fullt{A}{B}}$.
+ resolved this \end{eqnarray}
+
+	We call a rule $\text{RULE}\ r = s$ \define{satisfied} when its interpretation equals $\ti{\fullt{A}{B}}$.
 	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
@@
 \footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
-	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.
-	Consequently, a rule is sometimes called \define{invariant}.}.
+	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
+	Consequently, such a rule may also be referred to as \define{invariant}.
+	Some ways of resolving violations are presented in section~\ref{sct:Code Fragments}.
 	Each pair in the violation set of a rule is called a violation of that rule.
@@
 	Software systems that support legal arguments have been around for many decades.
-	Legal reasoning differs from logic reasoning because of the human judgments that are involved.
+	Legal reasoning differs from logical reasoning because of the human judgments that are involved.
 	Verheij~\cite{Verheij2003} distinguishes between argument assistance systems and automated reasoning systems.
@@
 	whereas mathematical logic eradicates any human influence.
-	Legal reasoning places human judgment in the core of the very notion of truth.
+	Legal reasoning places human judgment at the core of the very notion of truth.
 	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
@@
 	because they respect the professional freedom of legal professionals to construct their own line of argumentation.
-	AASs offer help in many different ways.
-	They can help by looking up legal references, jurisdiction from the past, scholarly works etc.
+	They offer help in many different ways.
+	For instance by looking up legal references, jurisdiction from the past, scholarly works etc.
 	They can also help to construct and validate arguments by keeping arguments and evidence organized.
@@
 	Even though our case study yields a model that is a bit too large for this paper,
-	figure~\ref{fig:conceptual model} gives a good impression of what it looks like.
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[scale=.23]{LogicalDataModel.pdf}
-\end{center}
-\caption{Conceptual data model}
-\label{fig:conceptual model}
-\end{figure}
-
-	To convey the flavour of software development in relation algebra, it suffices to discuss a tiny part of the whole program.
-	Therefore, we shall discuss the part of the conceptual model that is used in the sequel.
-	The context in which a user creates arguments and reasons about them is a legal case.
-	For this reason, validity, falsehood and truth of statements are related to the case.
-	A \define{statement} is a phrase in natural language that can be either true or false.
-	An example is \stmtText{The employee, John Brown, is entitled to 50 Euros}%
-\footnote{This type of statement is typical for cases. It is valid only in the case where this particular John Brown is known.
-	In other cases, where John Brown is unknown, this statement is meaningless}.
-	In MirrorMe, a user can define a template, such as \stmtText{The employee, [emp], is entitled to [increase]}.
+	we have printed a relevant part of it at the end of this paper in figure~\ref{fig:conceptual model}.
+	This diagram gives an impression of what it looks like.
+	It shows entities such as Statement, Argument, Binding, Template, Case, and Claim, together with their attributes.
+	Attributes are the univalent and injective relations, which Ampersand derives from the script.
+	Relations that are neither univalent nor injective are n-m relations between entities.
+	In practice, the conceptual models have shown to be readable without much explanation for experienced database developers.
+
+\section{Programming in Relation Algebra}
+\label{sct:Programming in RA}
+	This section illustrates the execution model of Ampersand by means of an example.
+	A demonstration of this example was given at RAMiCS-2017.
+	The example covers a tiny part of the whole MirrorMe program,
+	which is enough to give a taste of software development in relation algebra.
+	The use case for this example is the substitution of placeholders in text,
+	for instance in replacing "\stmtText{The employee, [emp], is entitled to [increase]}" by
+        "\stmtText{The employee, James, is entitled to 50 Euros.}".
+	A user does such replacements when formulating arguments and reasoning with them in a legal case.
+	In MirrorMe, a user can define a template with placeholders in square brackets.
 	The strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
@@
 	\stmtText{Every employee is entitled to an increase in salary}.
-	The user of MirrorMe will pick a legal text (from a source he trusts) and substitute parts of that text by placeholders.
-	When the facts are known, the argument can be completed by substituting placeholders by actual phrases.
-
-	It is precisely this substitution process that we have chosen to describe in this paper as a case study in
-	programming with relation algebra.
-
-\section{Programming in Relation Algebra}
-\label{sct:Programming in RA}
+	The user of MirrorMe will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.
+	When the values of the placeholders are known, a statement can be made by substituting placeholders by actual phrases.
+	When all substitutions are made, a statement remains.
+	A \define{statement} is a phrase in natural language that can be either true or false.
+	A \define{binding} assigns a piece of text to a placeholder.
+	That text is the \define{phrase of} the binding and
+	also the \define{value} of the placeholder.
+
 	At this point we have reached the core of this case study.
-	We focus on the substitution of placeholders when their values change.
-	By focusing on this tiny detail, we can discuss the mechanics ``under the hood'' of the application generated by Ampersand.
-
-	First we show how the computer solves the issue by looking at an exerpt of a log file (figure~\ref{fig:log file}).
+	First we show how the computer solves the issue by looking at an excerpt from a log file (figure~\ref{fig:log file}).
 	It shows an alternating sequence of the computer (ExecEngine) mentioning a rule,
@@
 	The machine will only act on rules that are violated.
-	The first rule to be violated is ``signal phrase update'', in which the computer signals that something or someone has made a change
+	The first rule violated is ``signal phrase update'', in which the computer signals that something or someone has made a change
 	in relation \id{phrase}.
@@
 
-	Let us now study the actual rules to see how these rules cause the right actions to take place in the correct order.
+	Before looking at the actual rules,
+	let us introduce an auxiliary relation, \id{differB}, to represent bindings that must be updated:
+\[\id{differB} = \ident{Binding}\cap\id{shownPhrase};\cmpl{\ident{Phrase}};\flip{\id{phrase}}\]
+	The relation \id{phrase} links bindings to texts.
+	It contains the actual phrase of each binding.
+	This is also known as the value of a placeholder%
+\footnote{A phrase and a placeholder are linked by a pair from $\flip{\id{phrase}};\id{placeholder}$.}.
+	The relation \id{shownPhrase} links bindings to text fields in the user interface.
+	So \id{differB} contains the bindings of which the phrase differs from what is shown on the screen.
+	Whether a placeholder has been edited can be observed by comparing its shown phrase to the phrase.
+	Bindings in \id{differB} are to be updated,
+	because the value of placeholders should always correspond with the text on the screen.
+
+	Let us now study the rules to see how these rules cause the right actions to take place in the correct order.
 	We will follow the log file from figure~\ref{fig:log file} in reverse direction,
@@
 
-	Whether a placeholder has been edited can be observed by comparing its new phrase to the shown phrase%
-\footnote{Note that we use the notion ``the phrase of a placeholder'' to indicate a pair from $\flip{\id{phrase}};\id{placeholder}$.}.
-	The phrase of a placeholder is kept in the relation \id{phrase}.
-	The shown phrase is kept in the relation \id{shownPhrase}.
-	The sole purpose for having the relation \id{shownPhrase} is to detect a change in \id{phrase}.
-	Let us introduce \id{differB} to represent the bindings with an updated phrase:
-\[\id{differB} = \ident{Binding}\cap\id{shownPhrase};\cmpl{\ident{Phrase}};\flip{\id{phrase}}\]
-	When the phrase of a placeholder changes,
-	that phrase must be updated in every statement in which the placeholder was used.
+	The sole purpose for having the relation \id{shownPhrase} is to detect a change
+	and propagate it as the new value of a placeholder.
+	Then, the new value of the placeholder must be propagated in every statement in which the placeholder was used.
 	That update action is specified in section~\ref{rule: substitute}.
@@
 \[\text{RULE}\ (\ident{Binding}\cap\id{substituted}/\id{inStatement} );\id{phrase}\ \subs\ \id{shownPhrase} \]
-        This rule says that for each binding that has been substituted in every statement it is used in,
-	the \id{phrase} must be equal to the \id{shownPhrase}.
-	When violated, it is satisfied by inserting all violations into \id{shownPhrase}.
+        This rule says that far a binding with all substitutions made, the \id{phrase} must be equal to the \id{shownPhrase}.
+	When violated, the software inserts all violations into \id{shownPhrase} to satisfy the rule.
 
@@
 \begin{table}[htb]
-{\small\begin{tabular}{|l|p{3.3in}|}\hline
+{\small\begin{tabular}{|l|p{3.2in}|}\hline
 rule&event types\\ \hline
@@
 	That was signaled by the rule ``signal phrase update''.
-\begin{figure}[bht]
-\begin{center}
-  \includegraphics[scale=.55]{eventFlow.png}
-\end{center}
-\caption{event flow}
-\label{fig:event flow}
-\end{figure}
 	In general, the software developer must decide how to resolve violations that occur as a result of some event.
@@
 	and the dual information.
+\begin{figure}[bht]
+\begin{center}
+  \includegraphics[scale=.55]{eventFlow.png}
+\end{center}
+\caption{event flow}
+\label{fig:event flow}
+\end{figure}
 	Figure~\ref{fig:event flow} shows part of that graph.
-	A circle represents a rule, a rectangle represents an event type, and the arrows connect them.
+	A circle represents a rule, a rectangle represents an event type, and the edges correspond to the flow of values between them.
+	The propagation of changes across the graph is initiated by the event sources (e.g., a user changing the set of pairs in a relation).
 	The software developer can work her way through the graph, to write code to resolve violations for every event type that might occur.
 	Figure~\ref{fig:event flow} shows just that part of the graph that corresponds with the case study in this article.
+	In actual programs the complete graph can be much larger.
 
@@
 	which is built around two basic operations: insert and delete.
-	The insert operation can be represented in terms of pre- and postconditions\footnote{this idea
-	was introduced by Floyd and Hoare~\cite{Floyd1967,Hoare1969}.
+	The insert operation can be represented in terms of pre- and postconditions\footnote{this notation
+	was conceived by Floyd and Hoare~\cite{Floyd1967,Hoare1969}.
 	It is well known for making assertions about computer programs and reasoning about them.}.
@@
 	which represent relations.
-	We assume that before executing any code, we have $\code{P}=p$, $\code{S}=s$, and $\code{T}=t$.
+	We assume that before executing any code, we have $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$.
 	The following example shows that inserting a set of pairs, $\Delta$, in \code{P}
@@
 	we must show that its value afterwards is equal to $\full$,
-	under the condition that its initial value equals $\full$.
-	The initial value is $\cmpl{p}\cup s\cup t$, which we may assume.
-	The final value is $(p\cup\Delta)\subs(s\cup\Delta)\cup t$, which must be proven equal to $\full$:
+	assuming that its initial value is $\full$.
 \[\begin{array}{cl}
@@
 \end{array}\]
-	From this derivation we may conclude that invariance of $\code{P}\subs\code{S}\cup\code{T}$ is restored.
+	From this derivation we may conclude that invariance of $\code{P}\subs\code{S}\cup\code{T}$ is preserved.
 	From this example, we can see that $\code{INSERT }\Delta\code{ IN T}$ would have worked just as well.
@@
 \end{equation}
-	Suppose an event occurs, e.g. $\Delta_r$ is inserted in $r$,
-	and a programmer must write code that preserves invariance of rule~\ref{rule:r subs s}.
-	This requires choosing a value for $\Delta_s$, such that equation~\ref{eqn:r subsRS s} holds:
-\begin{eqnarray}
-r\subs s&\Rightarrow&r\cup\Delta_r\subs s\cup \Delta_s\label{eqn:r subsRS s}
-\end{eqnarray}
-	The programmer can choose $\Delta_s=\Delta_r$, and call it just $\Delta$,
-	satisfying equation~\ref{eqn:r subs s} trivially by monotonicity of $\subs$:
+	Recall that $\subs$ is monotonous, so:
 \begin{eqnarray}
@@
 \end{eqnarray}
-	A programmer can take the antecedent of equation~\ref{eqn:r subs s}, $r\subs s$, as precondition.
-	The consequent, $r\cup\Delta\subs s\cup \Delta$, is the postcondition.
-	She then writes the following code fragment:
+	It immediately follows that inserting $\Delta$ into \code{S} is an adequate reaction to
+	an insertion of $\Delta$ into \code{R}:
 \[\begin{array}{r@{~}l}
@@
 \end{array}\]
-	The first step, $\code{INSERT }\Delta\code{ INTO R}$, represents the event that may violate invariance of $\code{R}\subs\code{S}$.
-	The second step, $\code{INSERT }\Delta\code{ INTO S}$, represents the reaction.
 
@@
 \end{array}\]
-	Likewise, the reaction on a deletion will be an insertion:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
-\code{DELETE}&\Delta\code{ FROM R};\\
-&\{\text{$\code{R}=r-\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
-\code{INSERT}&\Delta\code{ INTO S}\\
-&\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
-\end{array}\]
-
-	Now consider equations~\ref{lbl:Preserve t=r inter s with Ins Delta into t} and~\ref{lbl:Preserve t=r union s with Del Delta from t}.
-	If \code{T} changes, both \code{R} and \code{S} must be changed accordingly.
-	Since $\cup$ and $\cap$ are associative operators, the arguments can be evaluated with no interaction between them.
-	This allows us to define a primitive, \code{ALL}, without specifying in which order it processes its arguments\footnote{... leaving room to take advantage of parallelism.}.
-	We define \code{ALL} to perform all actions from a given list:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume }\forall\code{A}\in\code{AS}: (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
-\code{ALL}&\ \code{AS}\\
-&\{\forall\code{A}: \id{post}_{\code{\tiny A}}\}
-\end{array}\]
-	With this primitive, we can write code for inserting pairs in \code{T},
-	using equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t}:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
-\code{INSERT}&\Delta\code{ INTO T};\\
-&\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
-\code{ALL}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-&\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
-\end{array}\]
-	Similarly, we can derive code for equation~\ref{lbl:Preserve t=r union s with Del Delta from t}:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
-\code{DELETE}&\Delta\code{ FROM T};\\
-&\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
-\code{ALL}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-&\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
-\end{array}\]
-
-	In equations~\ref{lbl:Preserve t=r inter s with Del Delta from t} and~\ref{lbl:Preserve t=r union s with Ins Delta into t},
-	a choice must be made between $r$ and $s$. For that purpose we need a primitive \code{CHC} that chooses one action from a list $\code{AS}$, say $\code{A}$:
-\[\begin{array}{r@{~}l}
-&\{ (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
-\code{CHC}&\ \code{AS}\\
-&\{\id{post}_{\code{\tiny A}}\}
-\end{array}\]
-	Note that this definition leaves open which term is selected.
-	The following code shows how this works out for equation~\ref{lbl:Preserve t=r inter s with Del Delta from t}:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
-\code{DELETE}&\Delta\code{ FROM T};\\
-&\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
-\code{CHC}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-&\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
-\end{array}\]
-
-	The following code shows how this works out for equation~\ref{lbl:Preserve t=r union s with Ins Delta into t}:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
-\code{INSERT}&\Delta\code{ INTO T};\\
-&\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
-\code{CHC}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-&\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
-\end{array}\]
-
-	Section~\ref{ssct:Implication Rules} treated rules of the form $\code{R}\subs\code{S}$.
-	Note that this rule can be rewritten to $\cmpl{\code{R}}\cup\code{S}=\full$.
-	Consequently, rules did not have to be treated separately.
-	We could have treated them as a special case of equations with boolean operators only.
-	In fact, every rule with boolean operators $\cup$, $\cap$, and $-$ can be rewritten as a number of rules in disjunctive normal form\footnote{%
-	Normalizing a boolean algebra is existing knowledge, and is considered outside the scope of this paper.} (DNF~\cite{davey1990}).
-	We can represent a rule in DNF informally by:
-\begin{eqnarray}
-\code{RULE }\cmpl{r_0}\cup\ldots\cup\cmpl{r_n}\cup s_0\cup\ldots\cup s_m
-\end{eqnarray}
-	or more formally by
-\begin{eqnarray}
-\full=\cmpl{\bigcap\id{antecedents}}\cup\bigcup\id{consequents}
-\end{eqnarray}
-	in which \id{antecedents} are the complemented (negative) terms and \id{consequents} are the non-complemented (positive) terms.
-
-	Summarizing, we have obtained code to preserve invariants in a boolean algebra:
+	The proofs for the following table are similar and equally simple,
+	so we omit them for the sake of brevity.
+% 	Likewise, the reaction on a deletion will be an insertion:
+% \[\begin{array}{r@{~}l}
+% &\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
+% \code{DELETE}&\Delta\code{ FROM R};\\
+% &\{\text{$\code{R}=r-\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
+% \code{INSERT}&\Delta\code{ INTO S}\\
+% &\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
+% \end{array}\]
+% 
+% 	Now consider equations~\ref{lbl:Preserve t=r inter s with Ins Delta into t} and~\ref{lbl:Preserve t=r union s with Del Delta from t}.
+% 	If \code{T} changes, both \code{R} and \code{S} must be changed accordingly.
+% 	Since $\cup$ and $\cap$ are associative operators, the arguments can be evaluated with no interaction between them.
+% 	This allows us to define a primitive, \code{ALL}, without specifying in which order it processes its arguments\footnote{... leaving room to take advantage of parallelism.}.
+% 	We define \code{ALL} to perform all actions from a given list:
+% \[\begin{array}{r@{~}l}
+% &\{\text{Assume }\forall\code{A}\in\code{AS}: (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
+% \code{ALL}&\ \code{AS}\\
+% &\{\forall\code{A}: \id{post}_{\code{\tiny A}}\}
+% \end{array}\]
+% 	With this primitive, we can write code for inserting pairs in \code{T},
+% 	using equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t}:
+% \[\begin{array}{r@{~}l}
+% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
+% \code{INSERT}&\Delta\code{ INTO T};\\
+% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
+% \code{ALL}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
+% &\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
+% \end{array}\]
+% 	Similarly, we can derive code for equation~\ref{lbl:Preserve t=r union s with Del Delta from t}:
+% \[\begin{array}{r@{~}l}
+% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
+% \code{DELETE}&\Delta\code{ FROM T};\\
+% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
+% \code{ALL}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
+% &\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
+% \end{array}\]
+% 
+% 	In equations~\ref{lbl:Preserve t=r inter s with Del Delta from t} and~\ref{lbl:Preserve t=r union s with Ins Delta into t},
+% 	a choice must be made between $r$ and $s$. For that purpose we need a primitive \code{CHC} that chooses one action from a list $\code{AS}$, say $\code{A}$:
+% \[\begin{array}{r@{~}l}
+% &\{ (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
+% \code{CHC}&\ \code{AS}\\
+% &\{\id{post}_{\code{\tiny A}}\}
+% \end{array}\]
+% 	Note that this definition leaves open which term is selected.
+% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r inter s with Del Delta from t}:
+% \[\begin{array}{r@{~}l}
+% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
+% \code{DELETE}&\Delta\code{ FROM T};\\
+% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
+% \code{CHC}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
+% &\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
+% \end{array}\]
+% 
+% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r union s with Ins Delta into t}:
+% \[\begin{array}{r@{~}l}
+% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
+% \code{INSERT}&\Delta\code{ INTO T};\\
+% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
+% \code{CHC}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
+% &\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
+% \end{array}\]
+% 
+% 	Section~\ref{ssct:Implication Rules} treated rules of the form $\code{R}\subs\code{S}$.
+% 	Note that this rule can be rewritten to $\cmpl{\code{R}}\cup\code{S}=\full$.
+% 	Consequently, rules did not have to be treated separately.
+% 	We could have treated them as a special case of equations with boolean operators only.
+% 	In fact, every rule with boolean operators $\cup$, $\cap$, and $-$ can be rewritten as a number of rules in disjunctive normal form\footnote{%
+% 	Normalizing a boolean algebra is existing knowledge, and is considered outside the scope of this paper.} (DNF~\cite{davey1990}).
+% 	We can represent a rule in DNF informally by:
+% \begin{eqnarray}
+% \code{RULE }\cmpl{r_0}\cup\ldots\cup\cmpl{r_n}\cup s_0\cup\ldots\cup s_m
+% \end{eqnarray}
+% 	or more formally by
+% \begin{eqnarray}
+% \full=\cmpl{\bigcap\id{antecedents}}\cup\bigcup\id{consequents}
+% \end{eqnarray}
+% 	in which \id{antecedents} are the complemented (negative) terms and \id{consequents} are the non-complemented (positive) terms.
+% 
+	To preserve invariants in a boolean algebra is done as follows:
 \[\begin{array}{|l|l|l|l|}
@@
 \end{array}\]
-	Having treated the boolean operators, let us consider code that is needed when relational operators are involved.
+	So let us now turn to code that is needed when relational operators are involved.
 
@@
 &r;s\cup\Delta_r;\Delta_s\\
-=&\hspace{1in}\{\text{let \code{NEWPAIRS} ensure that }r;\Delta_s\subs t\cup\Delta\text{ and }\Delta_r;s\subs t\cup\Delta\}\\
+=&\hspace{1in}\{\text{let \code{NEWPAIRS} ensure: }r;\Delta_s\subs t\cup\Delta\text{ and }\Delta_r;s\subs t\cup\Delta\}\\
 &r;s\cup r;\Delta_s\cup \Delta_r;s\cup \Delta_r;\Delta_s\\
@@
 \code{NEWPAIRS}&(\code{R}, \code{S}, \code{DELTA}): (\code{DELTAR}, \code{DELTAS})\\
-&\{\code{DELTAR};\code{DELTAS}=\Delta\text{ and }\code{R};\code{DELTAS}\subs r;s\cup\Delta\text{ and }\code{DELTAR};\code{S}\subs r;s\cup\Delta\}
+&\begin{array}[t]{@{}l@{}l}
+        \{&\code{DELTR};\code{DELTAS}=\Delta - r;s\text{ and }\\
+          &\code{R};\code{DELTAS}\subs r;s\cup\Delta\text{ and }\code{DELTAR};\code{S}\subs r;s\cup\Delta\}
+\end{array}
 \end{array}\]
@@
 \end{equation}
-	The largest solution of equation~\ref{eqn:del r ins s} follows from the law $x;(x\backslash y)\subs y$:
+	The largest $\Delta_s$ follows from the law $x;(x\backslash y)\subs y$ for $x=r-\Delta$ and $y=r;s$:
 \begin{eqnarray}
-&&x;(x\backslash y)\ \subs\ y\nonumber\\
-&\Leftrightarrow&\hspace{1in}\{\text{Let }x=r-\Delta\text{ and }y=r;s)\}\nonumber\\
 &&(r-\Delta);((r-\Delta)\backslash(r;s))\ \subs\ r;s\nonumber\\
-&\Leftrightarrow&\hspace{1in}\{\text{Let }\Delta_s\ =\ (r-\Delta)\backslash(r;s)\}\nonumber\\
+&\Rightarrow&\hspace{1in}\{\text{Let }\Delta_s\ =\ (r-\Delta)\backslash(r;s)\}\nonumber\\
 &&(r-\Delta);\Delta_s\ \subs\ r;s\nonumber\\
@@
 \end{eqnarray}
-	This derivation shows that $(r-\Delta)\backslash(r;s)$, is the largest solution.
+	This derivation shows that we can choose $\Delta_s=(r-\Delta)\backslash(r;s)$ as the largest set that preserves invariant~\ref{eqn:del r ins s}.
 
@@
 	We illustrate this point by giving an example.
-	Suppose we have $\code{T}\subs\code{R};\code{S}$ and action $\code{DELETE}\ \Delta\code{ FROM R}$.
+	Suppose we have invariant $\code{T}\subs\code{R};\code{S}$ and action $\code{DELETE}\ \Delta\code{ FROM R}$.
 	So the question is which pairs must be added to $s$ after pairs $\Delta$ have been removed from $r$.
@@
 \end{equation}
-	The following derivation shows that $\flip{(r-\Delta)};(r\cap\Delta);s$ satisfies this requirement,
-	but on one condition: for every pair in $r;s$ there must be a pair remaining behind in $r-\Delta$.
+	The following derivation shows that $\flip{(r-\Delta)};(r\cap\Delta);s$ satisfies this requirement conditionally.
 	Here is the derivation:
@@
 	but only if $r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}$.
+	The part $r;s;\flip{s};\flip{r}$ is not necessary in this proof, but it is useful in practice.
+	It says that $r-\Delta$ must contain a pair for every element in the domain of $r;s$,
+	instead of every element in $I$.
+	This makes this result applicable in more situations.
 
@@
 \label{sct:Reflection}
-	Ampersand is being built on the belief that software development should be automated.
-	The holy grail is to derive deterministic ways to preserve invariants, to allow for programming by constraints.
+	Ampersand is being built to automate software development.
+	The holy grail is to derive deterministic ways to preserve invariants, to allow programming by constraints.
 	This is motivated by the benefits of describing a result in terms of constraints,
@@
 	That gives opportunities, for instance to make these choices explicit to a user, to optimize the number of choices,
-	or to use external knowledge to resolve the remaining choices on runtime.
+	or to use external knowledge to resolve the remaining choices at run time.
 
@@
 	This experience consists of inventing rules and choosing event types for resolving violations, as illustrated by the case study.
-	To relate this experience to programming as we know it, section~\ref{sct:Comparison} makes a comparison with established programming languages (prior art).
+	To relate this experience to programming as we know it, section~\ref{sct:Comparison} makes a comparison with established programming languages.
 	Section~\ref{sct:Contribution} summarizes the contributions to software development claimed by Ampersand.
@@
 
-	The first to compare with is the imperative programming paradigm, known from popular languages such as Java~\cite{Java} and C++~\cite{Stroustrup97a}.
-	Our case study cannot be called imperative, because the notion of control flow in imperative languages is very different.
-	In imperative languages, the control flow is defined by the software developer.
-	In this case study, the control flow emerges as a result of changes in relation content as illustrated
+	The first to compare with is the object oriented programming paradigm, known from popular languages such as Java~\cite{Java} and C++~\cite{Stroustrup97a}.
+	Ampersand cannot be called object oriented, because the notion of control flow in object oriented languages is very different.
+	In those languages, the threads in the control flow are controlled by the software developer.
+	As shown in our case study, the control flow in an Ampersand script emerges as a result of changes in relation content as illustrated
 	by figure~\ref{fig:event flow}.
 
-	The case study also differs from the logic programming paradigm of Prolog~\cite{Lloyd1984} and of all rule engines that can be considered to be an offspring of Prolog.
+	Ampersand also differs from the logic programming paradigm of Prolog~\cite{Lloyd1984}
+	and of all rule engines that can be considered to be offspring of Prolog.
 	A difference lies in the way rules are interpreted.
-	In logic programming, a program consists of Horn-clauses and a resolution-proof is constructed on run-time.
+	In logic programming, a program consists of Horn-clauses and a resolution-proof is constructed at run time.
 	The software developer works with notions such as backtracking (backward chaining) and unification, which are absent in Ampersand.
-	Ampersand is not restricted to Horn-clauses; any relation-algebraic equation over relations can be used as a rule.
-
-	The case study also differs from functional programming, of which Haskell and Scala are prominent representants.
+	On the other hand, Ampersand is not restricted to implications with one predicate on the consequent side, as Prolog does.
+        Ampersand allows rules, which are equations with arbitrary relational expressions on either side of the equals sign.
+
+	Ampersand also differs from functional programming, of which Haskell and Scala are prominent representants.
 	A core idea in functional programming is to evaluate a program as a function from input to output~\cite{Backus1978}.
-	A functional program consists of function definitions, that are evaluated by a term- or graph-rewriter, using various strategies such as lazy or eager evaluation.
-	In contrast, one might interpret our case study as a relaxation of the constraint that everything is a function.
+	A functional program consists of function definitions, that are evaluated by a term- or graph-rewriter,
+	using various strategies such as lazy or eager evaluation.
+	In contrast, our case study illustrates that not everything has to be a function.
 	In relation algebra, everything is a relation and a function is a restricted form of a relation.
-	A similarity to functional programming is the declarative style, because substitution of equal terms without changing the semantics is a property we see in both worlds.
+	A similarity to functional programming is shown by the declarative style,
+	because substitution of equal terms without changing the semantics is a property we see in both worlds.
 
 	A difference with database programming is found in the type of algebra that is used.
-	Relational databases are founded on relational algebra~\cite{Codd70}. They are typically programmed in SQL.
-	In contrast with database programming, Ampersand implements heterogeneous relation algebra~\cite{Schmidt1997}.
-	A software developer working with relational algebra sees n-ary tables, while Ampersand is restricted to binary relations.
-	The comparison between relation{\it al} algebra and relation algebra might relate to comparing relational databases with graph databases~\cite{Vicknair2010},
-	although no literature was found to corroborate this.
+	Conventional databases are typically programmed in SQL, which is based on n-ary relations in relational algebra~\cite{Codd70}.
+	In contrast, Ampersand uses binary relations%
+\footnote{At runtime, Ampersand uses the efficiency of n-ary tables. It does so by translating relation algebra to efficient SQL-code by avoiding as many joins as possible.}
+	based on heterogeneous relation algebra~\cite{Schmidt1997}.
+	In this comparison, Ampersand is closer to graph databases~\cite{Vicknair2010} than to relational databases.
 
@@
 
-	The similarity with the reactive programming paradigm~\cite{Bainomugisha2013} is obvious from figure~\ref{fig:event flow}.
-	This graph is part of the program graph as conceived in reactive programs.
-	The code fragments described in section~\ref{sct:Code Fragments}
-	might be generated towards a mature reactive programming language and executed directly.
-	Ampersand lets you think in constraints and (reactive) code fragments to preserve invariants can be generated with the theory in this paper.
-	Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
-	but generation to a reactive programming language is definitely on the wishlist.
+	The similarity with the reactive programming paradigm is obvious from figure~\ref{fig:event flow}.
+	This graph is part of the dependency graph, which is a familiar phenomenon in reactive programming.
+	Ampersand may be seen as a reactive programming language on binary relations.
+	It has a unidirectional push based evaluation model, static typing, implicit lifting, and glitch avoidance~\cite{Bainomugisha2013}.
+	Even though Ampersand generates code to execute scripts,
+	code fragments described in section~\ref{sct:Code Fragments}
+	might just as well be generated towards a mature reactive programming language and executed there%
+\footnote{Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
+	but generation to a mature reactive programming language is definitely on the wishlist.}.
+	A distinguishing feature of Ampersand is that it treats constraints (rules) as its central notion,
+	which is also seen in TBAG~\cite{Elliott94} and FHM~\cite{Giorgidze2011} but not common in most reactive programming languages.
+	In the author's opinion a constraint based approach is preferrable, because the correctness of a program is within closer range.
 
@@
 	Also, programming appears to be difficult in practice.
-	Ongoing research is being conducted to find better ways for learning,
-	which is necessary for large scale use.
+	Ongoing research tries to find better ways for learning, which is necessary for large scale use.
 
@@
 	Further use of relation algebra can be made by incorporating a model checker, such as the Alloy analyser~\cite{Alloy2006}, to detect inconsistent rules.
-	An exciting new development is Amperspiegel~\cite{Amperspiegel}, which brings notational flexibility at the fingertips of the user.
-	Developments in the Ampersand-compiler are going towards a rule-repository (written in Ampersand itself).
+	An exciting new development is Amperspiegel~\cite{Amperspiegel}, which brings notational flexibility to the fingertips of the user.
+	Developments in the Ampersand compiler are going towards a rule-repository (written in Ampersand itself).
 	This will make collaborative information systems development in Ampersand easier, because the repository can assist in automating
@@
 
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[angle=90,scale=.357]{LogicalDataModel.pdf}
+\end{center}
+\caption{Conceptual data model (from section~\ref{sct:Conceptual analysis})}
+\label{fig:conceptual model}
+\end{figure}
+
 \section{Bibliography}

### 2018-01-27 2017 JLAMP/articleRamics.tex
Fixes for reviewer #1 completed

@@
 
-\title{Ampersand uses Relation Algebra as Programming Language}
+\title{Ampersand uses Relation Algebra as a Programming Language}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
 	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
-	Beside being declarative, relation algebra comes with attractive promises for developing big software.
+	Beside being declarative, relation algebra comes with attractive prospects for developing big software.
 	The compiler that was used for this case study, Ampersand, is the result of an open source project.
@@
 
-	The author shares the conviction that a provably correct implementation of requirements is essential to the software development process~\cite{Boehm1981}.
+	The author shares the conviction that a provably correct implementation of requirements
+	is essential to the software development process~\cite{Boehm1981}.
 	Even though formal methods are not widely used in computer science,
-        Ampersand is built on formal foundations laid by specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
+        Ampersand is built on the foundations laid by formal specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
 	In contrast with such methods,
-	Ampersand is equipped with a software generator that generates information systems that consist of a database with user interfaces.
+	Ampersand is equipped with a software generator that generates information systems,
+	each of which consists of a database with web-based user interfaces.
 	Thus, specifying in Ampersand is developing software at the same time.
@@
 	Ampersand interprets every relation as a finite set of pairs, which are stored in the database.
-	The phase in which Ampersand takes a script, and turns it into a database, is what we will refer to as \define{compile-time}.
+	The phase in which Ampersand takes a script and turns it into a database, is what we will refer to as \define{compile-time}.
 	The phase in which a user interacts with the database, is what we will refer to as \define{run-time}.
@@
 	We call $A$ the source concept and $B$ the target concept of the relation.
-	The pair $\pair{A}{B}$ is called the type of the relation.
+	The pair $\pair{A}{B}$ is called the \emph{type} of the relation.
 	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
@@
 	It maps each relation to a set of pairs.
-	The type system of Ampersand guarantees that each pair satisfies:
+	The type system of Ampersand guarantees that each pair in $r$ respects the type of $r$:
 \begin{equation}
@@
 \footnote{Thanks to Elleke Baecke for pointing us towards this source.}.
-	Toulmin is regarded as the first scholar in modern history to come up with a usable theory of argumentation.
 	Recent work typically draws on Toulmin, so his doctrine offers a good starting point.
@@
 	whereas mathematical logic eradicates any human influence.
-	Legal reasoning places human judgment at the core of the very notion of truth.
 	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
-	Argument assistance systems~\cite{Verheij2005} have been more successful,
-	because they respect the professional freedom of legal professionals to construct their own line of argumentation.
-	They offer help in many different ways.
+	Argument assistance systems~\cite{Verheij2005} respect the professional freedom of each legal professional
+	in constructing her own line of argumentation.
+	AAS offer help in many different ways.
 	For instance by looking up legal references, jurisdiction from the past, scholarly works etc.
@@
 	They can store, disclose and share legal evidence.
-
-	In software development, logic predominates.
-	We have used it to build an AAS.
-	The argumentation principles of Toulmin are \emph{implemented in} logic rather than \emph{replaced by} logic.
+	The MirrorMe system lets each user (whether a plaintiff, judge, or defendant) decide individually on the truth of every statement.
+	Differences in truth assessments can then be made visible and manageable.
+
+	Software development is founded on mathematical reasoning, whereas legal work is built on defeasible reasoning.
+	So we have built an AAS in the mathematical formalism of relation algebra
+	by \emph{implementing} Toulmin's principles%
+\footnote{We deliberately did not want to squeeze legal reasoning into relation algebra, or any other objective formalism.}.
 	The structure of MirrorMe was designed by conceptually analysing ideas of Toulmin,
@@
 	which is enough to give a taste of software development in relation algebra.
+
 	The use case for this example is the substitution of placeholders in text,
@@
 	A user does such replacements when formulating arguments and reasoning with them in a legal case.
-	In MirrorMe, a user can define a template with placeholders in square brackets.
-	The strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
+	In MirrorMe, a user can define templates.
+	A \define{template} is a string in which substrings between square brackets serve as placeholders.
+	In our example, the strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
 	The reason for using placeholders is that legal rules are stated in general terms, e.g.
@@
 	When the values of the placeholders are known, a statement can be made by substituting placeholders by actual phrases.
+	The substitution of a placeholder by a phrase is called a \define{binding}
 	When all substitutions are made, a statement remains.
 	A \define{statement} is a phrase in natural language that can be either true or false.
-	A \define{binding} assigns a piece of text to a placeholder.
+	In software, a statement is a template without placeholders.
+	A binding assigns a piece of text to a placeholder.
 	That text is the \define{phrase of} the binding and
@@
 	The log file of figure~\ref{fig:log file} has been taken from a computer that carries out the procedure to satisfy all rules.
+	It illustrates how the machine is busy satisfying one rule after another by inserting or deleting elements in different relations.
+	This example follows the event flow in a dependency graph, figure~\ref{fig:event flow}, which is discussed in section~\ref{sct:Programming}.
 	The machine will only act on rules that are violated.

### 2018-02-25 2018 Transactions/articleTransactions.tex
New paper about the semantics of transactions

--- 2018-01-27 2017 JLAMP/articleRamics.tex
@@
 
-\title{Ampersand uses Relation Algebra as a Programming Language}
+\title{A Theory for Database Transactions with no Database}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
 \begin{abstract}
-	Relation Algebra can be used as a programming language for building information systems.
-	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
-	As a case study, we have developed a database application for legal reasoning.
-	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
-	Beside being declarative, relation algebra comes with attractive prospects for developing big software.
-	The compiler that was used for this case study, Ampersand, is the result of an open source project.
-	Ampersand has been tried and tested in practice and is available as free open source software\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
-	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}.
+	How to do transactions if data is scattered across the internet?
+	Users who must satisfy data integrity constraints are stuck with relational databases if they need transactions.
+	For most NoSQL databases lack transaction functionality.
+	This contribution proposes a theory of database transactions solely based on data,
+	with no reference to any database in which that data resides.
+	This makes transactions suitable for a wider class of databases than just relational ones.
+	The soundness of the theory has been established by means of category theory.
+	To show that the theory works, the theory is demonstrated by a prototype developed in Ampersand\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
+	To demonstrate that the theory is about real transactions, this contribution defines semantics of conventional database transactions.
+	To fulfill our promise, we also gives semantics of transactions on the semantic web.
 \end{abstract}
@@
 \begin{keyword}
-relation algebra\sep software development\sep legal reasoning\sep information systems design\sep Ampersand\sep MirrorMe\sep big software
+relation algebra\sep software development\sep legal reasoning\sep information systems design\sep Ampersand\sep MirrorMe\sep reactive programming
 \end{keyword}
@@
 \label{sct:Introduction}
-	This paper investigates how relation algebra can be used as a programming language for information systems.
-	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database application.
-	Ampersand is a syntactically sugared version of heterogeneous relation algebra \cite{Schmidt1997}.
-	We present a case study to demonstrate programming in relation algebra and its impact on the software development process.
-	The case study takes the reader by the hand in the thinking process of a software developer who programs with relation algebra.
-
-	The use of relation algebra as a programming language is rooted in relation algebra~\cite{Maddux06}, logic programming~\cite{Lloyd1984},
-	database application programming~\cite{Codd70}, and formal specification of software.
-	Ampersand uses these ideas to compile relation algebra to working software, an idea which is also found in RelView~\cite{Berghammer2005} by Berghammer (Univ. of Kiel).
- 	In Ampersand the user specifies constraints. The compiler then generates an information system that respects these constraints.
-	RelView also generates software but uses relational algebra operators in an otherwise imperative programming language.
-	This makes RelView stronger in specifying complex computational problems.
-	Some more differences with earlier programming languages are discussed in section~\ref{sct:Comparison}.
-
-	The author shares the conviction that a provably correct implementation of requirements
-	is essential to the software development process~\cite{Boehm1981}.
-	Even though formal methods are not widely used in computer science,
-        Ampersand is built on the foundations laid by formal specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
-	In contrast with such methods,
-	Ampersand is equipped with a software generator that generates information systems,
-	each of which consists of a database with web-based user interfaces.
-	Thus, specifying in Ampersand is developing software at the same time.
-
-	The user may regard Ampersand as a programming language, especially suited for designing the back-end of information systems.
-	The axioms of Tarski can be used to manipulate expressions in a way that preserves meaning~\cite{vdWoude2011}.
-	This makes Ampersand a declarative language.
-
-	Our case study shows an argument assistant for legal professionals, which was built as innovation project at Ordina.
-	The purpose of this argument assistant is to support legal professionals in constructing a legal brief%
-\footnote{A \define{brief} is a document that is meant to summarize a lawsuit for the judge and counterparty.
-	It provides legal reasons for claims in a lawsuit based on regulations, precedents, and other legally acceptable sources.
-	It shows how the reasoning applies to facts from the case.}.
-	To the programmer, this offers the challenge to represent the problem of legal reasoning in relation algebra.
-	This challenge is by no means simple, but outside the scope of this paper.
-	In the Ampersand project at the Open University of the Netherlands,
-	we are studying practical and didactic issues of software development in relation algebra~\cite{Michels2015}.
-
-	The contribution of this paper is twofold.
-	The case study contributes to the idea that programming can be done in relation algebra.
-	The theory in section~\ref{sct:Code Fragments} contributes a theory to assist in preserving invariants.
-	Section~\ref{sct:Code Fragments} is the extension of the paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}
-	It provides derivations of program code,
-	all of which have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
-
-	Section~\ref{sct:Ampersand} introduces Ampersand and its semantics.
-	Section~\ref{sct:Conceptual analysis} introduces part of a theory of legal reasoning,
-	which was developed for argument assistance.
-	Section~\ref{sct:Programming in RA} discusses the programming mechanism in the application,
-	and section~\ref{sct:Programming} visualizes that mechanism.
-	Section~\ref{sct:Code Fragments} presents one way (of infinitely many) to derive code for arbitrary rules in relation algebra.
-	Section~\ref{sct:Reflection} reflects on software development in Ampersand.
-	It also provides an overview of the use of Ampersand in practice and an outlook to its further development.
+
+\section{Consistency}
+\label{sct:Consistency}
+	Each user of a database issues a stream of events.
+	The order (in time) of events is given by a relation $\declare{pred}{Event}{Event}}$,
+	where $\pair{e1}{e2}\in\id{pred}$ means that event $e1$ precedes event $e2$.
+	From a user perspective this stream is a sequence, so every event except the first has exactly one predecessor.
+
+	For simplicity's sake, we assume the user sees items.
+	The items seen by the user at any moment are given by a relation $\declare{elem}{Event}{Item}}$,
+	where $\pair{e}{i}\in\id{elem}$ means that item $i$ exists immediately after $e$ has occured.
+	So the set of items visible to the user exists immediately after $e$ has occured can be computed by $\{ i | \pair{e}{i}\in\id{elem} \}.
+	Note that the relation $\id{elem}$ is interpreted in an open world.
+	If $\pair{e}{i}\not\in\id{elem}$ this does not mean $i$ does not exist.
+	It just means that the pair $\pair{e}{i}$ is not in $\id{elem}$, so the user will not see it.
+
+	To cope with change, a user may insert and/or delete items at will.
+	Let us specify these deletions and insertions by two relations, $\declare{ins}{Event}{Item}}$ and $\declare{del}{Event}{Item}}$,
+	where $\pair{e}{i}\in\id{ins}$ means that item $i$ exists immediately after $e$ has occured
+	and $\pair{e}{i}\in\id{del}$ means that existence of item $i$ is not visible immediately after $e$ has occured.
+	So for every action a user takes, she wants \define{consistent behaviour}.
+	This means that insert and delete actions must satisfy:
+\begin{equation}
+	\id{elem}\ =\ (\id{pred};\id{elem}-\id{del})\cup\id{ins}
+\label{consistent behaviour}
+\end{equation}
+	Equation~\ref{consistent behaviour} says that the items visible to the user immediately after an event $e$
+	are all items that are inserted together with items that were visible immediately after the preceding event, except those that were deleted.
+
+	We define $A_{(\id{ins},\id{del})}$ as the function\footnote{The name $A$ is chosen as a reminder to the word ``action''.}
+	that embodies this change:
+\begin{equation}
+	A_{(\id{ins},\id{del})}(x)\ =\ (x-\id{del})\cup\id{ins}
+\end{equation}
+
+	This raises the question whether we can compose two actions into a single action.
+	For this, we derive:
+\[\begin{array}{r@{~}l}
+	&(((x-\id{del})\cup\id{ins})-\id{del}')\cup\id{ins}'\\
+	=&\hspace{1in}\{x-y=x\cap\overline{y}\}\\
+	&(((x\cap\overline{\id{del}})\cup\id{ins})\cap\overline{\id{del}'})\cup\id{ins}'\\
+	=&\hspace{1in}\{\text{distribute $\cap$ over $\cup$}\}\\
+	&(x\cap\overline{\id{del}}\cap\overline{\id{del}'})\cup(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'\\
+	=&\hspace{1in}\{x-y=x\cap\overline{y}\}\\
+	&((x-\id{del})-\id{del}')\cup(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'\\
+	=&\hspace{1in}\{x-y=x\cap\overline{y}\}\\
+	&(x-(\id{del})\cup\id{del}'))\cup(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'\\
+\end{array}\]
+	So the function composition operator ($.$) is defined by:
+\begin{equation}
+	A_{(\id{ins}',\id{del}')}.A_{(\id{ins},\id{del})}\ =\ A_{(\id{del})\cup\id{del}', (\id{ins}\cap\overline{\id{del}'})\cup\id{ins}')}
+\label{definition of A}
+\end{equation}
+
+\section{Isolation}
+\label{sct:Isolation}
+	If a user were alone,
+	The moment a transaction starts, its user is isolated from other users.
+	That moment is represented by a relation $\declare{beginTransaction}{Event}{Transaction}}$,
+	where $\pair{e}{t}\in\id{beginTransaction}$ means that event $e$ marks the start of transaction $t$.
 
@@
 
-	A rule is an equality between two terms.
+	A rule is a constraint in the form of an equality between two terms.
 	Terms are built from relations.
@@
 	Either way, Ampersand ensures that all rules remain satisfied.
-	Rules in Ampersand are kept satisfied rather than being executed directly.
 
@@
 	Specialization is needed to allow statements such as: ``An orange is a fruit that ....''.
-	Specialization is not used the remainder of this paper.
+	Specialization is not used in the remainder of this paper.
 
@@
 
-	The use case for this example is the substitution of placeholders in text,
-	for instance in replacing "\stmtText{The employee, [emp], is entitled to [increase]}" by
-        "\stmtText{The employee, James, is entitled to 50 Euros.}".
-	A user does such replacements when formulating arguments and reasoning with them in a legal case.
+	The use case for this example is the substitution of placeholders in text.
+	For instance, the user might want to replace \stmtText{The employee, [emp], is entitled to [raise]} by
+		\stmtText{The employee, James, is entitled to 50 Euros.}.
+	The first text is called a \define{template}, because it contains placeholders in square brackets.
+	The strings \stmtText{[emp]} and \stmtText{[raise]} are the placeholders.
+	The second text is called a \define{statement},
+	because it contains no placeholders and can be read as a statement in natural language.
+	So, a statement can be either true or false.
+
+	A user makes such replacements when formulating arguments and reasoning with them.
+	MirrorMe helps her by making replacements consistently throughout each case.
 	In MirrorMe, a user can define templates.
-	A \define{template} is a string in which substrings between square brackets serve as placeholders.
-	In our example, the strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
 	The reason for using placeholders is that legal rules are stated in general terms, e.g.
-	\stmtText{Every employee is entitled to an increase in salary}.
-	The user of MirrorMe will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.
-	When the values of the placeholders are known, a statement can be made by substituting placeholders by actual phrases.
-	The substitution of a placeholder by a phrase is called a \define{binding}
-	When all substitutions are made, a statement remains.
-	A \define{statement} is a phrase in natural language that can be either true or false.
-	In software, a statement is a template without placeholders.
+	\stmtText{Every employee is entitled to a raise in salary}.
+	A user will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.
+
+	The substitution of a placeholder by a phrase is called a \define{binding}.
 	A binding assigns a piece of text to a placeholder.
@@
 	also the \define{value} of the placeholder.
-
-	At this point we have reached the core of this case study.
-	First we show how the computer solves the issue by looking at an excerpt from a log file (figure~\ref{fig:log file}).
-	It shows an alternating sequence of the computer (ExecEngine) mentioning a rule,
-	followed by insert or delete actions to satisfy that rule.
-	Then we zoom in further, one rule at a time, to explain precisely what each rule looks like and how programming is done.
-	We then discuss the same flow of events by means of a graph (figure~\ref{fig:event flow}) in which these rules and actions are nodes.
-	This graph serves as an event flow diagram to illustrate the process behind figure~\ref{fig:log file}.
+	When all placeholders in a template have been substituted, a statement remains.
+
+	The process of substitution is embodied by a sequence of events
+	that are triggered by rule violations.
+	Events may resolve violations in one rule, but cause violations in others.
+	In this manner, rule violations and events to resolve them alternate.
+	That is shown in figure~\ref{fig:log file}.
 
@@
  InsPair(stmtShowText,Statement,Stat623,StmtText,
-         The employee, [emp], is entitled to [increase].)
+         The employee, [emp], is entitled to [raise].)
  ExecEngine satisfying rule 'done initializing'
@@
  InsPair(stmtShowText,Statement,Stat623,StmtText,
-         The employee, James, is entitled to [increase].)
+         The employee, James, is entitled to [raise].)
  InsPair(substituted,Binding,Bind625,Statement,Stat623)
@@
 	The machine will only act on rules that are violated.
-	The first rule violated is ``signal phrase update'', in which the computer signals that something or someone has made a change
-	in relation \id{phrase}.
+	The first rule violated is ``signal phrase update'', in which the computer signals a change in relation \id{phrase}.
+	That change is caused by a user who has added or deleted pairs in this relation,
+	The computer will react by changing other relations such that the rule is satisfied again.
+	This may then violate other rules, which have to be restored in turn.
 	The procedure ends by doing the necessary substitutions,
@@
 
-	The sole purpose for having the relation \id{shownPhrase} is to detect a change
+\subsection{rule: fill shownPhrase}
+\label{rule: fill shownPhrase}
+	The sole purpose of having the relation \id{shownPhrase} is to detect a change
 	and propagate it as the new value of a placeholder.
@@
 	That update action is specified in section~\ref{rule: substitute}.
-	Section~\ref{rule: fill shownPhrase} specifies how \id{shownPhrase} is made equal to \id{phrase},
+	The following rule specifies how \id{shownPhrase} is made equal to \id{phrase},
 	after all necessary substitutions are done.
-
-\subsection{rule: fill shownPhrase}
-\label{rule: fill shownPhrase}
 
@@
 	Let us now look into the process of substituting placeholders by phrases.
-	The relation \id{tmplParsedText} contains the original text, provided by a user.
-	Placeholders are specified by enclosing them in brackets,
-	e.g. \stmtText{The employee, [emp], is entitled to [increase]}.
+	The relation \id{tmplParsedText} contains the template text,
+	e.g. \stmtText{The employee, [emp], is entitled to [raise]}.
 	The text in which a placeholder has been substituted by a phrase,
@@
 	After a placeholder has been substituted, it no longer occurs in \id{stmtShowText}.
-	This poses a problem if we want to substitute the new phrase in \id{stmtShowText}.
-    For the placeholder that defines the place in the text where to substitute, is no longer in that text.
-	Therefore, substitutions must be done in the original text of the statement.
-	All placeholders in that text must then be substituted again.
+	The absence of placeholders in \id{stmtShowText} poses a problem if we want to substitute a new phrase in \id{stmtShowText}.
+	That substitution must be carried out in the original template,
+	which contains all of the original placeholders.
 	So, the text in \id{stmtShowText} must first be reset to the original text from \id{tmplParsedText}.
@@
 	the text in \id{stmtShowText} must be made equal to the original text in the template.
+	That text is obtained from the relation \id{descriptor}.
 	The action is specified by the following rule:
@@
 	Violations of this rule are descriptors of templates that belong to statements that are being reset.
-	These violations can be resolved by inserting them in \id{stmtShowText}.
+	These violations are resolved by inserting them in \id{stmtShowText}.
 
@@
 	She considers how violations arise from changing content of relations.
-	And she thinks about insert and delete actions to resolve these violations.
+	She also thinks about insert and delete actions to resolve these violations.
 	In our case study, we have reasoned with the event types: Ins $\la\id{relation}\ra$ and Del $\la\id{relation}\ra$%
@@
 	For resolving the violations in the same relation would resolve the violations by undoing the change.
-        Typically, preserving invariance of a rule is done by changing (one or more) different relations than the one that caused the violations.
+    Typically, preserving invariance of a rule is done by changing one or more other relations than the one that caused the violations.
 	The software developer can draw a graph that contains all information from table~\ref{fig:violation of rules}
@@
 	which is built around two basic operations: insert and delete.
-	The insert operation can be represented in terms of pre- and postconditions\footnote{this notation
+	The insert operation can be represented in terms of pre- and postconditions\footnote{This notation
 	was conceived by Floyd and Hoare~\cite{Floyd1967,Hoare1969}.
@@
 \hline
-\text{invariant}&\text{action}&\text{reaction}&\text{eqn.}\\ \hline
+\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
 \full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{INSERT }\Delta\code{ INTO S}&\ref{eqn:r subs s}\\
@@
 \code{DELETE}&\Delta\code{ FROM S}\\
-&\{\text{$\code{R}=r\cup\Delta$ and $\code{S}=s-\Delta$, so by eqn.~\ref{lbl:Preserve r = -s with Ins Delta into r} we have $\code{R}=\cmpl{\code{S}}$}\}
+&\{\text{$\code{R}=r\cup\Delta$ and $\code{S}=s-\Delta$, so by equation~\ref{lbl:Preserve r = -s with Ins Delta into r} we have $\code{R}=\cmpl{\code{S}}$}\}
 \end{array}\]
@@
 % \code{INSERT}&\Delta\code{ INTO S}\\
-% &\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
+% &\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
 % \end{array}\]
@@
 % 	This allows us to define a primitive, \code{ALL}, without specifying in which order it processes its arguments\footnote{... leaving room to take advantage of parallelism.}.
-% 	We define \code{ALL} to perform all actions from a given list:
+% 	We define \code{ALL} to perform all actions from a given list of actions \code{AS}:
 % \[\begin{array}{r@{~}l}
@@
 % \code{ALL}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-% &\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
+% &\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
 % \end{array}\]
@@
 % \code{ALL}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-% &\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
+% &\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
 % \end{array}\]
@@
 % \code{CHC}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-% &\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
+% &\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
 % \end{array}\]
@@
 % \code{CHC}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-% &\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by eqn.~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
+% &\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
 % \end{array}\]
@@
 \hline
-\text{invariant}&\text{action}&\text{reaction}&\text{eqn.}\\ \hline
+\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
 \code{R}=\cmpl{\code{S}}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE }\Delta\code{ FROM S}&\ref{lbl:Preserve r = -s with Ins Delta into r}\\
@@
 	In order to solve equation~\ref{eqn:ins r del s} towards $\Delta_s$,
-	we derive\footnote{with thanks to Peter Jipsen, who pointed out this line of reasoning.}:
+	we derive\footnote{Thanks to Peter Jipsen, who pointed out this line of reasoning.}:
 \begin{eqnarray}
@@
 	Once again, we might not have all pairs from $r;s$,
-	so eqn.~\ref{eqn:del r ins s} needs to be completed with the missing pairs.
+	so equation~\ref{eqn:del r ins s} needs to be completed with the missing pairs.
 	The missing pairs are $r;s-(r-\Delta);(s\cup(r-\Delta)\backslash(r;s))$.
@@
 \hline
-\text{invariants}&\text{action}&\text{reaction}&\text{eqn.}\\ \hline
+\text{invariants}&\text{action}&\text{reaction}&\text{equation}\\ \hline
 \begin{array}{l}
@@
 	As a consequence, many tools that are readily available in the public domain can be put to good use.
-	For executives of large organizations it can be reassuring that the formalism is free of childhood diseases.
+	For executives of large organizations it can be reassuring that the formalism is well studied and understood.
 \item	Relation algebra facilitates composing software from reusable components, because a program consists of rules.

### 2018-02-25 2017 JLAMP/articleRamics.tex

--- 2018-01-20 2017 JLAMP/articleRamics.tex
@@
 	The theory in section~\ref{sct:Code Fragments} contributes a theory to assist in preserving invariants.
-	All lemmas in this paper have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
+	That theory has been consolidated in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/}~\cite{Joosten-AFP17}.
+	All proofs upon which this paper is based have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
 
@@
 \end{eqnarray}
-	In proof assistant Isabelle, these lemmas are proven in zero steps, so we omit these proofs here.
+	In proof assistant Isabelle, the proof of each of these lemmas is made in zero steps, so we omit these proofs here.
 
@@
 	The code fragments described in section~\ref{sct:Code Fragments}
-	might be generated towards a mature reactive programming language and executed directly.
+	might be generated towards a mature reactive programming language (e.g. ELM)
+	and executed directly.
 	Ampersand lets you think in constraints and (reactive) code fragments to preserve invariants can be generated with the theory in this paper.
 	Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
-	but generation to a reactive programming language is definitely on the wishlist.
+	but generation to a reactive programming language (e.g. ELM) is definitely on the wishlist.
 
@@
 	We have used that to formalize business requirements without exposing business stakeholders to any formal notation.
-\item	Ampersand is statically typed~\cite{Joosten2015}, which is an obvious consequence of being heterogeneous.
-	There is much evidence that static typing yields significantly lower software maintenance cost as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
+\item	Heterogeneous relation algebra in Ampersand is statically typed~\cite{Joosten2015}.
+	There is much evidence for significantly lower software maintenance cost due to static typing as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
 \item	Heterogeneous relation algebra is well studied.
@@
 
-	Ampersand also has a disadvantage. It appears to be difficult to learn for large groups of software professionals.
+	Ampersand also has disadvantages. It appears to be difficult to learn for large groups of software professionals.
 	Research~\cite{Michels2015} shows that this is largely due to deficits in prerequisite knowledge, especially skills in discrete mathematics.
@@
 	Also, programming appears to be difficult in practice.
-	Ongoing research is being conducted to find better ways for learning,
-	which is necessary for large scale use.
 

### 2018-03-25 2018 Transactions/articleTransactions.tex
complete set for coauthor

@@
 	Each user of a database issues a stream of events.
-	The order (in time) of events is given by a relation $\declare{pred}{Event}{Event}}$,
+	The order (in time) of events is given by a relation\footnote{%
+A relation $\declare{r}{A}{B}$ is a set of tuples $\pair{a}{b}$ with $a\in A$ and $b\in B$} $\declare{pred}{Event}{Event}$,
 	where $\pair{e1}{e2}\in\id{pred}$ means that event $e1$ precedes event $e2$.
@@
 
-	For simplicity's sake, we assume the user sees items.
-	The items seen by the user at any moment are given by a relation $\declare{elem}{Event}{Item}}$,
-	where $\pair{e}{i}\in\id{elem}$ means that item $i$ exists immediately after $e$ has occured.
-	So the set of items visible to the user exists immediately after $e$ has occured can be computed by $\{ i | \pair{e}{i}\in\id{elem} \}.
-	Note that the relation $\id{elem}$ is interpreted in an open world.
-	If $\pair{e}{i}\not\in\id{elem}$ this does not mean $i$ does not exist.
-	It just means that the pair $\pair{e}{i}$ is not in $\id{elem}$, so the user will not see it.
+	For simplicity's sake, we assume the user sees a finite set of items at any given moment.
+	The items seen by the user at any moment are given by a relation $\declare{state}{Event}{Item}$,
+	where $\pair{e}{i}\in\id{state}$ means that item $i$ exists immediately after $e$ has occured.
+	So the set of items visible to the user immediately after $e$ has occured can be computed by $\{ i | \pair{e}{i}\in\id{state} \}$.
+	Note that the relation $\id{state}$ is interpreted in an open world.
+	If $\pair{e}{i}\not\in\id{state}$ this does not mean $i$ does not exist.
+	It just means that the pair $\pair{e}{i}$ is not in $\id{state}$, so the user will not see it.
 
 	To cope with change, a user may insert and/or delete items at will.
-	Let us specify these deletions and insertions by two relations, $\declare{ins}{Event}{Item}}$ and $\declare{del}{Event}{Item}}$,
-	where $\pair{e}{i}\in\id{ins}$ means that item $i$ exists immediately after $e$ has occured
-	and $\pair{e}{i}\in\id{del}$ means that existence of item $i$ is not visible immediately after $e$ has occured.
-	So for every action a user takes, she wants \define{consistent behaviour}.
-	This means that insert and delete actions must satisfy:
+	Let us specify these deletions and insertions by two relations, $\declare{ins}{Event}{Item}$ and $\declare{del}{Event}{Item}$,
+	where $\pair{e}{i}\in\id{ins}$ means that pair $\pair{e}{i}$ is to be added to $\id{state}$,
+	and $\pair{e}{i}\in\id{del}$ means that pair $\pair{e}{i}$ is to be removed from $\id{state}$.
+	More precisely, let us consider an event $e$.
+	The items $\{ i | \pair{e}{i}\in\id{state} \}$ are the items that exist immediately after the preceding event has occured,
+	minus the items to be removed, plus the items to be inserted.
+	The following equation defines this \define{behaviour}:
 \begin{equation}
-	\id{elem}\ =\ (\id{pred};\id{elem}-\id{del})\cup\id{ins}
-\label{consistent behaviour}
+	\id{state}\ =\ (\id{pred};\id{state}-\id{del})\cup\id{ins}
+\label{behaviour}
 \end{equation}
-	Equation~\ref{consistent behaviour} says that the items visible to the user immediately after an event $e$
+	Equation~\ref{behaviour} says that the items visible to the user immediately after an event $e$
 	are all items that are inserted together with items that were visible immediately after the preceding event, except those that were deleted.
@@
 	&(x\cap\overline{\id{del}}\cap\overline{\id{del}'})\cup(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'\\
-	=&\hspace{1in}\{x-y=x\cap\overline{y}\}\\
-	&((x-\id{del})-\id{del}')\cup(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'\\
-	=&\hspace{1in}\{x-y=x\cap\overline{y}\}\\
-	&(x-(\id{del})\cup\id{del}'))\cup(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'\\
+	=&\hspace{1in}\{\text{De Morgan}\}\\
+	&(x-(\id{del}\cup\id{del}'))\cup(\id{ins}-\id{del}')\cup\id{ins}'\\
+	=&\hspace{1in}\{\text{Assume}\ \id{ins}\cap\id{del}=\emptyset\ \text{and}\ \id{ins}'\cap\id{del}'=\emptyset\}\\
+	&(x-(\id{del}\cup\id{del}'))\cup(\id{ins}\cup\id{ins}')
 \end{array}\]
-	So the function composition operator ($.$) is defined by:
+	This yields composition ($;$) of $A$:
 \begin{equation}
-	A_{(\id{ins}',\id{del}')}.A_{(\id{ins},\id{del})}\ =\ A_{(\id{del})\cup\id{del}', (\id{ins}\cap\overline{\id{del}'})\cup\id{ins}')}
-\label{definition of A}
+	A_{(\id{ins},\id{del})}\ ;\ A_{(\id{ins}',\id{del}')}\ =\ A_{((\id{ins}-\id{del}')\cup\id{ins}', \id{del}\cup\id{del}')}
+\label{composition of A}
 \end{equation}
+
+	The composition operator turns out to be commutative, but under a condition.
+	This will appear to be useful later on in this paper.
+	Equation~\ref{composition of A} shows that we only need to investigate commutativity in the first argument of $A$: 
+\[\begin{array}{r@{~}rcl}
+	&A_{(\id{ins},\id{del})}\ ;\ A_{(\id{ins}',\id{del}')}&=&A_{(\id{ins}',\id{del}')}\ ;\ A_{(\id{ins},\id{del})}\\
+	\Leftrightarrow\\
+	&A_{((\id{ins}-\id{del}')\cup\id{ins}', \id{del}\cup\id{del}')}&=&A_{((\id{ins}'-\id{del})\cup\id{ins}, \id{del}'\cup\id{del})}\\
+	\Leftrightarrow\\
+	&(\id{ins}-\id{del}')\cup\id{ins}'&=&(\id{ins}'-\id{del})\cup\id{ins}\\
+	\Leftrightarrow\\
+	&(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'&=&(\id{ins}'\cap\overline{\id{del}})\cup\id{ins}\\
+	\Leftrightarrow\\
+	&(\id{ins}\cup\id{ins}')\cap(\overline{\id{del}'}\cup\id{ins}')&=&(\id{ins}\cup\id{ins}')\cap(\overline{\id{del}}\cup\id{ins})\\
+	\Leftrightarrow\\
+	&(\id{ins}\cup\id{ins}')-\overline{\overline{\id{del}'}\cup\id{ins}'}&=&(\id{ins}\cup\id{ins}')-\overline{\overline{\id{del}}\cup\id{ins}}\\
+	\Leftrightarrow\\
+	&\id{del}'\cap\overline{\id{ins}'}&=&\id{del}\cap\overline{\id{ins}}\\
+	\Leftrightarrow\\
+	&\id{del}'-\id{ins}'&=&\id{del}-\id{ins}
+\end{array}\]
+	For the purpose of making commutativity independent of its arguments,
+	we must check the condition $\id{del}'-\id{ins}'=\id{del}-\id{ins}$.
 
@@
 	The moment a transaction starts, its user is isolated from other users.
-	That moment is represented by a relation $\declare{beginTransaction}{Event}{Transaction}}$,
+	That moment is represented by a relation $\declare{beginTransaction}{Event}{Transaction}$,
 	where $\pair{e}{t}\in\id{beginTransaction}$ means that event $e$ marks the start of transaction $t$.
@@
 \label{sct:Conceptual analysis}
-	As a case study, an argument assistance system, MirrorMe, was built.
-	We have chosen to implement the ideas of Toulmin \cite{Toulmin1958},
-	because his book ``The uses of Arguments'' is still one of the most influential works in the area of legal argument management%
-\footnote{Thanks to Elleke Baecke for pointing us towards this source.}.
-	Recent work typically draws on Toulmin, so his doctrine offers a good starting point.
-	Toulmin's ideas have been implemented before, for example in a tool called ArguMed \cite{Verheij1999}.
 
-	Software systems that support legal arguments have been around for many decades.
-	Legal reasoning differs from logical reasoning because of the human judgments that are involved.
-	Verheij~\cite{Verheij2003} distinguishes between argument assistance systems and automated reasoning systems.
-	An automated reasoning system (ARS) mimics human reasoning in an attempt to automate a decision.
-	An argument assisance system (AAS) helps its user to construct a decision.
-	The literature on legal reasoning~\cite{Lind2007} makes it quite clear why mathematical logic alone does not suffice.
-	Legal reasoning puts a human decision at the root of its notion of truth,
-	whereas mathematical logic eradicates any human influence.
-	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
-	Argument assistance systems~\cite{Verheij2005} respect the professional freedom of each legal professional
-	in constructing her own line of argumentation.
-	AAS offer help in many different ways.
-	For instance by looking up legal references, jurisdiction from the past, scholarly works etc.
-	They can also help to construct and validate arguments by keeping arguments and evidence organized.
-	They can store, disclose and share legal evidence.
-	The MirrorMe system lets each user (whether a plaintiff, judge, or defendant) decide individually on the truth of every statement.
-	Differences in truth assessments can then be made visible and manageable.
-
-	Software development is founded on mathematical reasoning, whereas legal work is built on defeasible reasoning.
-	So we have built an AAS in the mathematical formalism of relation algebra
-	by \emph{implementing} Toulmin's principles%
-\footnote{We deliberately did not want to squeeze legal reasoning into relation algebra, or any other objective formalism.}.
-	The structure of MirrorMe was designed by conceptually analysing ideas of Toulmin,
-	such as claim, warrant, argument, and rebuttal. They appear in MirrorMe as relations.
-	The Ampersand compiler generates a conceptual data model
-	to help the software developer to oversee all relations.
-	Even though our case study yields a model that is a bit too large for this paper,
-	we have printed a relevant part of it at the end of this paper in figure~\ref{fig:conceptual model}.
-	This diagram gives an impression of what it looks like.
-	It shows entities such as Statement, Argument, Binding, Template, Case, and Claim, together with their attributes.
-	Attributes are the univalent and injective relations, which Ampersand derives from the script.
-	Relations that are neither univalent nor injective are n-m relations between entities.
-	In practice, the conceptual models have shown to be readable without much explanation for experienced database developers.
-
-\section{Programming in Relation Algebra}
-\label{sct:Programming in RA}
-	This section illustrates the execution model of Ampersand by means of an example.
-	A demonstration of this example was given at RAMiCS-2017.
-	The example covers a tiny part of the whole MirrorMe program,
-	which is enough to give a taste of software development in relation algebra.
-
-	The use case for this example is the substitution of placeholders in text.
-	For instance, the user might want to replace \stmtText{The employee, [emp], is entitled to [raise]} by
-		\stmtText{The employee, James, is entitled to 50 Euros.}.
-	The first text is called a \define{template}, because it contains placeholders in square brackets.
-	The strings \stmtText{[emp]} and \stmtText{[raise]} are the placeholders.
-	The second text is called a \define{statement},
-	because it contains no placeholders and can be read as a statement in natural language.
-	So, a statement can be either true or false.
-
-	A user makes such replacements when formulating arguments and reasoning with them.
-	MirrorMe helps her by making replacements consistently throughout each case.
-	In MirrorMe, a user can define templates.
-	The reason for using placeholders is that legal rules are stated in general terms, e.g.
-	\stmtText{Every employee is entitled to a raise in salary}.
-	A user will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.
-
-	The substitution of a placeholder by a phrase is called a \define{binding}.
-	A binding assigns a piece of text to a placeholder.
-	That text is the \define{phrase of} the binding and
-	also the \define{value} of the placeholder.
-	When all placeholders in a template have been substituted, a statement remains.
-
-	The process of substitution is embodied by a sequence of events
-	that are triggered by rule violations.
-	Events may resolve violations in one rule, but cause violations in others.
-	In this manner, rule violations and events to resolve them alternate.
-	That is shown in figure~\ref{fig:log file}.
-
-\begin{figure}[htb]
-\begin{verbatim}
- ExecEngine run started
- ExecEngine satisfying rule 'signal phrase update'
- InsPair(resetS,Statement,Stat623,Statement,Stat623)
- ExecEngine satisfying rule 'flush substitutions'
- DelPair(substituted,Binding,Bind625,Statement,Stat623)
- ExecEngine satisfying rule 'reset statement text'
- InsPair(stmtShowText,Statement,Stat623,StmtText,
-         The employee, [emp], is entitled to [raise].)
- ExecEngine satisfying rule 'done initializing'
- DelPair(resetS,Statement,Stat623,Statement,Stat623)
- ExecEngine satisfying rule 'substitute'
- InsPair(stmtShowText,Statement,Stat623,StmtText,
-         The employee, James, is entitled to [raise].)
- InsPair(substituted,Binding,Bind625,Statement,Stat623)
- ExecEngine satisfying rule 'fill shownPhrase'
- InsPair(shownPhrase,Binding,Bind625,Phrase,James)
- ExecEngine run completed
-\end{verbatim}
-\caption{Log file of a substitution}
-\label{fig:log file}
-\end{figure}
-	The log file of figure~\ref{fig:log file} has been taken from a computer that carries out the procedure to satisfy all rules.
-	It illustrates how the machine is busy satisfying one rule after another by inserting or deleting elements in different relations.
-	This example follows the event flow in a dependency graph, figure~\ref{fig:event flow}, which is discussed in section~\ref{sct:Programming}.
-	The machine will only act on rules that are violated.
-	The first rule violated is ``signal phrase update'', in which the computer signals a change in relation \id{phrase}.
-	That change is caused by a user who has added or deleted pairs in this relation,
-	The computer will react by changing other relations such that the rule is satisfied again.
-	This may then violate other rules, which have to be restored in turn.
-	The procedure ends by doing the necessary substitutions,
-	ensuring that all statements have the actual phrase of a placeholder in their text.
-
-	Before looking at the actual rules,
-	let us introduce an auxiliary relation, \id{differB}, to represent bindings that must be updated:
-\[\id{differB} = \ident{Binding}\cap\id{shownPhrase};\cmpl{\ident{Phrase}};\flip{\id{phrase}}\]
-	The relation \id{phrase} links bindings to texts.
-	It contains the actual phrase of each binding.
-	This is also known as the value of a placeholder%
-\footnote{A phrase and a placeholder are linked by a pair from $\flip{\id{phrase}};\id{placeholder}$.}.
-	The relation \id{shownPhrase} links bindings to text fields in the user interface.
-	So \id{differB} contains the bindings of which the phrase differs from what is shown on the screen.
-	Whether a placeholder has been edited can be observed by comparing its shown phrase to the phrase.
-	Bindings in \id{differB} are to be updated,
-	because the value of placeholders should always correspond with the text on the screen.
-
-	Let us now study the rules to see how these rules cause the right actions to take place in the correct order.
-	We will follow the log file from figure~\ref{fig:log file} in reverse direction,
-	reasoning backwards from the result.
-
-\subsection{rule: fill shownPhrase}
-\label{rule: fill shownPhrase}
-	The sole purpose of having the relation \id{shownPhrase} is to detect a change
-	and propagate it as the new value of a placeholder.
-	Then, the new value of the placeholder must be propagated in every statement in which the placeholder was used.
-	That update action is specified in section~\ref{rule: substitute}.
-	The following rule specifies how \id{shownPhrase} is made equal to \id{phrase},
-	after all necessary substitutions are done.
-
-\[\text{RULE}\ (\ident{Binding}\cap\id{substituted}/\id{inStatement} );\id{phrase}\ \subs\ \id{shownPhrase} \]
-        This rule says that far a binding with all substitutions made, the \id{phrase} must be equal to the \id{shownPhrase}.
-	When violated, the software inserts all violations into \id{shownPhrase} to satisfy the rule.
-
-\subsection{rule: substitute}
-\label{rule: substitute}
-	Let us now look into the process of substituting placeholders by phrases.
-	The relation \id{tmplParsedText} contains the template text,
-	e.g. \stmtText{The employee, [emp], is entitled to [raise]}.
-	The text in which a placeholder has been substituted by a phrase,
-	e.g. \stmtText{The employee, John Brown, is entitled to 50 Euros},
-	is kept in relation \id{stmtShowText}.
-	Each substitution that has been done in a statement corresponds to a binding-statement pair in the relation \id{substituted}.
-	This relation keeps track of all substitutions.
-	After a placeholder has been substituted, it no longer occurs in \id{stmtShowText}.
-	The absence of placeholders in \id{stmtShowText} poses a problem if we want to substitute a new phrase in \id{stmtShowText}.
-	That substitution must be carried out in the original template,
-	which contains all of the original placeholders.
-	So, the text in \id{stmtShowText} must first be reset to the original text from \id{tmplParsedText}.
-	To keep track of substitutions correctly, all corresponding binding-statement pairs must be removed from the relation \id{substituted}.
-	Only after resetting is done, the substitutions can be put back in place with the new phrases filled in.
-
-	We define a relation \id{resetS} to register the statements that are being reset.
-    In statements that are not being reset, $\ident{Statement}-\id{resetS}$, substitutions can take place.
-	All placeholders that have a binding with a phrase can be substituted. 
-	The following rule specifies the action of substituting placeholders.
-\begin{displaymath}
-\begin{array}{rl}
-\multicolumn{2}{l}{\text{RULE}}\\
-&(\ident{Binding}\cap \id{phrase} ;\flip{\id{phrase}});\id{inStatement} ;(\ident{Statement} - \id{resetS})\\
-\subs\\
-&\id{substituted}
-\end{array}
-\end{displaymath}
-	Violations of this rule are binding-statement pairs, of which the binding has a phrase and the statement is not being reset.
-	Hence, this rule can be satisfied by inserting every violation into the relation substituted.
-
-\subsection{rule: done initializing}
-\label{rule: done initializing}
-	Resetting a statement is done when two conditions are met.
-	First, every statement that is (still) being reset may have no bindings in the relation \id{substituted}.
-	Second, the text in \id{stmtShowText} corresponds to the text in \id{tmplParsedText}.
-	So the rule that specifies the action is:
-
-\begin{displaymath}
-\begin{array}{rl}
-\multicolumn{2}{l}{\text{RULE}}\\
-&(\id{resetS} - \flip{\id{inStatement}};\id{substituted})\ \cap\\
-&\id{template} ;\id{tmplParsedText} ;\flip{\id{stmtShowText}}\\
-\subs\\
-&\overline{\id{resetS}}
-\end{array}
-\end{displaymath}
-	Violations of this rule are statements that are no longer being reset, but are still in the relation \id{resetS}.
-	The appropriate action is to remove them from \id{resetS}.
-
-\subsection{rule: reset statement text}
-\label{rule: reset statement text}
-	To satisfy one condition from section~\ref{rule: done initializing},
-	the text in \id{stmtShowText} must be made equal to the original text in the template.
-	That text is obtained from the relation \id{descriptor}.
-	The action is specified by the following rule:
-\[\text{RULE}\ \id{resetS} ;\id{template} ;\id{descriptor}\ \subs\ \id{stmtShowText} \]
-	Violations of this rule are descriptors of templates that belong to statements that are being reset.
-	These violations are resolved by inserting them in \id{stmtShowText}.
-
-\subsection{rule: flush substitutions}
-\label{rule: flush substitutions}
-	To satisfy the other condition from section~\ref{rule: done initializing},
-	the following rule specifies the action to be taken:
-\[\text{RULE}\ \fullt{Binding}{Binding};\id{inStatement} ;\id{resetS}\ \subs\ \overline{\id{substituted} }\]
-	Every binding in a statement that is being reset needs to be removed from the relation \id{substituted}.
-	The software developer can implement this by deleting all violations of this rule from the relation \id{substituted}.
-
-\subsection{rule: signal phrase update}
-\label{rule: signal phrase update}
-	When the phrase in a binding is edited and the new phrase differs from the shown phrase,
-	this signals that substitutions must be flushed (see~\ref{rule: flush substitutions}),
-	that the statement text must be reset to the original text (see~\ref{rule: reset statement text}),
-	that the reset-state must be revoked (see~\ref{rule: done initializing}),
-	that substitution must take place (see~\ref{rule: substitute}),
-	and finally that the phrase detection is switched off again (see~\ref{rule: fill shownPhrase}).
-	The initial condition occurs if a binding has been used (substituted) in a statement,
-	and the binding satisfies \id{differB}.
-	The following rule specifies the action that resetting a statment can start:
-\[\text{RULE}\ \ident{Statement}\cap \flip{\id{inStatement}};\id{{\it differB}};\id{substituted}\ \subs\ \id{resetS} \]
-	Violations of this rule are statements of which a substitution must be re-done.
-	The software developer can have these violations added to \id{resetS} to satisfy this rule.
-	In doing so, the chain of events is triggered that ends when all rules are satisfied.
-
-\section{Programming in the Small}
-\label{sct:Programming}
-	Let us take a closer look at the programming process.
-	Working in relation algebra, a software developer must think about satisfying constraints.
-	She considers how violations arise from changing content of relations.
-	She also thinks about insert and delete actions to resolve these violations.
-	In our case study, we have reasoned with the event types: Ins $\la\id{relation}\ra$ and Del $\la\id{relation}\ra$%
-\footnote{Ins and Del are called each others duals.}.
-\begin{table}[htb]
-{\small\begin{tabular}{|l|p{3.2in}|}\hline
-rule&event types\\ \hline
-signal phrase update&\small Ins \id{inStatement}, Ins \id{differB}, Ins \id{substituted}, Del \id{resetS}\\
-flush substitutions&\small Ins \id{inStatement}, Ins \id{resetS}, Del \id{substituted}\\
-reset statement text&\small Ins \id{resetS}, Ins \id{template}, Ins \id{descriptor}, Del \id{stmtShowText}\\
-done initializing&\small Ins \id{resetS}, Del \id{inStatement}, Del \id{substituted}, Ins \id{template}, Ins \id{tmplParsedText}, Ins \id{stmtShowText}\\
-substitute&\small Ins \id{phrase}, Ins \id{inStatement}, Del \id{resetS}, Del \id{substituted}\\
-fill shownPhrase&\small Ins \id{substituted}, Del \id{inStatement}, Del \id{shownPhrase}, Ins \id{phrase}\\ \hline
-\end{tabular}}
-\caption{By which type of events can rules be violated?}
-\label{fig:violation of rules}
-\end{table}
-	Table~\ref{fig:violation of rules} shows which event types may violate which rules.
-	Recall section~\ref{sct:Programming in RA}, where the process of substitution started by updating the value of a placeholder.
-	This meant doing an insert after a delete on the relation \id{phrase}, causing the updated phrase to appear in \id{differB}.
-	That was signaled by the rule ``signal phrase update''.
-	In general, the software developer must decide how to resolve violations that occur as a result of some event.
-	This can be done by choosing among the duals of the event types from table~\ref{fig:violation of rules}.
-	By systematically swapping Ins for Del and vice-versa,
-	the table shows by which type of events violations can be resolved.
-	To ensure progress, the software developer will pick a different relation.
-	For resolving the violations in the same relation would resolve the violations by undoing the change.
-    Typically, preserving invariance of a rule is done by changing one or more other relations than the one that caused the violations.
-	The software developer can draw a graph that contains all information from table~\ref{fig:violation of rules}
-	and the dual information.
-\begin{figure}[bht]
-\begin{center}
-  \includegraphics[scale=.55]{eventFlow.png}
-\end{center}
-\caption{event flow}
-\label{fig:event flow}
-\end{figure}
-	Figure~\ref{fig:event flow} shows part of that graph.
-	A circle represents a rule, a rectangle represents an event type, and the edges correspond to the flow of values between them.
-	The propagation of changes across the graph is initiated by the event sources (e.g., a user changing the set of pairs in a relation).
-	The software developer can work her way through the graph, to write code to resolve violations for every event type that might occur.
-	Figure~\ref{fig:event flow} shows just that part of the graph that corresponds with the case study in this article.
-	In actual programs the complete graph can be much larger.
-
-\section{Code Fragments}
-\label{sct:Code Fragments}
-	After the previous section, one question remains open: which code can preserve invariance and can we produce that code for arbitrary rules?
-	We shall provide an answer to this question not just for the case study at hand, but for all conceivable cases.
-	This section derives such code for every term in relation algebra.
-	The problem of generating such code suffers from arbitrariness: there are many ways to generate correct code.
-	We have not explored all possibilities, because there are too many.
-	We do give one implementation for every conceivable term, pointing out where choices are to be made.
-
-	The purpose of the code is to preserve invariance of rules by insertion and deletion into and from relations.
-	Inserting (a set of) pairs $\Delta$ into a relation $r$ yields $r\cup\Delta$.
-	Likewise, deleting pairs $\Delta$ from a relation $r$ yields $r-\Delta$.
-	In this section we derive code for computers,
-	which is built around two basic operations: insert and delete.
-	The insert operation can be represented in terms of pre- and postconditions\footnote{This notation
-	was conceived by Floyd and Hoare~\cite{Floyd1967,Hoare1969}.
-	It is well known for making assertions about computer programs and reasoning about them.}.
-\[\begin{array}{r@{~}l}
-&\{\text{pre: }\code{P}=p\}\\
-\code{INSERT}&\Delta\code{ IN P}\\
-&\{\text{post: }\code{P}=p\cup\Delta\}\\
-\end{array}\]
-
-	The delete operation is introduced similarly:
-\[\begin{array}{r@{~}l}
-&\{\text{pre: }\code{P}=p\}\\
-\code{DELETE}&\Delta\code{ FROM P}\\
-&\{\text{post: }\code{P}=p-\Delta\}\\
-\end{array}\]
-
-	Let the following example introduce the idea of breaking and restoring invariance.
-	Suppose the rule $\code{P}\subs\code{S}\cup\code{T}$ is to be kept satisfied by a computer,
-	and \code{P}, \code{S}, and \code{T} are variables (memory locations) in that computer,
-	which represent relations.
-	We assume that before executing any code, we have $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$.
-	The following example shows that inserting a set of pairs, $\Delta$, in \code{P}
-	may violate the rule.
-	Inserting the same $\Delta$ in \code{S} restores invariance.
-	In terms of code, we write:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
-\code{INSERT}&\Delta\code{ IN P}\\
-&\{\text{$\code{P}=p\cup\Delta$, $\code{S}=s$, and $\code{T}=t$, and invariance is broken: $\code{P}\not\subs\code{S}\cup\code{T}$}\}\\
-\code{INSERT}&\Delta\code{ IN S}\\
-&\{\text{postcondition: $\code{P}=p\cup\Delta$, $\code{S}=s\cup\Delta$, $\code{T}=t$, and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
-\end{array}\]
-
-	To prove that rule $\code{P}\subs\code{S}\cup\code{T}$ is kept satisfied,
-	we must show that its value afterwards is equal to $\full$,
-	assuming that its initial value is $\full$.
-\[\begin{array}{cl}
-&(p\cup\Delta)\subs(s\cup\Delta)\cup t\\
-=&\\
-&\cmpl{p\cup\Delta}\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
-=&\hspace{1in}\{\hbox{De Morgan's law}\}\\
-&(\cmpl{p}\cap\cmpl{\Delta})\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
-=&\hspace{1in}\{\hbox{Distribute $\cup$ over $\cap$}\}\\
-&(\cmpl{p}\cup s\cup\Delta\cup t)\ \cap\ (\cmpl{\Delta}\cup s\cup\Delta\cup t)\\
-=&\hspace{1in}\{\cmpl{\Delta}\cup\Delta=\full\}\\
-&\cmpl{p}\cup s\cup\Delta\cup t\\
-=&\hspace{1in}\{\full\ =\ \cmpl{p}\cup s\cup t\}\\
-&\full
-\end{array}\]
-	From this derivation we may conclude that invariance of $\code{P}\subs\code{S}\cup\code{T}$ is preserved.
-	From this example, we can see that $\code{INSERT }\Delta\code{ IN T}$ would have worked just as well.
-	In general, there are alternative ways of preserving invariance.
-	This calls for a systematic analysis.
-	The following sections analyze the various terms in Relation Algebra and provides some code to restore invariance of rules containing such terms.
-	We start with an analysis of rules, then we proceed with boolean operators, and finally we arrive at relational operators.
-
-\subsection{Implication Rules}
-\label{ssct:Implication Rules}
-	We will start by studying insertions and deletions in the simplest of rules:
-\begin{equation}
-\code{RULE }r\subs s\label{rule:r subs s}
-\end{equation}
-	Recall that $\subs$ is monotonous, so:
-\begin{eqnarray}
-r\subs s&\Rightarrow&r\cup\Delta\subs s\cup \Delta\label{eqn:r subs s}
-\end{eqnarray}
-	It immediately follows that inserting $\Delta$ into \code{S} is an adequate reaction to
-	an insertion of $\Delta$ into \code{R}:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
-\code{INSERT}&\Delta\code{ INTO R};\\
-&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
-\code{INSERT}&\Delta\code{ INTO S}\\
-&\{\text{postcondition: $\code{R}=r\cup\Delta$, $\code{S}=s\cup\Delta$, and $\code{R}\subs\code{S}$}\}
-\end{array}\]
-
-	A similar line of reasoning can be used for deletion as well:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
-\code{DELETE}&\Delta\code{ FROM S};\\
-&\{\text{$\code{R}=r$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
-\code{DELETE}&\Delta\code{ FROM R}\\
-&\{\text{postcondition: $\code{R}=r-\Delta$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$}\}
-\end{array}\]
-	The proof is noteworthy similar and equally trivial:
-\begin{eqnarray}
-r\subs s&\Rightarrow&r-\Delta\subs s-\Delta\label{eqn:r subs s by deletion}
-\end{eqnarray}
-	If the event is a deletion of $\Delta$ from $r$ or an insertion into $s$, no reaction is required.
-	Such events cannot violate the invariance of rule~\ref{rule:r subs s},
-	because equations~\ref{eqn:r subs s delfrom r} and~\ref{eqn:r subs s insinto s} hold for any $\Delta$:
-\begin{eqnarray}
-r\subs s&\Rightarrow&r-\Delta\subs s\label{eqn:r subs s delfrom r}\\
-r\subs s&\Rightarrow&r\subs s\cup\Delta\label{eqn:r subs s insinto s}
-\end{eqnarray}
-
-Summarizing, some code for preserving invariance of rules of form $\code{R}\subs\code{S}$ is:
-\[\begin{array}{|l|l|l|l|}
-\hline
-\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
-\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{INSERT }\Delta\code{ INTO S}&\ref{eqn:r subs s}\\
-\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM S}&\code{DELETE }\Delta\code{ FROM R}&\ref{eqn:r subs s by deletion}\\
-\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\ref{eqn:r subs s delfrom r}\\
-\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO S}&\code{Nop}&\ref{eqn:r subs s insinto s}\\ \hline
-\end{array}\]
-Preserving an invariant of the form $\code{R}=\code{S}$ is equivalent to preserving both $\code{R}\subs\code{S}$ and $\code{S}\subs\code{R}$ simultaneously.
-
-\subsection{Rules with Boolean Operators}
-	To write code for boolean expressions,
-	we revert to the following lemma's:
-\begin{eqnarray}
-r=\cmpl{s}&\Rightarrow&r\cup\Delta=\cmpl{s-\Delta}\label{lbl:Preserve r = -s with Ins Delta into r}\\
-r=\cmpl{s}&\Rightarrow&r-\Delta=\cmpl{s\cup\Delta}\label{lbl:Preserve r = -s with Del Delta from r}\\
-t=r\cap s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cap(s\cup\Delta)\label{lbl:Preserve t=r inter s with Ins Delta into t}\\
-t=r\cup s&\Rightarrow&t-\Delta=(r-\Delta)\cup(s-\Delta)\label{lbl:Preserve t=r union s with Del Delta from t}\\
-t=r\cap s&\Rightarrow&t-\Delta=(r-\Delta)\cap s\label{lbl:Preserve t=r inter s with Del Delta from t}\\
-t=r\cup s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cup s\label{lbl:Preserve t=r union s with Ins Delta into t}
-\end{eqnarray}
-	In proof assistant Isabelle, these lemmas are proven in zero steps, so we omit these proofs here.
-
-	Equations~\ref{lbl:Preserve r = -s with Ins Delta into r} and~\ref{lbl:Preserve r = -s with Del Delta from r}
-	show that insertion in a term means deletion from its complement and vice versa.
-	So the programmer can write the following code:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
-\code{INSERT}&\Delta\code{ INTO R};\\
-&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
-\code{DELETE}&\Delta\code{ FROM S}\\
-&\{\text{$\code{R}=r\cup\Delta$ and $\code{S}=s-\Delta$, so by equation~\ref{lbl:Preserve r = -s with Ins Delta into r} we have $\code{R}=\cmpl{\code{S}}$}\}
-\end{array}\]
-	The proofs for the following table are similar and equally simple,
-	so we omit them for the sake of brevity.
-% 	Likewise, the reaction on a deletion will be an insertion:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
-% \code{DELETE}&\Delta\code{ FROM R};\\
-% &\{\text{$\code{R}=r-\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
-% \code{INSERT}&\Delta\code{ INTO S}\\
-% &\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
-% \end{array}\]
-% 
-% 	Now consider equations~\ref{lbl:Preserve t=r inter s with Ins Delta into t} and~\ref{lbl:Preserve t=r union s with Del Delta from t}.
-% 	If \code{T} changes, both \code{R} and \code{S} must be changed accordingly.
-% 	Since $\cup$ and $\cap$ are associative operators, the arguments can be evaluated with no interaction between them.
-% 	This allows us to define a primitive, \code{ALL}, without specifying in which order it processes its arguments\footnote{... leaving room to take advantage of parallelism.}.
-% 	We define \code{ALL} to perform all actions from a given list of actions \code{AS}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume }\forall\code{A}\in\code{AS}: (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
-% \code{ALL}&\ \code{AS}\\
-% &\{\forall\code{A}: \id{post}_{\code{\tiny A}}\}
-% \end{array}\]
-% 	With this primitive, we can write code for inserting pairs in \code{T},
-% 	using equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{INSERT}&\Delta\code{ INTO T};\\
-% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{ALL}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-% &\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
-% \end{array}\]
-% 	Similarly, we can derive code for equation~\ref{lbl:Preserve t=r union s with Del Delta from t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{DELETE}&\Delta\code{ FROM T};\\
-% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{ALL}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-% &\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
-% \end{array}\]
-% 
-% 	In equations~\ref{lbl:Preserve t=r inter s with Del Delta from t} and~\ref{lbl:Preserve t=r union s with Ins Delta into t},
-% 	a choice must be made between $r$ and $s$. For that purpose we need a primitive \code{CHC} that chooses one action from a list $\code{AS}$, say $\code{A}$:
-% \[\begin{array}{r@{~}l}
-% &\{ (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
-% \code{CHC}&\ \code{AS}\\
-% &\{\id{post}_{\code{\tiny A}}\}
-% \end{array}\]
-% 	Note that this definition leaves open which term is selected.
-% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r inter s with Del Delta from t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{DELETE}&\Delta\code{ FROM T};\\
-% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{CHC}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-% &\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
-% \end{array}\]
-% 
-% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r union s with Ins Delta into t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{INSERT}&\Delta\code{ INTO T};\\
-% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{CHC}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-% &\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
-% \end{array}\]
-% 
-% 	Section~\ref{ssct:Implication Rules} treated rules of the form $\code{R}\subs\code{S}$.
-% 	Note that this rule can be rewritten to $\cmpl{\code{R}}\cup\code{S}=\full$.
-% 	Consequently, rules did not have to be treated separately.
-% 	We could have treated them as a special case of equations with boolean operators only.
-% 	In fact, every rule with boolean operators $\cup$, $\cap$, and $-$ can be rewritten as a number of rules in disjunctive normal form\footnote{%
-% 	Normalizing a boolean algebra is existing knowledge, and is considered outside the scope of this paper.} (DNF~\cite{davey1990}).
-% 	We can represent a rule in DNF informally by:
-% \begin{eqnarray}
-% \code{RULE }\cmpl{r_0}\cup\ldots\cup\cmpl{r_n}\cup s_0\cup\ldots\cup s_m
-% \end{eqnarray}
-% 	or more formally by
-% \begin{eqnarray}
-% \full=\cmpl{\bigcap\id{antecedents}}\cup\bigcup\id{consequents}
-% \end{eqnarray}
-% 	in which \id{antecedents} are the complemented (negative) terms and \id{consequents} are the non-complemented (positive) terms.
-% 
-	To preserve invariants in a boolean algebra is done as follows:
-\[\begin{array}{|l|l|l|l|}
-\hline
-\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
-\code{R}=\cmpl{\code{S}}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE }\Delta\code{ FROM S}&\ref{lbl:Preserve r = -s with Ins Delta into r}\\
-\code{R}=\cmpl{\code{S}}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT }\Delta\code{ INTO S}&\ref{lbl:Preserve r = -s with Del Delta from r}\\
-\code{T}=\bigcap\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{ALL }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Ins Delta into t}\\
-\code{T}=\bigcup\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{ALL }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Del Delta from t}\\
-\code{T}=\bigcap\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{CHC }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Del Delta from t}\\
-\code{T}=\bigcup\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{CHC }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Ins Delta into t}\\ \hline
-\end{array}\]
-	So let us now turn to code that is needed when relational operators are involved.
-
-\subsection{Rules with relational operators}
-	This section studies how to insert pairs into a term $r;s$.
-	To formulate this precisely, we introduce an auxiliary relation $t$ to represent the term $r;s$.
-	First we will insert pairs in $t$ while preserving invariance of $t=r;s$
-	Then we will insert pairs in $r$ preserving the same rule.
-
-	To insert pairs $\Delta$ into $t$ involves finding relations $\Delta_r$ and $\Delta_s$
-	that satisfy equation~\ref{eqn:t subs r;s}.
-\begin{equation}
-t=r;s\ \Rightarrow\ t\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:t subs r;s}
-\end{equation}
-	Let us define a software primitive to compute both $\Delta_r$ and $\Delta_s$ and call it \code{NEWPAIRS}.
-	It computes $\Delta_r$ and $\Delta_s$ from $r$, $s$, and $\Delta$.
-	We can derive its postconditions from equation~\ref{eqn:t subs r;s} by means of the following derivation:
-\[\begin{array}{r@{~}l}
-&t\cup\Delta\\
-=&\hspace{1in}\{\text{since }t\cup\Delta=t\cup(\Delta-t)\}\\
-&t\cup(\Delta-t)\\
-=&\hspace{1in}\{t=r;s\text{ and let \code{NEWPAIRS} ensure that }\Delta_r;\Delta_s=\Delta-t\}\\
-&r;s\cup\Delta_r;\Delta_s\\
-=&\hspace{1in}\{\text{let \code{NEWPAIRS} ensure: }r;\Delta_s\subs t\cup\Delta\text{ and }\Delta_r;s\subs t\cup\Delta\}\\
-&r;s\cup r;\Delta_s\cup \Delta_r;s\cup \Delta_r;\Delta_s\\
-=&\hspace{1in}\{\text{distributivity}\}\\
-&(r\cup \Delta_r);(s\cup \Delta_s)
-\end{array}\]
-	The conditions in this derivation must be satisfied by \code{NEWPAIRS}. 
-	This yields the following specification of \code{NEWPAIRS}:
-\[\begin{array}{r@{~}l}
-&\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{DELTA}=\Delta\}\\
-\code{NEWPAIRS}&(\code{R}, \code{S}, \code{DELTA}): (\code{DELTAR}, \code{DELTAS})\\
-&\begin{array}[t]{@{}l@{}l}
-        \{&\code{DELTR};\code{DELTAS}=\Delta - r;s\text{ and }\\
-          &\code{R};\code{DELTAS}\subs r;s\cup\Delta\text{ and }\code{DELTAR};\code{S}\subs r;s\cup\Delta\}
-\end{array}
-\end{array}\]
-	The pre- and postconditions of \code{NEWPAIRS} are sufficient to satisfy equation~\ref{eqn:t subs r;s}.
-	Since $t$ does not occur in the pre- and postconditions, \code{NEWPAIRS} implements equation~\ref{eqn:ins delta r;s}:
-\begin{equation}
-r;s\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:ins delta r;s}
-\end{equation}
-	Note that the specification of \code{NEWPAIRS} does not determine its implementation uniquely.
-	There are multiple ways to implement \code{NEWPAIRS}.
-	If \code{NEWPAIRS} creates one new pair for every pair in $\Delta-t$,
-	we get $r;\Delta_s=\emptyset$ and $\Delta_r;s=\emptyset$.
-	The weaker condition from the derivation, i.e. $r;\Delta_s\subs t\cup\Delta$ and $\Delta_r;s\subs t\cup\Delta$ allows some reuse of existing pairs.
-	However, any implementation that preserves invariance of equation~\ref{eqn:ins delta r;s} will do.
-
-	At this point we have the ingredients for code that preserves invariance of $\code{T}=\code{R};\code{S}$
-\[\begin{array}{l}
-\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\text{ and }\code{T}=\code{R};\code{S}\}\\
-\code{INSERT }\Delta\code{ INTO T};\\
-\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}\neq\code{R};\code{S}\}\\
-(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R}, \code{S}, \Delta-\code{T});\\
-\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{ and }\\
-\hspace{0.575in}\code{DELTAR};\code{DELTAS}=\Delta-\code{T}\text{ and }\\
-\hspace{0.575in}\code{R};\code{DELTAS}\subs\code{T}\text{ and }\code{DELTAR};\code{S}\subs\code{T}\text{, so equation~\ref{eqn:ins delta r;s} is satisfied}\}\\
-\code{INSERT DELTAR INTO R};\\
-\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\}\\
-\code{INSERT DELTAS INTO S}\\
-\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\cup\code{DELTAS}\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}=\code{R};\code{S}\}
-\end{array}\]
-    This shows how to satisfy invariant $\code{T}=\code{R};\code{S}$ when pairs $\Delta$ are inserted into \code{T}.
-
-	Let us now study how to keep $r;s$ constant when $r$ changes.
-	Inserting pairs $\Delta$ into $r$ involves deleting pairs from $s$ such that:
-\begin{equation}
-(r\cup \Delta);(s-\Delta_s)=r;s\label{eqn:ins r del s}
-\end{equation}
-	In order to solve equation~\ref{eqn:ins r del s} towards $\Delta_s$,
-	we derive\footnote{Thanks to Peter Jipsen, who pointed out this line of reasoning.}:
-\begin{eqnarray}
-&&(r\cup \Delta);(s-\Delta_s)\ \subs\ r;s\label{eqn:rs subs t by ins r del s}\\
-&\Leftrightarrow&\nonumber\\
-&&s-\Delta_s\ \subs\ (r\cup \Delta)\backslash(r;s)\nonumber\\
-&\Leftrightarrow&\hspace{1in}\{\text{apply }x-y=z\ \Leftrightarrow\ x=z\cup y\text{ twice.}\}\nonumber\\
-&&s-(r\cup \Delta)\backslash(r;s)\ \subs\ \Delta_s\nonumber
-\end{eqnarray}
-	This derivation shows that every solution $\Delta_s$ contains $s-(r\cup \Delta)\backslash(r;s)$.
-	So the latter, $s-(r\cup \Delta)\backslash(r;s)$, is the smallest solution.
-
-	The smallest solution might not contain all pairs from $r;s$.
-	The missing pairs are $r;s-(r\cup\Delta);(s-(r\cup \Delta)\backslash(r;s))$.
-	Therefore it may be necessary to reintroduce these missing pairs afterwards.
-	Fortunately, we have just shown how to do just that with \code{NEWPAIRS}.
-	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
-\[\begin{array}{l}
-\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
-\code{INSERT }\Delta\code{ INTO R};\\
-\hspace{0.5in}\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}\neq\code{T}\}\\
-\code{DELETE R}\backslash\code{T FROM S};\\
-\hspace{0.5in}\{\text{Since }\code{R}=r\cup\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r\cup \Delta)\backslash(r;s)\\
-\hspace{0.575in}\text{So }\code{S}=s-(r\cup \Delta)\backslash(r;s)\text{ and therefore}\\
-\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
-(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}\\
-\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
-\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
-\end{array}\]
-
-	Let us now study how to delete pairs $\Delta$ from $r$ while keeping $r;s$ constant.
-	This involves inserting pairs into $s$ such that:
-\begin{equation}
-(r-\Delta);(s\cup\Delta_s)=r;s\label{eqn:del r ins s}
-\end{equation}
-	The largest $\Delta_s$ follows from the law $x;(x\backslash y)\subs y$ for $x=r-\Delta$ and $y=r;s$:
-\begin{eqnarray}
-&&(r-\Delta);((r-\Delta)\backslash(r;s))\ \subs\ r;s\nonumber\\
-&\Rightarrow&\hspace{1in}\{\text{Let }\Delta_s\ =\ (r-\Delta)\backslash(r;s)\}\nonumber\\
-&&(r-\Delta);\Delta_s\ \subs\ r;s\nonumber\\
-&\Leftrightarrow&\hspace{1in}\{p;x\subs p;q\ \Leftrightarrow\ p;(q\cup x)\subs p;q\}\nonumber\\
-&&(r-\Delta);(s\cup\Delta_s)\ \subs\ r;s\nonumber
-\end{eqnarray}
-	This derivation shows that we can choose $\Delta_s=(r-\Delta)\backslash(r;s)$ as the largest set that preserves invariant~\ref{eqn:del r ins s}.
-
-	Once again, we might not have all pairs from $r;s$,
-	so equation~\ref{eqn:del r ins s} needs to be completed with the missing pairs.
-	The missing pairs are $r;s-(r-\Delta);(s\cup(r-\Delta)\backslash(r;s))$.
-	In this case too, these missing pairs can be reintroduced afterwards.
-	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
-\[\begin{array}{l}
-\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
-\code{DELETE }\Delta\code{ FROM R};\\
-\hspace{0.5in}\{\code{R}=r-\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so possibly }\code{R};\code{S}\neq\code{T}\}\\
-\code{INSERT R}\backslash\code{T INTO S};\\
-\hspace{0.5in}\{\text{Since }\code{R}=r-\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r-\Delta)\backslash(r;s)\\
-\hspace{0.575in}\text{So }\code{S}=s\cup(r-\Delta)\backslash(r;s)\text{ and therefore}\\
-\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
-(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}\\
-\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
-\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
-\end{array}\]
-
-	Summarizing, we have obtained some more code to preserve invariants in a relation algebra:
-\[\begin{array}{|l|l|l|l|}
-\hline
-\text{invariants}&\text{action}&\text{reaction}&\text{equation}\\ \hline
-\begin{array}{l}
-\code{R};\code{S}\subs\code{T}\\
-\code{S}\subs\code{R}\backslash\code{T}\\
-\code{R}\subs\code{T}\slash\code{S}
-\end{array}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE R}\backslash\code{T FROM S}&\ref{eqn:rs subs t by ins r del s}\\ \hline
-\begin{array}{l}
-\code{R};\code{S}\subs\code{T}\\
-\code{S}\subs\code{R}\backslash\code{T}\\
-\code{R}\subs\code{T}\slash\code{S}
-\end{array}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO T}&
-\begin{array}{l}
-(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \Delta-\code{T});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}
-\end{array}&\ref{eqn:ins delta r;s}\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM T}&\code{Nop}&\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{Nop}&\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT R}\backslash\code{T INTO S}&\\ \hline
-\code{R};\code{S}=\code{T}&\code{INSERT}\ \Delta\code{ INTO R}&
-\begin{array}{l}
-\code{DELETE R}\backslash\code{T FROM S};\\
-(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}
-\end{array}&\ref{eqn:ins r del s}\\ \hline
-\code{R};\code{S}=\code{T}&\code{DELETE}\ \Delta\code{ FROM R}&
-\begin{array}{l}
-\code{INSERT R}\backslash\code{T INTO S};\\
-(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}
-\end{array}&\ref{eqn:del r ins s}\\ \hline
-\end{array}\]
-	The solutions given here are far from being exhaustive.
-	We illustrate this point by giving an example.
-	Suppose we have invariant $\code{T}\subs\code{R};\code{S}$ and action $\code{DELETE}\ \Delta\code{ FROM R}$.
-	So the question is which pairs must be added to $s$ after pairs $\Delta$ have been removed from $r$.
-	In normal circumstances, we might just look it up in the table above, picking $\code{INSERT R}\backslash\code{T INTO S}$ as a reaction.
-	In the circumstance that initially $\code{T}=r;s$, we might reason differently.
-	We must still find a smallest value for $\Delta_s$ that satisfies:
-\begin{equation}
-r;s\ \subs\ (r-\Delta);(s\cup\Delta_s)\label{lbl:nonshrinking r;s when Del from r}
-\end{equation}
-	The following derivation shows that $\flip{(r-\Delta)};(r\cap\Delta);s$ satisfies this requirement conditionally.
-	Here is the derivation:
-\begin{eqnarray}
-&&\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ \Delta_s\nonumber\\
-&\Rightarrow\nonumber\\
-&&(r-\Delta);\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Rightarrow&\hspace{1in}\{r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}\}\nonumber\\
-&&(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Leftrightarrow\nonumber\\
-&&(r-(r-\Delta));s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Rightarrow\nonumber\\
-&&(r;s)-((r-\Delta);s)\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Leftrightarrow\nonumber\\
-&&r;s\ \subs\ (r-\Delta);s \cup (r-\Delta);\Delta_s\nonumber\\
-&\Leftrightarrow\nonumber\\
-&&r;s\ \subs\ (r-\Delta);(s \cup \Delta_s)\nonumber
-\end{eqnarray}
-	This derivation justifies the reaction $\code{INSERT }\flip{(r-\Delta)};(r\cap\Delta);s\code{ INTO S}$,
-	but only if $r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}$.
-	The part $r;s;\flip{s};\flip{r}$ is not necessary in this proof, but it is useful in practice.
-	It says that $r-\Delta$ must contain a pair for every element in the domain of $r;s$,
-	instead of every element in $I$.
-	This makes this result applicable in more situations.
-
-\section{Reflection}
-\label{sct:Reflection}
-	Ampersand is being built to automate software development.
-	The holy grail is to derive deterministic ways to preserve invariants, to allow programming by constraints.
-	This is motivated by the benefits of describing a result in terms of constraints,
-	rather than having to invent all the steps that lead to that result.
-	For example, describing the situation that any financial transaction must preserve the total amount of money is simpler
-	than having to describe all the steps involved in such a transaction.
-	The attraction is not only for the programmer, but even more so for auditors and users for who the constraints are easier to grasp and validate.
-	In this paper we have shown how to derive specific steps,
-	helping the programmer to invent code and providing mathematical assurance of correctness (compliance-by-design).
-
-	The analysis in this paper shows once more that inventing algorithms for preserving invariants is far from trivial.
-	Algorithms cannot be derived from constraints alone. The choices that remain must be resolved first.
-	This paper does show, however, that these choices occur in predictable ways.
-	That gives opportunities, for instance to make these choices explicit to a user, to optimize the number of choices,
-	or to use external knowledge to resolve the remaining choices at run time.
-
-	In practice we have experienced that programming by constraints in relation algebra yields an unconventional way of programming.
-	This experience consists of inventing rules and choosing event types for resolving violations, as illustrated by the case study.
-	To relate this experience to programming as we know it, section~\ref{sct:Comparison} makes a comparison with established programming languages.
-	Section~\ref{sct:Contribution} summarizes the contributions to software development claimed by Ampersand.
-	Section~\ref{sct:Ampersand in practice} summarizes the use of Ampersand in practice and
-	section~\ref{sct:Further Research} gives an outlook on research that is required in the near future.
-
-\subsection{Comparison}
-\label{sct:Comparison}
-	This section compares Ampersand with existing programming paradigms by mentioning the most important differences and similarities.
-
-	The first to compare with is the object oriented programming paradigm, known from popular languages such as Java~\cite{Java} and C++~\cite{Stroustrup97a}.
-	Ampersand cannot be called object oriented, because the notion of control flow in object oriented languages is very different.
-	In those languages, the threads in the control flow are controlled by the software developer.
-	As shown in our case study, the control flow in an Ampersand script emerges as a result of changes in relation content as illustrated
-	by figure~\ref{fig:event flow}.
-
-	Ampersand also differs from the logic programming paradigm of Prolog~\cite{Lloyd1984}
-	and of all rule engines that can be considered to be offspring of Prolog.
-	A difference lies in the way rules are interpreted.
-	In logic programming, a program consists of Horn-clauses and a resolution-proof is constructed at run time.
-	The software developer works with notions such as backtracking (backward chaining) and unification, which are absent in Ampersand.
-	On the other hand, Ampersand is not restricted to implications with one predicate on the consequent side, as Prolog does.
-        Ampersand allows rules, which are equations with arbitrary relational expressions on either side of the equals sign.
-
-	Ampersand also differs from functional programming, of which Haskell and Scala are prominent representants.
-	A core idea in functional programming is to evaluate a program as a function from input to output~\cite{Backus1978}.
-	A functional program consists of function definitions, that are evaluated by a term- or graph-rewriter,
-	using various strategies such as lazy or eager evaluation.
-	In contrast, our case study illustrates that not everything has to be a function.
-	In relation algebra, everything is a relation and a function is a restricted form of a relation.
-	A similarity to functional programming is shown by the declarative style,
-	because substitution of equal terms without changing the semantics is a property we see in both worlds.
-
-	A difference with database programming is found in the type of algebra that is used.
-	Conventional databases are typically programmed in SQL, which is based on n-ary relations in relational algebra~\cite{Codd70}.
-	In contrast, Ampersand uses binary relations%
-\footnote{At runtime, Ampersand uses the efficiency of n-ary tables. It does so by translating relation algebra to efficient SQL-code by avoiding as many joins as possible.}
-	based on heterogeneous relation algebra~\cite{Schmidt1997}.
-	In this comparison, Ampersand is closer to graph databases~\cite{Vicknair2010} than to relational databases.
-
-	In the tradition of formal specification, there are many relational approaches, such as Z~\cite{Z}, CSP~\cite{CSP}, LOTOS~\cite{LOTOS}, VDM~\cite{VDM}.
-	Where formal specification techniques typically analyse and diagnose specifications,
-	Ampersand actually synthesizes (generates) information systems.
-
-	The similarity with the reactive programming paradigm is obvious from figure~\ref{fig:event flow}.
-	This graph is part of the dependency graph, which is a familiar phenomenon in reactive programming.
-	Ampersand may be seen as a reactive programming language on binary relations.
-	It has a unidirectional push based evaluation model, static typing, implicit lifting, and glitch avoidance~\cite{Bainomugisha2013}.
-	Even though Ampersand generates code to execute scripts,
-	code fragments described in section~\ref{sct:Code Fragments}
-	might just as well be generated towards a mature reactive programming language and executed there%
-\footnote{Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
-	but generation to a mature reactive programming language is definitely on the wishlist.}.
-	A distinguishing feature of Ampersand is that it treats constraints (rules) as its central notion,
-	which is also seen in TBAG~\cite{Elliott94} and FHM~\cite{Giorgidze2011} but not common in most reactive programming languages.
-	In the author's opinion a constraint based approach is preferrable, because the correctness of a program is within closer range.
-
-	If Ampersand represents a programming style at all, we might call it a relation-algebraic style of programming or maybe constraint programming\footnote{https://en.wikipedia.org/wiki/Constraint\_programming}.
-	Ampersand's style would be characterized by a programmer who specifies constraints and a computer trying to satisfy these constraints by resolving violations.
-
-\subsection{Contribution}
-\label{sct:Contribution}
-	Contributions of Ampersand to the software development process are:
-\begin{itemize}
-\item	Ampersand has the usual benefits of a declarative language:
-	This means that terms can be manipulated by Tarski's axioms without changing their semantics~\cite{vdWoude2011}.
-	It also means that the order in which rules are written has no consequence for their semantics.
-\item	Heterogeneous relation algebra has a straightforward interpretation in natural language~\cite{RBD}.
-	We have used that to formalize business requirements without exposing business stakeholders to any formal notation.
-\item	Ampersand is statically typed~\cite{Joosten2015}, which is an obvious consequence of being heterogeneous.
-	There is much evidence that static typing yields significantly lower software maintenance cost as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
-\item	Heterogeneous relation algebra is well studied.
-	As a consequence, many tools that are readily available in the public domain can be put to good use.
-	For executives of large organizations it can be reassuring that the formalism is well studied and understood.
-\item	Relation algebra facilitates composing software from reusable components, because a program consists of rules.
-	Since the union of sets of rules is a set of rules, compositionality comes from the union operator.
-	In practice, when components are brought together in larger assemblies, hardly any adjustments have to be made%
-\footnote{This is (unsubstantiated) experience collected from projects we have done with Ampersand.}.
-\end{itemize}
-
-	Ampersand also has a disadvantage. It appears to be difficult to learn for large groups of software professionals.
-	Research~\cite{Michels2015} shows that this is largely due to deficits in prerequisite knowledge, especially skills in discrete mathematics.
-%	For this reason, we have made an addendum to the course we teach at the Open University of the Netherlands.
-	Also, programming appears to be difficult in practice.
-	Ongoing research tries to find better ways for learning, which is necessary for large scale use.
-
-\subsection{Ampersand in Practice}
-\label{sct:Ampersand in practice}
-	Ampersand has been used in practice both in education (Open University of the Netherlands)
-	and in industry (Ordina and TNO).
-	For example, Ordina designed a proof-of-concept in 2007 of the INDIGO-system,
-	which is still in use today as the core information system of the Dutch immigration authority, IND.
-	The design of INDIGO was made with Ampersand, to obtain correct, detailed results in the least amount of time.
-	More recently, Ampersand was used to design an information system called DTV for the Dutch food authority, NVWA.
-	A prototype of DTV was built by Ampersand and was used as a model to build the actual system.
-	TNO, a major Dutch industrial research laboratory, is using Ampersand for research purposes.
-	For example, TNO did a study of international standardizations efforts such as
-	RBAC (Role Based Access Control) in 2003 and architecture (IEEE 1471-2000)~\cite{IEEE1471} in 2004.
-	Several inconsistencies were found in the last (draft) RBAC standard~\cite{RBAC}.
-	TNO has also used the technique in conceiving several patents%
-\footnote{e.g. patents DE60218042D, WO2006126875, EP1727327, WO2004046848, EP1563361, NL1023394C, EP1420323, WO03007571, and NL1013450C.}.
-	At the Open University of the Netherlands, Ampersand is being taught in a course called Rule Based Design~\cite{RBD}.
-	In this course, students use a platform called RAP, which has been built in Ampersand~\cite{Michels2015}.
-	RAP has been the first Ampersand-application that has run in production.
-
-\subsection{Further Research}
-\label{sct:Further Research}
-	Further research on this topic is required to bring relation algebra still closer to the community of practitioners.
-	Further use of relation algebra can be made by incorporating a model checker, such as the Alloy analyser~\cite{Alloy2006}, to detect inconsistent rules.
-	An exciting new development is Amperspiegel~\cite{Amperspiegel}, which brings notational flexibility to the fingertips of the user.
-	Developments in the Ampersand compiler are going towards a rule-repository (written in Ampersand itself).
-	This will make collaborative information systems development in Ampersand easier, because the repository can assist in automating
-	the software development process further.
-	Other research is needed towards a comprehensive theory of information systems.
-	Currently, there is no theory (in the mathematical meaning of the word) for information systems.
-	In the Ampersand project, a sub-project called ``Formal Ampersand'' is being conducted to achieve this goal.
-\subsection{Acknowledgements}
-\label{sct:Acknowledgements}
-	Firstly, I wish to thank Sebastiaan Joosten for introducing me to Isabelle and for inspiring me with his challenging critiques.
-	I owe many thanks to Wolfram Kahl, Michiel Stornebrink, Han Joosten, and the anonymous reviewers of RAMiCS for their useful feedback.
-	And I am very grateful to the Open University and Ordina for funding this research.
-
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[angle=90,scale=.357]{LogicalDataModel.pdf}
-\end{center}
-\caption{Conceptual data model (from section~\ref{sct:Conceptual analysis})}
-\label{fig:conceptual model}
-\end{figure}
+% \begin{figure}[htb]
+% \begin{center}
+%   \includegraphics[angle=90,scale=.357]{LogicalDataModel.pdf}
+% \end{center}
+% \caption{Conceptual data model (from section~\ref{sct:Conceptual analysis})}
+% \label{fig:conceptual model}
+% \end{figure}
 

### 2019-02-23 2018 Transactions/articleTransactions.tex
renewed sections 2 and 3

@@
 \hyphenation{ExecEngine}
+\def\id#1{\text{\em #1\/}}
+\def\Events{{\mathit E}}
 \begin{document}
@@
 \begin{abstract}
-	How to do transactions if data is scattered across the internet?
-	Users who must satisfy data integrity constraints are stuck with relational databases if they need transactions.
-	For most NoSQL databases lack transaction functionality.
-	This contribution proposes a theory of database transactions solely based on data,
-	with no reference to any database in which that data resides.
-	This makes transactions suitable for a wider class of databases than just relational ones.
-	The soundness of the theory has been established by means of category theory.
+	How can we make transactions work in distributed information systems without a shared database?
+	Transaction theory as we know it assumes a shared data space.
+	Multiple processes are acting simultaneously, each making changes to that shared data space.
+	The observation that transaction mechanisms in NoSQL databases are not common
+	makes us feel that a solution to this problem is not so obvious.
+
+	This contribution proposes a theory of database transactions without a shared data space.
+	This makes transactions suitable for more databases than just relational ones.
+	The soundness of the theory has been established by means of formal proof.
 	To show that the theory works, the theory is demonstrated by a prototype developed in Ampersand\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
@@
 \label{sct:Introduction}
+	A distributed information system (DIS) serves its users by registering data, guarding the consistency, protecting the data,
+	and providing that data accurately and meaningfully for its intended purpose.
+	Users want predictable behaviour when they are simultaneously changing a shared state (e.g.\ flight reservation).
+	For that purpose a transaction mechanism preserves system-wide invariants.
+	Many transaction mechanisms are known that presume a shared database [refs].
+	However, shared databases are not always an option in distributed cloud architectures.
+	Globally distributed applications such as Netflix and LinkedIn cannot afford to rely on a shared database for obvious reasons of performance and reliability.
+	Many cloud applications use distributed storage models instead.
+	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
+	These so-called NoSQL databases often lack functionality for transactions.
+	For the purpose of doing transactions in a non-transactional situation, several architectural patterns exist.
+	Some of these patterns simply circumvent the problem.
+	Others will isolate transactionality into a smaller sub-problem.
+	Still others introduce locking mechanisms to do the job.
+	Yet, in the world of event-driven distributed systems, a dire need is felt for a theory that brings what transactions do for shared databases.
+
+	This paper presents such a theory.
+	It describes transactions without the assumption of a shared database.
+	In this theory we express the characteristic properties of transactions: Atomicity, Consistency, Isolation, and Durability.
+	{\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
+	{\em Consistency} means that integrity rules are warranted by the database management system.
+	{\em Isolation} means that actions can be done without any knowledge of what other users are doing simultaneously.
+	{\em Durability} means that the contents of a database cannot change unintendedly.
+	To motivate this theory, let us formulate requirements to characterize a DIS:
+
+\section{Events}
+\label{sct:Events}
+	We need the notion of {em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a DIS.
+	Let us study the requirements for events in formal terms.
+	In this contribution we use variables $e$, $e_1$, and $e_2$ to represent events.
+	Let $\mathbb E$ represent the set of all conceivable events.
+
+	For the purpose of discussing history, we introduce a structure that represents a history at one moment in time.
+	A {\em historic structure} is a tuple $\la \Events, \id{prec}\ra$ in which
+	\begin{eqnarray}
+		&&\Events\text{ is a finite set of events.}\\
+		&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
+		&&\id{prec}^*\text{ is antisymmetric.}\label{req:precAntisymmetric}
+	\end{eqnarray}
+	Let set $\Events$ represent all events that the DIS has registered up to some point in time.
+	So for every event $e\in\Events$ we can say that it has occured (at some time in the past).
+	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has occured before event $e_2$,
+	making $\id{prec}$ a relation on events in $\Events$.
+	\marginpar{claim}
+	To represent time faithfully, we need an order on time that is a partial order,
+	i.e.\ an order which is reflexive, transitive, and antisymmetric.
+	For this reason requirement~\ref{req:precAntisymmetric} requires $\id{prec}^*$
+	(which is the reflexive and transitive closure of $\id{prec}$)
+	to be antisymmetric.
+	With these requirements in place,  $\la\Events,\id{prec}^*\ra$ is the partial order that we need to represent time.
+
+	An event $e$ that precedes another event is called a {\em past} event,
+	which we define as \ $\exists e_1: e\ \id{prec}\ e_1$.
+	All other events in a historic structure are called {\em current} events.
+	The special case in which $\id{prec}^*$ is a total order is called a {\em linear historic structure}.
+	Note that any nonempty linear historic structure has one current event.
+	Other historic structures may have more.
+
+	As time passes by, new events occur.
+	We describe each occurrence as a mapping, $\id{occur}_{P,e}$, from one historic structure to another:
+\begin{equation}
+	\id{occur}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
+\label{req:def occur}
+\end{equation}
+	The occurrence of $e$ adds one event to $\Events$.
+	It also enlarges relation $\id{prec}$ with one tuple for every immediate predecessor of $e$.
+	The immediate predecessors of $e$ are specified in set $P$, which we call the predecessor set.
+	Note that this enlargement can make current events become past events.
+	If $h'=\id{occur}_{P,e}(h)$, the events in $P$ become past events in $h'$ and event $e$ becomes current in $h'$.
+
+	We need the predecessor set to cater for the distributed nature of a DIS.
+	In a linear historic structure,
+\marginpar{claim}
+	$P$ is alway a singleton set.
+	In that case, every event except the first one has precisely one preceding event.
+	In general, however, events may have multiple preceding events and the predecessor set is not necessarily a singleton set.
+
+	Note that the definition of $\id{occur}_{P,e}$ (\ref{req:def occur}) must respect requirement~\ref{req:precInE}.
+	If $P$ would contain events that are not in $\Events$, these events appear in relation $\id{prec}$ and hence violate requirement~\ref{req:precInE}.
+\marginpar{claims}
+	So this requires that $P\subseteq\Events$.
+	Similarly, event $e$ cannot be an element of $\Events$, because that would violate the antisymmetry of $\id{prec}^*$ (requirement~\ref{req:precAntisymmetric}).
+	So in a specification style with pre- and postconditions%
+	\footnote{Also known as ``Floyd-Hoare style''}
+	we specify:
+\[\begin{array}{ll}
+	\text{pre:}&P\subseteq\Events\\
+	&e\notin\Events\\
+	&\id{Events}=\Events\\
+	&\id{pr}=\id{prec}\\[1em]
+	\multicolumn{2}{l}{\ \ \ \id{occur}_{P,e}\la \id{Events}, \id{pr}\ra}\\[1em]
+	\text{post:}&\id{Events}=\Events\cup\{e\}\\
+	&\id{pr}=\id{prec}\cup\{(e_1,e) | e_1\in P\}\\
+\end{array}\]
+
+	We define $\la \Events, \id{prec}\ra$ a {\em historic substructure} of $\la \Events', \id{prec}'\ra$
+	if $\Events\subseteq\Events'$ and $\id{prec}\subseteq\id{prec}'$.
+	We denote this as
+\begin{equation}
+	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra
+\end{equation}
+\marginpar{claim}
+	Note that $h'=\id{occur}_{P,e}(h)$ implies that $h\subseteq h'$.
+
+	Having defined historic structures and the way they grow, let us proceed to determine requirements for threads and states.
+	In the sequel, we will use variables $h$ and $h'$ to denote history structures.
+
+\section{States}
+	This section defines the requirements for predictable behavior of distributed information systems (DIS).
+	First we need to introduce the notion of state.
+	Let $\mathbb S$ represent the set of all conceivable states,
+	with one special state $\mathcal I$ that represents the initial state.
+	Let $\id{state}:\Events\rightarrow{\mathbb S}$ be a function that maps every event in a DIS to a state.
+	We pronounce $\id{state}(e)$ as ``the state immediately after event $e$ has occured'',
+	or shorter: ``the state after event $e$''.
+	A {\em current state} of a DIS is the state after a current event of that system
+	and a {\em past state} is the state after a past event.
+
+	Each time a new event occurs, a change of state takes place.
+	To accommodate such changes,
+	we use triples $(e_1,e,\Delta)$ instead of tuples $(e_1,e)$.
+	$\Delta$ represents the change from the state after event $e_1$ to the state after event $e$.
+	Besides an event $e$ and immediate predecessors $P$,
+	it takes a function $\delta$ to represent the change of state between two events:
+\begin{eqnarray}
+	\id{occur}_{P,e,\delta}\la \Events, \id{precd}\ra\ =\ \la \Events\cup\{e\}, \id{precd}\cup\{(e_1,e,\delta(e_1,e)) | e_1\in P\}\ra
+\end{eqnarray}
+	Additionally we require an operator $\oplus$ to compute the next state.
+	This provides the means to define a {\em stateful historic structure} as a tuple $\la \Events, \id{precd}\ra$ in which
+	$\id{precd}$ is a set of triples such that
+\begin{eqnarray}
+	&&\la \Events, \{(e_1,e)| (e_1,e,\Delta)\in\id{precd}\}\ra\text{ is a historic structure.}\\
+	&&(e_1,e,\Delta)\in\id{precd}\ \Rightarrow\ \id{state}(e)=\id{state}(e_1)\oplus\Delta\label{eqn:nextState}\\
+	&&(\exists e_1:(e_1,e,\Delta)\in\id{precd})\ \vee\ \id{state}(e)=\mathcal I\label{eqn:totState}\\
+	&&\begin{array}{@{}l}
+		(e_1,e,\Delta_1)\in\id{precd}\ \wedge\ (e_2,e,\Delta_2)\in\id{precd}\\
+		\ \ \Rightarrow\ \id{state}(e_1)\oplus\Delta_1\ =\ \id{state}(e_2)\oplus\Delta_2\label{eqn:uniState}
+	\end{array}
+\end{eqnarray}
+	Requirement~\ref{eqn:nextState} says that the state of non-initial events can be computed using $\oplus$.
+	Requirement~\ref{eqn:totState} says that the state of an initial event (i.e.\ events without predecessor) is the initial state.
+	Together, requirements~\ref{eqn:nextState} and~\ref{eqn:totState} define the function $\id{state}$ recursively.
+\marginpar{claim}
+	Requirement~\ref{eqn:uniState} ensures that $\id{state}$ is defined unambiguously,
+	meaning that every event in $\Events$ has precisely one state.
+	Thus, states are defined predictably in a DIS.
+
+	We define $\la \Events, \id{precd}\ra$ a {\em stateful historic substructure} of $\la \Events', \id{precd}'\ra$
+	if $\Events\subseteq\Events'$ and $\id{precd}\subseteq\id{precd}'$.
+	We denote this as
+\begin{equation}
+	\la \Events, \id{precd}\ra\ \subseteq\ \la \Events', \id{precd}'\ra
+\end{equation}
+\marginpar{claim}
+	Note that $\id{occur}_{P,e,\delta}(h)=h'$ implies that $h\subseteq h'$.
+
+	\marginpar{claim}
+	Can we prove $\oplus$ to be associative, commutative, and idempotent? Why would we do that?
+
+\section{Bibliography}
+\bibliographystyle{elsarticle-harv}
+\bibliography{doc}
+
+\end{document}
+
+	\section{Threads}
+	So how does a stateful historic structure represent a distributed information system?
+\marginpar{claim}
+	Part of the answer is that a substructure of a stateful historic structure is a stateful historic structure by itself.
+
+	The answer is that we have made no assumptions about location,
+	like for instance that there is a shared memory.
+	Even if we assume that each state exists in one location,
+	a stateful historic structure can still have multiple current states.
+	Each of these states can exist in a separate location.
+
+	The purpose of threads is to support deterministic computations.
+	We consider any nonempty historic structure that is totally ordered to be a thread.
+	A {\em thread} is a historic structure $\la \Events, \id{prec}\ra$ that satisfies
+\begin{eqnarray}
+	\forall e_1,e_2&:&e_1\in\Events\ \wedge\ e_2\in\Events\ \Rightarrow\ e_1\ \id{prec}^*\ e_2\vee e_2\ \id{prec}^*\ e_1\label{eqn:order is total}\\
+	\exists e&:&e\in\Events\label{eqn:nonempty thread}
+\end{eqnarray}
+	The total order (\ref{eqn:order is total}) is required to get deterministic results from a computation in a thread.
+	In order to talk about the first and the last event in a thread, we require at least one event (\ref{eqn:nonempty thread}) in a thread.
+	We shall use variable $t$ to represent threads.
+	The first event in a thread $t$ is the infimum of the events in $t$;
+	the last event is the supremum.
+
+	We need the notion of state%
+	\footnote{Cambridge Dictionary meaning of ``state'':
+	a condition or way of being that exists at a particular time}
+	to describe things that change over time.
+	We shall use variables $s$, $s_1$, and $s_2$ to denote states.
+	Let $\id{state}(e)$ represent the state after event $e$ has occured.
+
+\section{Distributed Information System}
+
+	Now equation~\ref{eqn:eval step} is the recursion step and the initial state is
+	the starting point of the recursive computation of the final state of $t$.
+	We consider a distributed information system as a set of processes that share a history.
+	The system is distributed because processes need not share all histories.
+	We need the notion of process%
+	\footnote{Cambridge Dictionary meaning of ``process'':
+		a series of actions that you take in order to achieve a result}
+	to describe how components of a DIS interact.
+	Let $\mathbb P$ be the set of all processes.
+	We have formulated the following requirements:
+	\begin{itemize}
+	\item Belong/Contain\\
+		To represent the fact that an event {\em belongs to} a process or
+		(equivalently) that a process {\em contains} certain events,
+		let $e\in p$ represent the statement that event $e$ belongs to process $p$.
+	\item Shared history\\
+		To abstract away from location, we make no assumptions with respect to the place where a process lives.
+		However, to partake in a distributed system, processes must be able to share history.
+		For that purpose we define what it means to be part of a history:
+		\begin{equation}
+			p\ \id{partof}\ h\ \Leftrightarrow\ (\forall e: e\in p\Rightarrow e\in h)
+		\end{equation}
+		We define an order between processes that share a history.
+		\begin{equation}
+			p_1\ \preceq\ p_2\ \Leftrightarrow(\forall e: e\in p_1\Rightarrow e\in p_2)
+		\end{equation}
+		We can prove that
+		\begin{equation}
+			p_1\ \preceq\ p_2\ \Rightarrow\ (\exists h:p_1\ \id{partof}\ h\wedge p_2\ \id{partof}\ h)
+		\end{equation}
+	\item Total order\\
+		We require a process to have predictable behaviour, for which we need the process to contain a series (sequence) of events.
+		So we require that $(\{e|e\in p\}, \preceq)$ is a total order.
+		Apart from being a partial order, $\preceq$ must satisfy (for every $e_1$, $e_2$, and $p$):
+		\begin{equation}
+			e_1\in p\ \wedge\ e_2\in p\ \Rightarrow\ e_1\preceq e_2\vee e_2\preceq e_1
+		\end{equation}
+	\end{itemize}
+
+\section{States}
+	\begin{itemize}
+	\item Revert\\
+		A change can be used to compute a previous state.
+		For this purpose we require an operator $\ominus$ such that
+	\begin{equation}
+		s_1\oplus\Delta=s_2\ \Leftrightarrow\ s_1=s_2\ominus\Delta
+	\end{equation}
+		As a consequence, we have for every event $e$:
+	\begin{equation}
+		\id{state}(\id{past}(e))=\id{state}(\id{pres}(e))\ominus\id{delta}(e)
+	\end{equation}
+		With $\preceq$ being reflexive, antisymmetric, and transitive,
+		it follows immediately that $\oplus$ is idempotent, commutative, and associative.
+	\end{itemize}
+
+\section{Atomicity}
+\label{sct:Atomicity}
+	{\em Atomicity} means that multiple primitive actions are executed as one.
+	So let us first define the notion of atomic action.
+
+\section{Consistency}
+\label{sct:Consistency}
+	{\em Consistency} means that integrity rules are warranted by the database management system.
+
+	If $s$ is a state and $r$ is an integrity rule, then $s\ \id{sat}\ r$ means that $s$ satisfies rule $r$
+	and $s\ \overline{\id{sat}}\ r$ means that $s$ does not satisfy rule $r$.
+
+	Let us call the set of integrity rules $\mathcal R$.
+	A state $s$ is called consistent if it satisfies every integrity rule $r\in\mathcal R$
+\begin{equation}
+	\id{consistent}(s)\ =\ \forall r\in\mathcal R: s\ \id{sat}\ r
+\end{equation}
+	The system is consistent if the initial state is consistent and it can only reach consistent states by means of atomic actions.
+	So all states in the history graph are consistent with $\mathcal R$.
+
+\section{Isolation}
+\label{sct:Isolation}
+{\em Isolation} means that actions can be done without any knowledge of what other users are doing simultaneously.
+
+\section{Durability}
+\label{sct:Durability}
+{\em Durability} means that the contents of a database cannot change unintendedly.
+
+\section{Bibliography}
+\bibliographystyle{elsarticle-harv}
+\bibliography{doc}
+
+\end{document}
+
+%% Obsolete
 
@@
 	The order (in time) of events is given by a relation\footnote{%
-A relation $\declare{r}{A}{B}$ is a set of tuples $\pair{a}{b}$ with $a\in A$ and $b\in B$} $\declare{pred}{Event}{Event}$,
+A relation $\declare{r}{A}{B}$ is a set of tuples. Each tuple $\pair{a}{b}$ in that set satisfies $a\in A$ and $b\in B$}
+	$\declare{pred}{Event}{Event}$,
 	where $\pair{e1}{e2}\in\id{pred}$ means that event $e1$ precedes event $e2$.

### 2019-02-24 2018 Transactions/articleTransactions.tex
Include the proven lemma 1 and its proof in Isabelle

@@
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage[english]{babel}
@@
 \hyphenation{ExecEngine}
+\newtheorem{lemma}{Lemma}
 \def\id#1{\text{\em #1\/}}
@@
 
+\section{Related work}
+\label{sct:Related work}
+\cite{Foster17c}
+
 \section{Events}
 \label{sct:Events}
-	We need the notion of {em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a DIS.
+	We need the notion of {\em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a DIS.
 	Let us study the requirements for events in formal terms.
-	In this contribution we use variables $e$, $e_1$, and $e_2$ to represent events.
+	In this work we use variables $e$, $e_1$, and $e_2$ to represent events.
 	Let $\mathbb E$ represent the set of all conceivable events.
@@
 	&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
-		&&\id{prec}^*\text{ is antisymmetric.}\label{req:precAntisymmetric}
+	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
+	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
 \end{eqnarray}
@@
 	making $\id{prec}$ a relation on events in $\Events$.
-	\marginpar{claim}
-	To represent time faithfully, we need an order on time that is a partial order,
-	i.e.\ an order which is reflexive, transitive, and antisymmetric.
-	For this reason requirement~\ref{req:precAntisymmetric} requires $\id{prec}^*$
-	(which is the reflexive and transitive closure of $\id{prec}$)
-	to be antisymmetric.
-	With these requirements in place,  $\la\Events,\id{prec}^*\ra$ is the partial order that we need to represent time.
+	For this reason, requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has occured before itself.
+	To represent time faithfully, we require that $\la\Events,\id{prec}^*\ra$ is a partially ordered set,
+	meaning that $\id{prec}^*$ is reflexive, transitive, and antisymmetric.
+	For this reason equation~\ref{req:precPlus asy} requires the transitive closure $\id{prec}^+$ to be antisymmetric.
 
-	An event $e$ that precedes another event is called a {\em past} event,
-	which we define as \ $\exists e_1: e\ \id{prec}\ e_1$.
-	All other events in a historic structure are called {\em current} events.
+	In a historic structure $\la \Events, \id{prec}\ra$, an event $e$ that precedes another event is called a {\em past} event,
+	which we define as \ $\exists e_1\in\Events: e\ \id{prec}\ e_1$.
+	All other events are called {\em current} events in that historic structure.
+
 	The special case in which $\id{prec}^*$ is a total order is called a {\em linear historic structure}.
@@
 	The occurrence of $e$ adds one event to $\Events$.
-	It also enlarges relation $\id{prec}$ with one tuple for every immediate predecessor of $e$.
+	It also enlarges relation $\id{prec}$ with one pair for every immediate predecessor of $e$.
 	The immediate predecessors of $e$ are specified in set $P$, which we call the predecessor set.
-	Note that this enlargement can make current events become past events.
-	If $h'=\id{occur}_{P,e}(h)$, the events in $P$ become past events in $h'$ and event $e$ becomes current in $h'$.
-
 	We need the predecessor set to cater for the distributed nature of a DIS.
 	In a linear historic structure,
-\marginpar{claim}
 	$P$ is alway a singleton set.
@@
 	In general, however, events may have multiple preceding events and the predecessor set is not necessarily a singleton set.
+\begin{lemma}
+\label{lemma:occur preserves historic structure}
+	If $h=\la \Events, \id{prec}\ra$ is a historic structure and $P\subseteq\Events$ and $e\notin\Events$
+	then $\id{occur}_{P,e}(h)$ is a historic structure.
+\end{lemma}
+	This lemma has been proven in proof assistant Isabelle/HOL.
+	To understand why $P\subseteq\Events$ is necessary,
+	let $\la \Events', \id{prec}'\ra=\id{occur}_{P,e}(h)$.
+	Now suppose event $e_1\in P$ but $e_1\notin\Events$.
+	Equation~\ref{req:def occur} tells us that $(e_1,e)\in\id{prec}'$.
+	However, this violates requirement~\ref{req:precInE}.
+	So it is necessary that every $e_1\in P$ is an element of $\Events$.
+	Similarly, $e\notin\Events$ is a necessary condition,
+	because any $e\in\Events$ violates the antisymmetry of $\id{prec}^+$ (requirement~\ref{req:precPlus asy}).
 
-	Note that the definition of $\id{occur}_{P,e}$ (\ref{req:def occur}) must respect requirement~\ref{req:precInE}.
-	If $P$ would contain events that are not in $\Events$, these events appear in relation $\id{prec}$ and hence violate requirement~\ref{req:precInE}.
-\marginpar{claims}
-	So this requires that $P\subseteq\Events$.
-	Similarly, event $e$ cannot be an element of $\Events$, because that would violate the antisymmetry of $\id{prec}^*$ (requirement~\ref{req:precAntisymmetric}).
-	So in a specification style with pre- and postconditions%
-	\footnote{Also known as ``Floyd-Hoare style''}
-	we specify:
-\[\begin{array}{ll}
-	\text{pre:}&P\subseteq\Events\\
-	&e\notin\Events\\
-	&\id{Events}=\Events\\
-	&\id{pr}=\id{prec}\\[1em]
-	\multicolumn{2}{l}{\ \ \ \id{occur}_{P,e}\la \id{Events}, \id{pr}\ra}\\[1em]
-	\text{post:}&\id{Events}=\Events\cup\{e\}\\
-	&\id{pr}=\id{prec}\cup\{(e_1,e) | e_1\in P\}\\
-\end{array}\]
+	Note that the occurence of an event can make current events become past events.
+	If $h'=\id{occur}_{P,e}(h)$, the events in $P$ become past events in $h'$ and event $e$ becomes current in $h'$.
 
-	We define $\la \Events, \id{prec}\ra$ a {\em historic substructure} of $\la \Events', \id{prec}'\ra$
-	if $\Events\subseteq\Events'$ and $\id{prec}\subseteq\id{prec}'$.
-	We denote this as
+	Let $h$ and $h'$ be historic structures.
+	We use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em historic substructure} of $h'$.
+	This is defined as:
 \begin{equation}
-	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra
+\label{def:historic substructure}
+	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra\ \Leftrightarrow\ \Events\subseteq\Events'\ \wedge\ \id{prec}\subseteq\id{prec}'
 \end{equation}
-\marginpar{claim}
-	Note that $h'=\id{occur}_{P,e}(h)$ implies that $h\subseteq h'$.
+	Note that $\id{occur}_{P,e}(h)=h'$ implies that $h\subseteq h'$.
 
-	Having defined historic structures and the way they grow, let us proceed to determine requirements for threads and states.
+	Having defined historic structures and the way they grow, let us proceed to determine requirements for states.
 	In the sequel, we will use variables $h$ and $h'$ to denote history structures.
@@
 \begin{eqnarray}
-	\forall e_1,e_2&:&e_1\in\Events\ \wedge\ e_2\in\Events\ \Rightarrow\ e_1\ \id{prec}^*\ e_2\vee e_2\ \id{prec}^*\ e_1\label{eqn:order is total}\\
+	\forall e_1,e_2&:&e_1\in\Events\ \wedge\ e_2\in\Events\ \Rightarrow\ e_1\ \id{prec}+\ e_2\vee e_2\ \id{prec}^+\ e_1\label{eqn:order is total}\\
 	\exists e&:&e\in\Events\label{eqn:nonempty thread}

### 2019-03-02 2018 Transactions/articleTransactions.tex
Straightened out the text with one question for @sjc
Sebastiaan, please take a look at the claim that state:E->S is well
defined?

@@
 \begin{eqnarray}
-	&&\Events\text{ is a finite set of events.}\\
 	&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
@@
 	Let set $\Events$ represent all events that the DIS has registered up to some point in time.
-	So for every event $e\in\Events$ we can say that it has occured (at some time in the past).
-	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has occured before event $e_2$,
+	So for every event $e\in\Events$ we can say that it has happened (at some time in the past).
+	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
 	making $\id{prec}$ a relation on events in $\Events$.
-	For this reason, requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has occured before itself.
+	For this reason, requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
 	To represent time faithfully, we require that $\la\Events,\id{prec}^*\ra$ is a partially ordered set,
@@
 
-	As time passes by, new events occur.
-	We describe each occurrence as a mapping, $\id{occur}_{P,e}$, from one historic structure to another:
+	We describe each time an event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one historic structure to another:
 \begin{equation}
-	\id{occur}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
-\label{req:def occur}
+	\id{happens}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
+\label{req:def happens}
 \end{equation}
-	The occurrence of $e$ adds one event to $\Events$.
-	It also enlarges relation $\id{prec}$ with one pair for every immediate predecessor of $e$.
+	This happening adds one event to $\Events$ and enlarges relation $\id{prec}$ with one pair for every immediate predecessor of $e$.
 	The immediate predecessors of $e$ are specified in set $P$, which we call the predecessor set.
 	We need the predecessor set to cater for the distributed nature of a DIS.
+
+	If a DIS is initialized with an empty historic structure
+	and it grows by means of $\id{happens}$, i.e. by one event at a time,
+	then at any moment in time it consists of a finite number of events.
+	Therefore, we can safely assume that set $\Events$ is finite in every practical situation.
+
 	In a linear historic structure,
@@
 \begin{lemma}
-\label{lemma:occur preserves historic structure}
+\label{lemma:happens preserves historic structure}
 	If $h=\la \Events, \id{prec}\ra$ is a historic structure and $P\subseteq\Events$ and $e\notin\Events$
-	then $\id{occur}_{P,e}(h)$ is a historic structure.
+	then $\id{happens}_{P,e}(h)$ is a historic structure.
 \end{lemma}
+	Lemma~\ref{lemma:happens preserves historic structure} says that $\id{happens}_{P,e}$ preserves the historic structure.
 	This lemma has been proven in proof assistant Isabelle/HOL.
 	To understand why $P\subseteq\Events$ is necessary,
-	let $\la \Events', \id{prec}'\ra=\id{occur}_{P,e}(h)$.
+	let $\la \Events', \id{prec}'\ra=\id{happens}_{P,e}(h)$.
 	Now suppose event $e_1\in P$ but $e_1\notin\Events$.
-	Equation~\ref{req:def occur} tells us that $(e_1,e)\in\id{prec}'$.
+	Equation~\ref{req:def happens} tells us that $(e_1,e)\in\id{prec}'$.
 	However, this violates requirement~\ref{req:precInE}.
@@
 
-	Note that the occurence of an event can make current events become past events.
-	If $h'=\id{occur}_{P,e}(h)$, the events in $P$ become past events in $h'$ and event $e$ becomes current in $h'$.
+	Note that $\id{happens}_{P,e}$ can make current events become past events.
+	If $h'=\id{happens}_{P,e}(h)$, every event in $P$ becomes a past event in $h'$ and event $e$ becomes current in $h'$.
 
@@
 \end{equation}
-	Note that $\id{occur}_{P,e}(h)=h'$ implies that $h\subseteq h'$.
+	Note that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.
 
@@
 \section{States}
-	This section defines the requirements for predictable behavior of distributed information systems (DIS).
-	First we need to introduce the notion of state.
+	This section defines the requirements for predictable behavior of a distributed information system (DIS).
+	We consider behavior to be predictable if every state can be computed unambiguously from its history.
+	So let us introduce the notion of state.
+
 	Let $\mathbb S$ represent the set of all conceivable states,
 	with one special state $\mathcal I$ that represents the initial state.
-	Let $\id{state}:\Events\rightarrow{\mathbb S}$ be a function that maps every event in a DIS to a state.
-	We pronounce $\id{state}(e)$ as ``the state immediately after event $e$ has occured'',
-	or shorter: ``the state after event $e$''.
-	A {\em current state} of a DIS is the state after a current event of that system
+	A {\em stateful historic structure} (SHS) is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
+	$\la \Events, \id{prec}\ra$ is a historic structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
+	that maps every event in $\Events$ to a state.
+	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
+	Or shorter: $s$ is the state after event $e$.
+	Alternatively, we may write $(s,e)\in\id{state}$ because a function is a univalent and total relation
+	(and therefore a set of pairs).
+	A {\em current state} is the state after a current event
 	and a {\em past state} is the state after a past event.
 
-	Each time a new event occurs, a change of state takes place.
-	To accommodate such changes,
-	we use triples $(e_1,e,\Delta)$ instead of tuples $(e_1,e)$.
-	$\Delta$ represents the change from the state after event $e_1$ to the state after event $e$.
-	Besides an event $e$ and immediate predecessors $P$,
-	it takes a function $\delta$ to represent the change of state between two events:
+	In a stateful historic structure, the occurrence of an event $e$ requires that $\id{state}(e)$ exists.
+	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another:
+\begin{equation}
+\begin{array}{l}
+	\id{occurs}_{P,e,s}(\la \Events, \id{prec}, \id{state}\ra)\\
+	\ =\ \la \Events\cup\{e\},\ \id{prec}\cup\{(e_1,e) | e_1\in P\},\ \id{state}\cup\{(e,s)\}\ra
+\end{array}	
+\label{eqn:def occurs}
+\end{equation}
+	In fact, this definition expands the definition of historic structures (eqn.~\ref{req:def happens}).
+	By using lemma~\ref{lemma:happens preserves historic structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
+	the stateful historic structure.
+	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
+	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
+
+	Let $\mathbb D$ represent the set of all conceivable state changes.
+	Let $s_2\ominus s_1$ represent the change from state $s_1$ to state $s_2$.
+	Let $\delta(e_1,e_2)$ represent the change $\id{state}(e_2)\ominus\id{state}(e_1)$.
+
+	Each time a new event happens, a change of state takes place.
+	This change is described in the following requirements:
 \begin{eqnarray}
-	\id{occur}_{P,e,\delta}\la \Events, \id{precd}\ra\ =\ \la \Events\cup\{e\}, \id{precd}\cup\{(e_1,e,\delta(e_1,e)) | e_1\in P\}\ra
+	&&(\exists e_2:(e_1,e_2)\in\id{prec})\ \vee\ \id{state}(e_1)=\mathcal I\label{eqn:totState}\\
+	&&(e_1,e_2)\in\id{prec}\ \Rightarrow\ \id{state}(e_2)=\id{state}(e_1)\oplus\delta(e_1,e_2)\label{eqn:nextState}
 \end{eqnarray}
-	Additionally we require an operator $\oplus$ to compute the next state.
-	This provides the means to define a {\em stateful historic structure} as a tuple $\la \Events, \id{precd}\ra$ in which
-	$\id{precd}$ is a set of triples such that
-\begin{eqnarray}
-	&&\la \Events, \{(e_1,e)| (e_1,e,\Delta)\in\id{precd}\}\ra\text{ is a historic structure.}\\
-	&&(e_1,e,\Delta)\in\id{precd}\ \Rightarrow\ \id{state}(e)=\id{state}(e_1)\oplus\Delta\label{eqn:nextState}\\
-	&&(\exists e_1:(e_1,e,\Delta)\in\id{precd})\ \vee\ \id{state}(e)=\mathcal I\label{eqn:totState}\\
-	&&\begin{array}{@{}l}
-		(e_1,e,\Delta_1)\in\id{precd}\ \wedge\ (e_2,e,\Delta_2)\in\id{precd}\\
-		\ \ \Rightarrow\ \id{state}(e_1)\oplus\Delta_1\ =\ \id{state}(e_2)\oplus\Delta_2\label{eqn:uniState}
-	\end{array}
-\end{eqnarray}
+	Requirement~\ref{eqn:totState} says that the state of an initial event (i.e.\ events without predecessor) is the initial state.
 	Requirement~\ref{eqn:nextState} says that the state of non-initial events can be computed using $\oplus$.
-	Requirement~\ref{eqn:totState} says that the state of an initial event (i.e.\ events without predecessor) is the initial state.
-	Together, requirements~\ref{eqn:nextState} and~\ref{eqn:totState} define the function $\id{state}$ recursively.
+	Together, requirements~\ref{eqn:totState} and~\ref{eqn:nextState} define the function $\id{state}$ recursively.
 \marginpar{claim}
-	Requirement~\ref{eqn:uniState} ensures that $\id{state}$ is defined unambiguously,
-	meaning that every event in $\Events$ has precisely one state.
-	Thus, states are defined predictably in a DIS.
-
-	We define $\la \Events, \id{precd}\ra$ a {\em stateful historic substructure} of $\la \Events', \id{precd}'\ra$
-	if $\Events\subseteq\Events'$ and $\id{precd}\subseteq\id{precd}'$.
-	We denote this as
-\begin{equation}
-	\la \Events, \id{precd}\ra\ \subseteq\ \la \Events', \id{precd}'\ra
-\end{equation}
-\marginpar{claim}
-	Note that $\id{occur}_{P,e,\delta}(h)=h'$ implies that $h\subseteq h'$.
-
-	\marginpar{claim}
-	Can we prove $\oplus$ to be associative, commutative, and idempotent? Why would we do that?
+	So for every finite SHS that has been constructed from the initial state by means of event occurrences,
+	the function $\id{state}$ is defined.
+	This ensures that every event in $\Events$ has precisely one state, which means that states are defined predictably in a DIS.
 
@@
 	We shall use variables $s$, $s_1$, and $s_2$ to denote states.
-	Let $\id{state}(e)$ represent the state after event $e$ has occured.
+	Let $\id{state}(e)$ represent the state after event $e$ has happened.
 
@@
 	The items seen by the user at any moment are given by a relation $\declare{state}{Event}{Item}$,
-	where $\pair{e}{i}\in\id{state}$ means that item $i$ exists immediately after $e$ has occured.
-	So the set of items visible to the user immediately after $e$ has occured can be computed by $\{ i | \pair{e}{i}\in\id{state} \}$.
+	where $\pair{e}{i}\in\id{state}$ means that item $i$ exists immediately after $e$ has happened.
+	So the set of items visible to the user immediately after $e$ has happened can be computed by $\{ i | \pair{e}{i}\in\id{state} \}$.
 	Note that the relation $\id{state}$ is interpreted in an open world.
@@
 	More precisely, let us consider an event $e$.
-	The items $\{ i | \pair{e}{i}\in\id{state} \}$ are the items that exist immediately after the preceding event has occured,
+	The items $\{ i | \pair{e}{i}\in\id{state} \}$ are the items that exist immediately after the preceding event has happened,
 	minus the items to be removed, plus the items to be inserted.

### 2019-03-04 2018 Transactions/articleTransactions.tex
introducing transactions

@@
 	For this reason equation~\ref{req:precPlus asy} requires the transitive closure $\id{prec}^+$ to be antisymmetric.
+	In the sequel, we will use predicate $\id{HS}\la \Events, \id{prec}\ra$
+	to denote that $\la \Events, \id{prec}\ra$ is a historic structure
+	and we will use variables $h$ and $h'$ to denote historic structures.
 
@@
 	Other historic structures may have more.
+	In a linear historic structure, every event except the first one has precisely one preceding event.
+	In general, however, events may have multiple preceding events and multiple succeeding events.
 
@@
 
-	If a DIS is initialized with an empty historic structure
-	and it grows by means of $\id{happens}$, i.e. by one event at a time,
-	then at any moment in time it consists of a finite number of events.
-	Therefore, we can safely assume that set $\Events$ is finite in every practical situation.
-
-	In a linear historic structure,
-	$P$ is alway a singleton set.
-	In that case, every event except the first one has precisely one preceding event.
-	In general, however, events may have multiple preceding events and the predecessor set is not necessarily a singleton set.
 \begin{lemma}
@@
 
+	Together with an empty historic structure, $\la \emptyset, \emptyset\ra$,
+	we can define a historic structure recursively by means of $\id{happens}$:
+\begin{eqnarray}
+	&&\id{HS}\la \emptyset, \emptyset\ra\label{def:zeroHS}\\
+&&\begin{array}{@{}l}
+	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{HS}\la \Events, \id{prec}\ra\ \Rightarrow\ \id{HS}(\id{happens}_{P,e}\la \Events, \id{prec}\ra)
+\end{array}
+\label{req:recursive def happens}
+\end{eqnarray}
+	Equation~\ref{def:zeroHS} says that $\la \emptyset, \emptyset\ra$ is a historic structure.
+	Equation~\ref{req:recursive def happens} specifies recursively that $\id{happens}_{P,e}(h)$ is a historic structure iff
+	$h$ is a historic structure and if $P$ and $e$ satisfy the specified conditions.
+	This definition allows us to prove properties of finite historic structures by induction.
+	In practical situations, every historic structure is finite.
+
 	Note that $\id{happens}_{P,e}$ can make current events become past events.
@@
 	Having defined historic structures and the way they grow, let us proceed to determine requirements for states.
-	In the sequel, we will use variables $h$ and $h'$ to denote history structures.
 
@@
 
-	Let $\mathbb S$ represent the set of all conceivable states,
-	with one special state $\mathcal I$ that represents the initial state.
-	A {\em stateful historic structure} (SHS) is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
+	Let $\mathbb S$ represent the set of all conceivable states.
+	A {\em stateful historic structure} is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
 	$\la \Events, \id{prec}\ra$ is a historic structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
 	that maps every event in $\Events$ to a state.
+	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
+	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful historic structure.
 	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
 	Or shorter: $s$ is the state after event $e$.
-	Alternatively, we may write $(s,e)\in\id{state}$ because a function is a univalent and total relation
+	Instead of $s=\id{state}(e)$, we may also write $(s,e)\in\id{state}$ because a function is a univalent and total relation
 	(and therefore a set of pairs).
@@
 
-	In a stateful historic structure, the occurrence of an event $e$ requires that $\id{state}(e)$ exists.
-	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another:
+	In a stateful historic structure, the occurrence of an event $e$ defines a new historic structure.
+	This is the way systems evolve over time.
+	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another SHS:
 \begin{equation}
@@
 	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
+	To describe the growth of a stateful historic structure we can define stateful historic structures recursively:
+\begin{eqnarray}
+	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
+&&\begin{array}{@{}l}
+	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
+	\ \ \Rightarrow\ \id{SHS}(\id{occurs}_{P,e,s}\la \Events, \id{prec}, \id{state}\ra)
+\end{array}
+\label{req:recursive def occurs}
+\end{eqnarray}
 
-	Let $\mathbb D$ represent the set of all conceivable state changes.
-	Let $s_2\ominus s_1$ represent the change from state $s_1$ to state $s_2$.
-	Let $\delta(e_1,e_2)$ represent the change $\id{state}(e_2)\ominus\id{state}(e_1)$.
+\section{Transactions}
+\label{sct:Transactions}
+	Information systems can register changes (as opposed to states) to avoid duplicating most of a previous state into a new state.
+	Such changes are called {\em transactions}.
+	For the purpose of computing a new state ($s_2$) from a given state ($s_1$),
+	we need two operators, $\oplus$ and $\ominus$, such that
+\begin{equation}
+	s_2\ =\ s_1\oplus(s_2\ominus s_1)
+\label{req:oplus and ominus}
+\end{equation}
+	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
+	In a stateful historic structure we can represent transactions by means of a function $\delta$,
+	where $\delta(e_1,e_2)$ represents the change of state from event $e_1$ to $e_2$.
+\begin{equation}
+	\delta(e_1,e_2)\ =\ \id{state}(e_2)\ominus\id{state}(e_1)
+\label{def:delta}
+\end{equation}
+	The transaction caused by event $e$ can be described by:
+\begin{equation}
+	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
+		\id{state}(e)\ =\ s&\wedge\\
+		\forall e_1\in P: s\ =\ \id{state}(e_1)\oplus\delta(e_1,e)
+	\end{array}
+\label{req:transaction}
+\end{equation}
+	We can define transactions recursively, in which we designate one specific state,
+	$\mathcal I\in\mathbb S$,
+	as initial state.
+\begin{equation}
+\begin{array}{l}
+	e\notin\Events\ \wedge\ P\subseteq\Events\ \Rightarrow\\
+	\id{trans}_{P,e_1,e,\Delta}(\la\Events, \id{prec}, \id{state}\ra)\\
+	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
+		\text{Let }s_1&=&\id{state}(e_1)\text{,}&\text{if }e_1\in P\\
+				&=&\mathcal I\text{,}&\text{otherwise}\\
+		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{state}\cup\{(e,s_1\oplus\Delta)\}\ra}
+		\end{array}
+\end{array}
+\label{req:recursive transaction}
+\end{equation}
 
-	Each time a new event happens, a change of state takes place.
-	This change is described in the following requirements:
-\begin{eqnarray}
-	&&(\exists e_2:(e_1,e_2)\in\id{prec})\ \vee\ \id{state}(e_1)=\mathcal I\label{eqn:totState}\\
-	&&(e_1,e_2)\in\id{prec}\ \Rightarrow\ \id{state}(e_2)=\id{state}(e_1)\oplus\delta(e_1,e_2)\label{eqn:nextState}
-\end{eqnarray}
-	Requirement~\ref{eqn:totState} says that the state of an initial event (i.e.\ events without predecessor) is the initial state.
-	Requirement~\ref{eqn:nextState} says that the state of non-initial events can be computed using $\oplus$.
-	Together, requirements~\ref{eqn:totState} and~\ref{eqn:nextState} define the function $\id{state}$ recursively.
-\marginpar{claim}
-	So for every finite SHS that has been constructed from the initial state by means of event occurrences,
-	the function $\id{state}$ is defined.
-	This ensures that every event in $\Events$ has precisely one state, which means that states are defined predictably in a DIS.
+\section{Atomicity}
+\label{sct:Atomicity}
+	{\em Atomicity} means that a state change caused by a sequence of events can be reproduced as a single event.
+
+	So let us first discuss event sequences.
+	Let us define an event sequence as a totally ordered historic structure.
+	Note that a historic structure that is not totally ordered may contain multiple historic substructures
+	that are totally ordered.
+	Typically, event sequences occur multiply in a historic structure.
 
@@
 	\end{itemize}
-
-\section{Atomicity}
-\label{sct:Atomicity}
-	{\em Atomicity} means that multiple primitive actions are executed as one.
-	So let us first define the notion of atomic action.
 

### 2019-03-05 2018 Transactions/articleTransactions.tex
Added Sebastiaan as author
Elaborate on transactions and introduce the incremental history structure

@@
 \ead{stef.joosten@ou.nl}
+\author[utwente]{Sebastiaan Joosten\fnref{fn2}}
 \address[ou]{Open Universiteit Nederland, Postbus 2960, 6401 DL Heerlen, the Netherlands}
 \address[ordina]{Ordina NV, Nieuwegein, the Netherlands}
+\address[utwente]{University of Twente, Enschede, the Netherlands}
 \fntext[fn1]{ORCID 0000-0001-8308-0189}
+\fntext[fn2]{ORCID 0000-0002-6590-6220}
 
@@
 	For the purpose of discussing history, we introduce a structure that represents a history at one moment in time.
-	A {\em historic structure} is a tuple $\la \Events, \id{prec}\ra$ in which
+	A {\em history structure} is a tuple $\la \Events, \id{prec}\ra$ in which
 \begin{eqnarray}
@@
 	In the sequel, we will use predicate $\id{HS}\la \Events, \id{prec}\ra$
-	to denote that $\la \Events, \id{prec}\ra$ is a historic structure
-	and we will use variables $h$ and $h'$ to denote historic structures.
+	to denote that $\la \Events, \id{prec}\ra$ is a history structure
+	and we will use variables $h$ and $h'$ to denote history structures.
 
-	In a historic structure $\la \Events, \id{prec}\ra$, an event $e$ that precedes another event is called a {\em past} event,
+	In a history structure $\la \Events, \id{prec}\ra$, an event $e$ that precedes another event is called a {\em past} event,
 	which we define as \ $\exists e_1\in\Events: e\ \id{prec}\ e_1$.
-	All other events are called {\em current} events in that historic structure.
+	All other events are called {\em current} events in that history structure.
 
-	The special case in which $\id{prec}^*$ is a total order is called a {\em linear historic structure}.
-	Note that any nonempty linear historic structure has one current event.
-	Other historic structures may have more.
-	In a linear historic structure, every event except the first one has precisely one preceding event.
+	The special case in which $\id{prec}^*$ is a total order is called a {\em linear history structure}.
+	Note that any nonempty linear history structure has one current event.
+	Other history structures may have more.
+	In a linear history structure, every event except the first one has precisely one preceding event.
 	In general, however, events may have multiple preceding events and multiple succeeding events.
 
-	We describe each time an event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one historic structure to another:
+	We describe each time an event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one history structure to another:
 \begin{equation}
@@
 \begin{lemma}
-\label{lemma:happens preserves historic structure}
-	If $h=\la \Events, \id{prec}\ra$ is a historic structure and $P\subseteq\Events$ and $e\notin\Events$
-	then $\id{happens}_{P,e}(h)$ is a historic structure.
+\label{lemma:happens preserves history structure}
+	If $h=\la \Events, \id{prec}\ra$ is a history structure and $P\subseteq\Events$ and $e\notin\Events$
+	then $\id{happens}_{P,e}(h)$ is a history structure.
 \end{lemma}
-	Lemma~\ref{lemma:happens preserves historic structure} says that $\id{happens}_{P,e}$ preserves the historic structure.
+	Lemma~\ref{lemma:happens preserves history structure} says that $\id{happens}_{P,e}$ preserves the history structure.
 	This lemma has been proven in proof assistant Isabelle/HOL.
@@
 
-	Together with an empty historic structure, $\la \emptyset, \emptyset\ra$,
-	we can define a historic structure recursively by means of $\id{happens}$:
+	Together with an empty history structure, $\la \emptyset, \emptyset\ra$,
+	we can define a history structure recursively by means of $\id{happens}$:
 \begin{eqnarray}
@@
 \end{eqnarray}
-	Equation~\ref{def:zeroHS} says that $\la \emptyset, \emptyset\ra$ is a historic structure.
-	Equation~\ref{req:recursive def happens} specifies recursively that $\id{happens}_{P,e}(h)$ is a historic structure iff
-	$h$ is a historic structure and if $P$ and $e$ satisfy the specified conditions.
-	This definition allows us to prove properties of finite historic structures by induction.
-	In practical situations, every historic structure is finite.
+	Equation~\ref{def:zeroHS} says that $\la \emptyset, \emptyset\ra$ is a history structure.
+	Equation~\ref{req:recursive def happens} specifies recursively that $\id{happens}_{P,e}(h)$ is a history structure iff
+	$h$ is a history structure and if $P$ and $e$ satisfy the specified conditions.
+	This definition allows us to prove properties of finite history structures by induction.
+	In practical situations, every history structure is finite.
 
@@
 
-	Let $h$ and $h'$ be historic structures.
+	Let $h$ and $h'$ be history structures.
 	We use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em historic substructure} of $h'$.
@@
 
-	Having defined historic structures and the way they grow, let us proceed to determine requirements for states.
+	Having defined history structures and the way they grow, let us proceed to determine requirements for states.
 
@@
 	Let $\mathbb S$ represent the set of all conceivable states.
-	A {\em stateful historic structure} is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
-	$\la \Events, \id{prec}\ra$ is a historic structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
+	A {\em stateful history structure} is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
+	$\la \Events, \id{prec}\ra$ is a history structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
 	that maps every event in $\Events$ to a state.
 	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
-	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful historic structure.
+	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
 	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
@@
 
-	In a stateful historic structure, the occurrence of an event $e$ defines a new historic structure.
+	In a stateful history structure, the occurrence of an event $e$ defines a new history structure.
 	This is the way systems evolve over time.
@@
 \end{equation}
-	In fact, this definition expands the definition of historic structures (eqn.~\ref{req:def happens}).
-	By using lemma~\ref{lemma:happens preserves historic structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
-	the stateful historic structure.
+	This definition expands the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
+	By using lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
+	the stateful history structure.
 	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
 	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
-	To describe the growth of a stateful historic structure we can define stateful historic structures recursively:
+	To describe the growth of a stateful history structure we can define stateful history structures recursively:
 \begin{eqnarray}
@@
 	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
-	In a stateful historic structure we can represent transactions by means of a function $\delta$,
+	In a stateful history structure we can represent transactions by means of a function $\delta$,
 	where $\delta(e_1,e_2)$ represents the change of state from event $e_1$ to $e_2$.
@@
 	The transaction caused by event $e$ can be described by:
-\begin{equation}
-	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
+\begin{eqnarray}
+&&	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
 		\id{state}(e)\ =\ s&\wedge\\
@@
 \label{req:transaction}
-\end{equation}
-	We can define transactions recursively, in which we designate one specific state,
-	$\mathcal I\in\mathbb S$,
-	as initial state.
-\begin{equation}
-\begin{array}{l}
-	e\notin\Events\ \wedge\ P\subseteq\Events\ \Rightarrow\\
-	\id{trans}_{P,e_1,e,\Delta}(\la\Events, \id{prec}, \id{state}\ra)\\
+\end{eqnarray}
+	We can define a stateful history structure,
+	in which the state after every event is computed from a change, $\Delta$.
+	For this purpose we must designate one specific state, $\mathcal I\in\mathbb S$, as the initial state.
+	We must additionally specify an existing event $e_1$ in order to compute $\id{state}(e_1)\oplus\Delta$.
+	The following mapping $\id{occurByDelta}_{P,e_1,e,\Delta}$ defines how an SHS grows by one event ($e$)
+	using only the change between the state of event $e_1$ to the new event $e$:
+\begin{eqnarray}
+&&\begin{array}{l}
+	\id{occurByDelta}_{P,e_1,e,\Delta}(\la\Events, \id{prec}, \id{state}\ra)\\
 	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
@@
 \end{array}
-\label{req:recursive transaction}
-\end{equation}
+\label{req:SHS by Delta}
+\end{eqnarray}
+	This mapping serves as recursion step in the following definition of a stateful history structure
+\begin{eqnarray}
+	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
+&&\begin{array}{@{}l}
+	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
+	\ \ \Rightarrow\ \id{SHS}(\id{occurByDelta}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \id{state}\ra)
+\end{array}
+\label{req:recursive def SHS by Delta}
+\end{eqnarray}
+
+	At this point, let us reflect on storage.
+	If we store a stateful history structure entirely,
+	one state is added to the storage for every event in $\Events$.
+	However, if changes are on average much smaller than states, it can be more economical to store only the changes.
+	The following mapping transforms a stateful history structure to an incremental history structure:
+\begin{eqnarray}
+&&\begin{array}{l}
+	\id{shs2ihs}(\la\Events, \id{prec}, \id{state}\ra)\\
+	\ \ =\ \begin{array}[t]{@{}l}
+		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \delta(e_1,e_2)=\id{state}(e_2)\ominus\id{state}(e_1)\\
+		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
+		\text{Let }\forall e\in\id{initial}:\ \iota(e)=\id{state}(e)\ominus\mathcal I\\
+		\text{in }\la\Events, \id{prec}, \delta, \iota\ra
+		\end{array}
+\end{array}
+\label{req:shs2ihs}
+\end{eqnarray}
+	If we want to build an IHS event-by-event, we need an incremental definition.
+	For this purpose we define an {\em incremental history structure} ($\id{IHS}$)
+	as a history structure that ``stores'' changes instead of states.
+	The following mapping, $\id{increment}_{P,e_1,e,\Delta,\iota}$,
+	adds one event ($e$) to an IHS:
+\begin{eqnarray}
+\begin{array}{l}
+	\id{increment}_{P,e_1,e,\Delta}\la\Events, \id{prec}, \delta, \iota\ra\\
+	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
+		\text{Let }\delta'(x,y)&=&\Delta\text{, }&\text{if }x=e_1\wedge y=e\wedge x\in P\\
+				       &=&(\id{state}(e_1)\oplus\Delta)\ominus\id{state}(x)\text{, }&\text{if }x\neq e_1\wedge y=e\wedge x\in P\\
+				       &=&\delta(x,y)\text{, }&\text{if }(x,y)\in\id{prec}\\
+		\text{Let }\iota'(x)&=&\Delta\text{, }&\text{if }x=e\wedge e_1\notin P\\
+				    &=&\iota(x)\text{, }&\text{if }x\neq e\wedge e_1\notin P\\
+			\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \delta', \iota'\}\ra}
+		\end{array}
+\end{array}
+\label{req:increment}
+\end{eqnarray}
+	This mapping can be used as recursive step in the following (recursive) definition of predicate $\id{IHS}$, which characterizes an incremental history structure:
+\begin{eqnarray}
+	&&\id{IHS}\la \emptyset, \emptyset, \emptyset, \emptyset\ra\\
+&&\begin{array}{@{}l}
+	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{IHS}\la \Events, \id{prec}, \delta, \iota\ra\\
+	\ \ \Rightarrow\ \id{IHS}(\id{increment}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \delta, \iota\ra)
+\end{array}
+\label{req:recursive def IHS}
+\end{eqnarray}
 
@@
 	So let us first discuss event sequences.
-	Let us define an event sequence as a totally ordered historic structure.
-	Note that a historic structure that is not totally ordered may contain multiple historic substructures
+	Let us define an event sequence as a totally ordered history structure.
+	Note that a history structure that is not totally ordered may contain multiple historic substructures
 	that are totally ordered.
-	Typically, event sequences occur multiply in a historic structure.
+	Typically, event sequences occur multiply in a history structure.
 
@@
 	\section{Threads}
-	So how does a stateful historic structure represent a distributed information system?
+	So how does a stateful history structure represent a distributed information system?
 \marginpar{claim}
-	Part of the answer is that a substructure of a stateful historic structure is a stateful historic structure by itself.
+	Part of the answer is that a substructure of a stateful history structure is a stateful history structure by itself.
 
@@
 	Even if we assume that each state exists in one location,
-	a stateful historic structure can still have multiple current states.
+	a stateful history structure can still have multiple current states.
 	Each of these states can exist in a separate location.
@@
 	The purpose of threads is to support deterministic computations.
-	We consider any nonempty historic structure that is totally ordered to be a thread.
-	A {\em thread} is a historic structure $\la \Events, \id{prec}\ra$ that satisfies
+	We consider any nonempty history structure that is totally ordered to be a thread.
+	A {\em thread} is a history structure $\la \Events, \id{prec}\ra$ that satisfies
 \begin{eqnarray}

### 2019-03-05 2018 Transactions/articleTransactions.tex
minor improvements

@@
 \cite{Foster17c}
+
+	An interesting read is {\tt\small https://towardsdatascience.com/distributed-transactions-and-why-you-should-care-116b6da8d72}.
+	It explains how modern distributed database cope with ACID properties in a was that makes sense.
 
@@
 	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
-	To describe the growth of a stateful history structure we can define stateful history structures recursively:
+	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
+	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
+	To describe the growth of a stateful history structure by the event,
+	we define $\id{SHS}$ recursively using $\id{occurs}_{P,e,s}$ as recurring step:
 \begin{eqnarray}
@@
 \end{equation}
-	The transaction caused by event $e$ can be described by:
+	A transaction caused by event $e$ can be described by:
 \begin{eqnarray}
@@
 	If we store a stateful history structure entirely,
-	one state is added to the storage for every event in $\Events$.
-	However, if changes are on average much smaller than states, it can be more economical to store only the changes.
-	The following mapping transforms a stateful history structure to an incremental history structure:
-\begin{eqnarray}
-&&\begin{array}{l}
-	\id{shs2ihs}(\la\Events, \id{prec}, \id{state}\ra)\\
-	\ \ =\ \begin{array}[t]{@{}l}
-		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \delta(e_1,e_2)=\id{state}(e_2)\ominus\id{state}(e_1)\\
-		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
-		\text{Let }\forall e\in\id{initial}:\ \iota(e)=\id{state}(e)\ominus\mathcal I\\
-		\text{in }\la\Events, \id{prec}, \delta, \iota\ra
-		\end{array}
-\end{array}
-\label{req:shs2ihs}
-\end{eqnarray}
-	If we want to build an IHS event-by-event, we need an incremental definition.
-	For this purpose we define an {\em incremental history structure} ($\id{IHS}$)
+	one state is added to the total storage requirement for every event in $\Events$.
+	However, if the average change is typically smaller than the average state, it can be more economical to store only the changes.
+	For this purpose we define an incremental history structure (IHS)
 	as a history structure that ``stores'' changes instead of states.
+	A {\em incremental history structure} is a quadruple $\la\Events, \id{prec}, \delta, \iota\ra$ in which
+	$\la \Events, \id{prec}\ra$ is a history structure,
+	$\delta(e_1,e_2)$ represents the change $\id{state}(e_2)\ominus\id{state}(e_1)$ for every pair $(e_1,e_2)\in\id{prec}$ (equation~\ref{def:delta}),
+	and $\iota(e)$ represents the change $\id{state}(e_1)\ominus\mathcal I$ for every event $e\in\Events$ that has no predecessor in $\id{prec}$.
+	Since histories are built event-by-event, we give an incremental definition.
 	The following mapping, $\id{increment}_{P,e_1,e,\Delta,\iota}$,
@@
 \end{eqnarray}
-	This mapping can be used as recursive step in the following (recursive) definition of predicate $\id{IHS}$, which characterizes an incremental history structure:
+	This definition inserts a new event, $e$, into a history structure with $e_1$ as its predecessor.
+	The new state, $\id{state}(e)$, is computed by $\id{state}(e_1)\oplus\Delta$; let us call it $s$.
+	Beside $e_1$, there may be other events preceding $e$ in the predecessor set $P$.
+	The change from every predecessor $x$ can be computed by $s\ominus\id{state}(x)$.
+	All these changes are added to the function $\delta$.
+	The function $\iota$ is there for initialization purposes.
+	To get an initial history structure (with no events in it) going,
+	we must add events while there are no other events available as predecessor.
+	We use the empty predecessor set as condition to signal that there is no previous state.
+	The mapping $\id{increment}_{P,e_1,e,\Delta}$ can be used as recursive step in defining predicate $\id{IHS}$,
+	which characterizes an incremental history structure:
 \begin{eqnarray}
@@
 \end{eqnarray}
+The following mapping transforms a stateful history structure to an incremental history structure:
+\begin{eqnarray}
+&&\begin{array}{l}
+	\id{shs2ihs}(\la\Events, \id{prec}, \id{state}\ra)\\
+	\ \ =\ \begin{array}[t]{@{}l}
+		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \delta(e_1,e_2)=\id{state}(e_2)\ominus\id{state}(e_1)\\
+		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
+		\text{Let }\forall e\in\id{initial}:\ \iota(e)=\id{state}(e)\ominus\mathcal I\\
+		\text{in }\la\Events, \id{prec}, \delta, \iota\ra
+		\end{array}
+\end{array}
+\label{req:shs2ihs}
+\end{eqnarray}
+	Vice versa, we can also map an incremental history structure to a stateful history structure:
+\begin{eqnarray}
+&&\begin{array}{l}
+	\id{ihs2shs}(\la\Events, \id{prec}, \delta, \iota\ra)\\
+	\ \ =\ \begin{array}[t]{@{}l}
+		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \id{state}(e_2)=\id{state}(e_1)\oplus\delta(e_1,e_2)\\
+		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
+		\text{Let }\forall e\in\id{initial}:\ \id{state}(e)=\mathcal I\oplus\iota(e)\\
+		\text{in }\la\Events, \id{prec}, \id{state}\ra
+		\end{array}
+\end{array}
+\label{req:ihs2shs}
+\end{eqnarray}
+	The\marginpar{claim} fact that $\id{shs2ihs}$ and $\id{ihs2shs}$ are each other's inverse proves that the choice to represent
+	history by an IHS or SHS can be made on practical grounds.
 
@@
 \label{sct:Atomicity}
-	{\em Atomicity} means that a state change caused by a sequence of events can be reproduced as a single event.
+	{\em Atomicity} guarantees all-or-nothing outcomes.
+	It means that the overall state change caused by a sequence of events can be committed as a whole or rejected completely.
 

### 2019-03-17 2018 Transactions/articleTransactions.tex
WIP

@@
 \label{sct:Introduction}
-	A distributed information system (DIS) serves its users by registering data, guarding the consistency, protecting the data,
+	A distributed information system (DIS) serves its users with things such as
+	registering data, guarding the consistency, protecting the data,
 	and providing that data accurately and meaningfully for its intended purpose.
@@
 
-	An interesting read is {\tt\small https://towardsdatascience.com/distributed-transactions-and-why-you-should-care-116b6da8d72}.
-	It explains how modern distributed database cope with ACID properties in a was that makes sense.
+	An interesting read is~\cite{Foster17c}\footnote{\tt\tiny https://towardsdatascience.com/distributed-transactions-and-why-you-should-care-116b6da8d72}.
+	It explains how modern distributed database cope with ACID properties in a way that makes sense.
 
-\section{Events}
-\label{sct:Events}
+	ACID 2.0 - Associative, Commutative, Idempotent, Distributed
+
+\section{History}
+\label{sct:History}
+	Distributed information systems are dynamic systems,
+	so time is the first aspect we focus on.
+	In this paper we simplify the notion of time to precedence of events,
+	allowing us to study the evolution over time by events that happen.
 	We need the notion of {\em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a DIS.
-	Let us study the requirements for events in formal terms.
 	In this work we use variables $e$, $e_1$, and $e_2$ to represent events.
 	Let $\mathbb E$ represent the set of all conceivable events.
-
-	For the purpose of discussing history, we introduce a structure that represents a history at one moment in time.
+	We also introduce the notion of {\em history structure}
+	to represent all events that have happend up to some point in time and the precedences between those events.
+	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.
+	Let $\mathbb H$ be the set of all conceivable history structures.
 	A {\em history structure} is a tuple $\la \Events, \id{prec}\ra$ in which
@@
 	making $\id{prec}$ a relation on events in $\Events$.
-	For this reason, requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
+	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
 	To represent time faithfully, we require that $\la\Events,\id{prec}^*\ra$ is a partially ordered set,
@@
 	For this reason equation~\ref{req:precPlus asy} requires the transitive closure $\id{prec}^+$ to be antisymmetric.
-	In the sequel, we will use predicate $\id{HS}\la \Events, \id{prec}\ra$
-	to denote that $\la \Events, \id{prec}\ra$ is a history structure
-	and we will use variables $h$ and $h'$ to denote history structures.
+
+	The special case in which $\la\Events,\id{prec}^*\ra$ is a total order is called a {\em sequential history structure}.
+	Note that any nonempty finite sequential history structure has one current event and one first event.
+	Other finite history structures may have more.
+	In a sequential history structure, every event except the first one has precisely one preceding event.
+	In general, however, events may have multiple preceding events and multiple succeeding events.
 
@@
 	All other events are called {\em current} events in that history structure.
+	Events in $\Events$ that have no preceding events are called {\em first} events.
 
-	The special case in which $\id{prec}^*$ is a total order is called a {\em linear history structure}.
-	Note that any nonempty linear history structure has one current event.
-	Other history structures may have more.
-	In a linear history structure, every event except the first one has precisely one preceding event.
-	In general, however, events may have multiple preceding events and multiple succeeding events.
-
+\section{Growth}
+\label{sct:Growth}
+	A DIS grows by adding events to it.
 	We describe each time an event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one history structure to another:
 \begin{equation}
-	\id{happens}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
+\begin{array}{l}
+	P\subseteq\Events\ \wedge\ e\notin E\\
+	\ \ \Rightarrow\ \ \id{happens}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
+\end{array}
 \label{req:def happens}
@@
 	We need the predecessor set to cater for the distributed nature of a DIS.
-
 \begin{lemma}
 \label{lemma:happens preserves history structure}
-	If $h=\la \Events, \id{prec}\ra$ is a history structure and $P\subseteq\Events$ and $e\notin\Events$
-	then $\id{happens}_{P,e}(h)$ is a history structure.
+	If $\la \Events, \id{prec}\ra$ is a history structure and $P\subseteq\Events$ and $e\notin\Events$
+	then $\id{happens}_{P,e}(\la \Events, \id{prec}\ra)$ is a history structure.
 \end{lemma}
-	Lemma~\ref{lemma:happens preserves history structure} says that $\id{happens}_{P,e}$ preserves the history structure.
+	Lemma~\ref{lemma:happens preserves history structure} is a preservation property.
+	It says that $\id{happens}_{P,e}$ preserves the history structure,
+	provided that $P\subseteq\Events$ and $e\notin\Events$.
 	This lemma has been proven in proof assistant Isabelle/HOL.
@@
 
-	Together with an empty history structure, $\la \emptyset, \emptyset\ra$,
-	we can define a history structure recursively by means of $\id{happens}$:
+	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
+	Let us represent the fact that $h$ is a {\em historic substructure} of $h'$ by:
+\begin{equation}
+\label{def:historic substructure}
+	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra\ \Leftrightarrow\ \Events\subseteq\Events'\ \wedge\ \id{prec}\subseteq\id{prec}'
+\end{equation}
+	From lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.
+
+	Let us define predicate $\id{HS}\la \Events, \id{prec}\ra$
+	to denote that $\la \Events, \id{prec}\ra$ is a history structure.
 \begin{eqnarray}
@@
 \end{array}
-\label{req:recursive def happens}
+\label{def:HS}
 \end{eqnarray}
 	Equation~\ref{def:zeroHS} says that $\la \emptyset, \emptyset\ra$ is a history structure.
-	Equation~\ref{req:recursive def happens} specifies recursively that $\id{happens}_{P,e}(h)$ is a history structure iff
+	Equation~\ref{def:HS} specifies recursively that $\id{happens}_{P,e}(h)$ is a history structure iff
 	$h$ is a history structure and if $P$ and $e$ satisfy the specified conditions.
@@
 	Note that $\id{happens}_{P,e}$ can make current events become past events.
-	If $h'=\id{happens}_{P,e}(h)$, every event in $P$ becomes a past event in $h'$ and event $e$ becomes current in $h'$.
+	If $h'=\id{happens}_{P,e}(h)$, every event in $P$ is a past event in $h'$ and event $e$ becomes current in $h'$.
 
-	Let $h$ and $h'$ be history structures.
-	We use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em historic substructure} of $h'$.
-	This is defined as:
+	Let us reconsider the relation $\id{prec}$ for a moment and define $\preceq$ as follows:
+	Requirement~\ref{req:precPlus asy} from section~\ref{sct:History} ensures that $\preceq$ is reflexive, antisymmetric, and transitive.
+	Let us introduce operator $\curlyvee$ (the {\em meet} of two events)
 \begin{equation}
-\label{def:historic substructure}
-	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra\ \Leftrightarrow\ \Events\subseteq\Events'\ \wedge\ \id{prec}\subseteq\id{prec}'
+	e\ =\ e_1\curlyvee e_2\ \ \Leftrightarrow\ \ e\ \preceq\ e_1\ \wedge\ e\ \preceq\ e_2
 \end{equation}
-	Note that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.
+	If $\la \Events, \id{prec}, \id{state}\ra$ satisfies definition~\ref{req:recursive def occurs},
+	we can prove
+\begin{equation}
+		e_1,e_2\in\Events\ \Rightarrow\ (e_1\curlyvee e_2)\in\Events
+\end{equation}
+	This means that $\la\Events,\preceq\ra$ is an order-theoretic meet-semilattice.
+	From order theory, we know that $\curlyvee$ is idempotent, commutative, and associative.
 
@@
 	We consider behavior to be predictable if every state can be computed unambiguously from its history.
-	So let us introduce the notion of state.
+	For that purpose we introduce the notion of state.
+
+	A history structure can be perceived as the (global) state of an information system at any point in time,
+	even though that global state is observable only in its local parts.
 
@@
 \end{equation}
-	This definition expands the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
+	This definition extends the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
 	By using lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
@@
 \label{sct:Transactions}
-	Information systems can register changes (as opposed to states) to avoid duplicating most of a previous state into a new state.
+	In practice, a new state is computed from the current state by specifying a change, $\Delta$.
 	Such changes are called {\em transactions}.
-	For the purpose of computing a new state ($s_2$) from a given state ($s_1$),
-	we need two operators, $\oplus$ and $\ominus$, such that
+	To describe a DIS in terms of changes, we need operators to compute a new state ($s_2$) from a given state ($s_1$).
+	In fact, we need two operators, $\oplus$ and $\ominus$, such that
 \begin{equation}
@@
 	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
+	This difference represents a change.
+	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.
 	In a stateful history structure we can represent transactions by means of a function $\delta$,
@@
 	We must additionally specify an existing event $e_1$ in order to compute $\id{state}(e_1)\oplus\Delta$.
-	The following mapping $\id{occurByDelta}_{P,e_1,e,\Delta}$ defines how an SHS grows by one event ($e$)
+	The following mapping $\id{occurByDelta}_{P,e_1,\Delta,e}$ defines how an SHS grows by one event ($e$)
 	using only the change between the state of event $e_1$ to the new event $e$:
@@
 &&\begin{array}{l}
-	\id{occurByDelta}_{P,e_1,e,\Delta}(\la\Events, \id{prec}, \id{state}\ra)\\
+	\id{occurByDelta}_{P,e_1,\Delta,e}(\la\Events, \id{prec}, \id{state}\ra)\\
 	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
@@
 	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
-	\ \ \Rightarrow\ \id{SHS}(\id{occurByDelta}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \id{state}\ra)
+	\ \ \Rightarrow\ \id{SHS}(\id{occurByDelta}_{P,e_1,\Delta,e}\la \Events, \id{prec}, \id{state}\ra)
 \end{array}
@@
 \end{eqnarray}
-
-	At this point, let us reflect on storage.
-	If we store a stateful history structure entirely,
-	one state is added to the total storage requirement for every event in $\Events$.
-	However, if the average change is typically smaller than the average state, it can be more economical to store only the changes.
-	For this purpose we define an incremental history structure (IHS)
-	as a history structure that ``stores'' changes instead of states.
-	A {\em incremental history structure} is a quadruple $\la\Events, \id{prec}, \delta, \iota\ra$ in which
-	$\la \Events, \id{prec}\ra$ is a history structure,
-	$\delta(e_1,e_2)$ represents the change $\id{state}(e_2)\ominus\id{state}(e_1)$ for every pair $(e_1,e_2)\in\id{prec}$ (equation~\ref{def:delta}),
-	and $\iota(e)$ represents the change $\id{state}(e_1)\ominus\mathcal I$ for every event $e\in\Events$ that has no predecessor in $\id{prec}$.
-	Since histories are built event-by-event, we give an incremental definition.
-	The following mapping, $\id{increment}_{P,e_1,e,\Delta,\iota}$,
-	adds one event ($e$) to an IHS:
+	In practice, information systems build states incrementally.
+	Each event comes with an increment ($\Delta$) that changes the current state.
+	For this purpose we define an incremental history structure (IHS).
+	A {\em incremental history structure} is a quadruple $\la\Events, \id{prec}, \delta, \iota\ra$ in which:
+\begin{itemize}
+\item	$\la \Events, \id{prec}\ra$ is a history structure;
+\item	$\delta(e_1,e_2)$ represents a change, $\id{state}(e_2)\ominus\id{state}(e_1)$, for every pair $(e_1,e_2)\in\id{prec}$ (equation~\ref{def:delta});
+\item	$\iota(e)$ represents the change $\id{state}(e_1)\ominus\mathcal I$ for every event $e\in\Events$ that has no predecessor in $\id{prec}$.
+\end{itemize}
+	Since histories are built event-by-event, we define IHSs incrementally.
+	For that purpose we use the following mapping, $\id{increment}_{P,e_1,e,\Delta,\iota}$,
+	which adds one event ($e$) to an IHS:
 \begin{eqnarray}
-\begin{array}{l}
-	\id{increment}_{P,e_1,e,\Delta}\la\Events, \id{prec}, \delta, \iota\ra\\
-	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
+\begin{array}{l@{}l}
+	\multicolumn{2}{l}{\id{increment}_{P,e_1,e,\Delta}\la\Events, \id{prec}, \delta, \iota\ra}\\
+	\ \ =\ &\begin{array}[t]{@{}l@{\hspace{0.2cm}}c@{\hspace{0.2cm}}ll}
 		\text{Let }\delta'(x,y)&=&\Delta\text{, }&\text{if }x=e_1\wedge y=e\wedge x\in P\\
 				       &=&(\id{state}(e_1)\oplus\Delta)\ominus\id{state}(x)\text{, }&\text{if }x\neq e_1\wedge y=e\wedge x\in P\\
-				       &=&\delta(x,y)\text{, }&\text{if }(x,y)\in\id{prec}\\
-		\text{Let }\iota'(x)&=&\Delta\text{, }&\text{if }x=e\wedge e_1\notin P\\
-				    &=&\iota(x)\text{, }&\text{if }x\neq e\wedge e_1\notin P\\
-			\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \delta', \iota'\}\ra}
-		\end{array}
+				       &=&\delta(x,y)\text{, }&\text{if }(x,y)\in\id{prec}
+		\end{array}\\
+		&\begin{array}[t]{@{}l@{\hspace{0.2cm}}c@{\hspace{0.2cm}}ll}
+			\text{Let }\iota'(x)&=&\Delta\text{, }&\text{if }x=e\wedge P=\emptyset\\
+				    &=&\iota(x)\text{, }&\text{if }x\neq e\wedge P=\emptyset
+		\end{array}\\
+		&\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \delta', \iota'\}\ra
 \end{array}
@@
 
+	We must prove\marginpar{claim} that
+\begin{equation}
+	\id{ihs2shs}(\id{increment}_{P,e_1,e,\Delta}(\id{shs2ihs}(h)))\ =\ \id{occurs}_{P,e,\id{state}(e_1)\oplus\Delta}(h)
+\end{equation}
+
+\section{Isolation}
+\label{sct:Isolation}
+	How does a user perceive a distributed information system (DIS)?
+	There are different options.
+	An example is an information screen with flight arrival times in an airport,
+	where a user sees the current state of flight arrivals and sees it change each time a new event occurs.
+	This is an example of non-transactional use: the user consumes information passively.
+	However, the topic of this paper is transactional use,
+	an example of which is to make a flight reservation.
+	In that example, the user interacts with the system for the purpose of reserving a seat on a flight.
+	She sees a current state and provides events to the system by filling out fields in a reservation screen.
+	These events cause changes in the state (e.g.\ from {\em The flight is not paid} to {\em The flight has been paid}).
+	With this type of use, i.e transactional use, the user expects a predictable result.
+	That is: if the current state is $s$ and the user specifies a change $\Delta$,
+	the resulting state must be $s\oplus\Delta$.
+	This requirement is the incremental variant of requirement~\ref{req:transaction}:
+\begin{equation}
+	h'=\id{occurByDelta}_{P,e_1,\Delta,e}(h)\ \Rightarrow\ \id{state}(e)\ =\ \id{state}(e_1)\oplus\Delta
+\label{req:occurByDelta}
+\end{equation}
+	Since $\id{occurByDelta}$ preserves a stateful history structure (equation~\ref{req:recursive def SHS by Delta}),
+	the user can interact repeatedly with the system and update the state predictably in a sequence of changes.
+	This yields a user experience of being the only user of the system.
+	That property is called {\em isolation}.
+
+\section{Consistency}
+\label{sct:Consistency}
+	From a user perspective, {\em consistency} means that the system is free of contradictions.%
+\footnote{A user might have wanted the system to provide the truth itself,
+	but that is beyond what technical systems can guarantee. For people can feed the system with lies.
+	However, freedom of contradictions is quite desirable as second-best, because truth is free of contradictions.}
+	We require a predicate on states to determine whether a state is consistent.
+	So we extend the notion of stateful historic structure with that predicate.
+
+	A {\em nonconflicting history structure} is a quadruple $\la\Events, \id{prec}, \id{state}, \id{consistent}\ra$ in which
+	$\la \Events, \id{prec}, \id{state}\ra$ is a SHS and $\id{consistent}$ is a predicate on $\mathbb S$,
+	which holds for every state in the SHS.
+	We will use predicate $\id{NHS}$ as a definition:
+\begin{equation}
+\begin{array}{l@{}ll}
+	\multicolumn{3}{l}{\id{NHS}\la \Events, \id{prec}, \id{state}, \id{consistent}\ra}\\
+	\ \Leftrightarrow\ &\id{SHS}\la \Events, \id{prec}, \id{state}\ra&\wedge\\
+	&\id{consistent}\subseteq\mathbb S&\wedge\\
+	&\forall e\in\Events:\ \id{consistent}(\id{state}(e))&
+\end{array}
+\label{def:NHS}
+\end{equation}
+	Equation~\ref{def:NHS} defines a nonconflicting history structure:
+	For practical use, we require $\id{consistent}$ to be decidable.
+
 \section{Atomicity}
@@
 	{\em Atomicity} guarantees all-or-nothing outcomes.
+	Let an example clarify this.
+	If I purchase a song on the internet, I must pay.
+	As a result, it is acceptable that the payment is done and the song is delivered.
+	It is also acceptable that the payment is not done and the song is not delivered.
+	However, it is unacceptable that the payment is done and the song is not delivered or vice-versa.
+
+	Consider this:
+	\[h'\ =\ \id{increment}_{P,e_2,e_3,\Delta_2}(\id{increment}_{P,e_1,e_2,\Delta_1}(h))\]
+	Imagine that $\Delta_1$ represents my payment and $\Delta_2$ represents the delivery of my song.
+	In the corresponding history structure, we will see that
+	\[\{(e_1,e_2), (e_2,e_3)\}\subseteq\id{prec} \]
 	It means that the overall state change caused by a sequence of events can be committed as a whole or rejected completely.
@@
 
+\section{theorems}
+\label{sct:theorems}
+	An order theoretic meet-semilattice $\la S, \leq\ra$ gives rise to a binary operation $\oplus$ such that $\la S, \oplus\ra$ is an algebraic meet-semilattice.
+	With $\preceq$ being reflexive, antisymmetric, and transitive,
+	it follows immediately that $\oplus$ is idempotent, commutative, and associative.
 \section{Bibliography}
@@
 
-\section{Consistency}
-\label{sct:Consistency}
-	{\em Consistency} means that integrity rules are warranted by the database management system.
-
-	If $s$ is a state and $r$ is an integrity rule, then $s\ \id{sat}\ r$ means that $s$ satisfies rule $r$
-	and $s\ \overline{\id{sat}}\ r$ means that $s$ does not satisfy rule $r$.
-
-	Let us call the set of integrity rules $\mathcal R$.
-	A state $s$ is called consistent if it satisfies every integrity rule $r\in\mathcal R$
-\begin{equation}
-	\id{consistent}(s)\ =\ \forall r\in\mathcal R: s\ \id{sat}\ r
-\end{equation}
-	The system is consistent if the initial state is consistent and it can only reach consistent states by means of atomic actions.
-	So all states in the history graph are consistent with $\mathcal R$.
-
 \section{Isolation}
@@
 	The order (in time) of events is given by a relation\footnote{%
-A relation $\declare{r}{A}{B}$ is a set of tuples. Each tuple $\pair{a}{b}$ in that set satisfies $a\in A$ and $b\in B$}
+A relation $\declare{r}{A}{B}$ is a set of pairs. Each pair $\pair{a}{b}$ in that set satisfies $a\in A$ and $b\in B$}
 	$\declare{pred}{Event}{Event}$,

### 2019-03-18 2018 Transactions/articleTransactions.tex
I have made things consistent with "the state of my mind".

@@
 
-\section{Growth}
-\label{sct:Growth}
+	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
+	Let us represent the fact that $h$ is a {\em historic substructure} of $h'$ by:
+\begin{equation}
+\label{def:historic substructure}
+	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra\ \Leftrightarrow\ \Events\subseteq\Events'\ \wedge\ \id{prec}\subseteq\id{prec}'
+\end{equation}
+
 	A DIS grows by adding events to it.
-	We describe each time an event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one history structure to another:
+	We describe each time a new event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one history structure to another:
 \begin{equation}
@@
 
-	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
-	Let us represent the fact that $h$ is a {\em historic substructure} of $h'$ by:
-\begin{equation}
-\label{def:historic substructure}
-	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra\ \Leftrightarrow\ \Events\subseteq\Events'\ \wedge\ \id{prec}\subseteq\id{prec}'
-\end{equation}
-	From lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.
+	With lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.
 
@@
 
-	Let us reconsider the relation $\id{prec}$ for a moment and define $\preceq$ as follows:
-	Requirement~\ref{req:precPlus asy} from section~\ref{sct:History} ensures that $\preceq$ is reflexive, antisymmetric, and transitive.
-	Let us introduce operator $\curlyvee$ (the {\em meet} of two events)
-\begin{equation}
-	e\ =\ e_1\curlyvee e_2\ \ \Leftrightarrow\ \ e\ \preceq\ e_1\ \wedge\ e\ \preceq\ e_2
-\end{equation}
-	If $\la \Events, \id{prec}, \id{state}\ra$ satisfies definition~\ref{req:recursive def occurs},
-	we can prove
-\begin{equation}
-		e_1,e_2\in\Events\ \Rightarrow\ (e_1\curlyvee e_2)\in\Events
-\end{equation}
-	This means that $\la\Events,\preceq\ra$ is an order-theoretic meet-semilattice.
-	From order theory, we know that $\curlyvee$ is idempotent, commutative, and associative.
-
-	Having defined history structures and the way they grow, let us proceed to determine requirements for states.
-
 \section{States}
@@
 	For that purpose we introduce the notion of state.
+	Let $\mathbb S$ represent the set of all conceivable states.
 
-	A history structure can be perceived as the (global) state of an information system at any point in time,
-	even though that global state is observable only in its local parts.
+	This section expands history structures to stateful history substructures
+	to represent the (global) state of an information system at any point in time.
+	Since this paper is about distributed systems,
+	we may not assume that the global state of a DIS at some point in time is observable.
+	The state of a DIS is observable only in its local parts.
+	So we shall define multiple states, each of which may be interpreted locally.
 
-	Let $\mathbb S$ represent the set of all conceivable states.
-	A {\em stateful history structure} is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
+	We define a {\em stateful history structure} as a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
 	$\la \Events, \id{prec}\ra$ is a history structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
@@
 	Or shorter: $s$ is the state after event $e$.
-	Instead of $s=\id{state}(e)$, we may also write $(s,e)\in\id{state}$ because a function is a univalent and total relation
-	(and therefore a set of pairs).
+%	Instead of $s=\id{state}(e)$, we may also write $(s,e)\in\id{state}$ because a function is a univalent and total relation
+%	(and therefore a set of pairs).
 	A {\em current state} is the state after a current event
@@
 \begin{array}{l}
-	\id{occurs}_{P,e,s}(\la \Events, \id{prec}, \id{state}\ra)\\
-	\ =\ \la \Events\cup\{e\},\ \id{prec}\cup\{(e_1,e) | e_1\in P\},\ \id{state}\cup\{(e,s)\}\ra
+	P\subseteq\Events\ \wedge\ e\notin E\\
+	\ \ \Rightarrow\ \ 
+\begin{array}[t]{l}
+	\id{occurs}_{P,e,s}(\la\Events, \id{prec}, \id{state}\ra)\\
+	\ \ =\ \begin{array}[t]{@{}l@{\hspace{.2em}}c@{\hspace{.2em}}ll}
+		\text{Let }\id{newState}(x)&=&s\text{,}&\text{if }x=e\\
+				&=&\id{state}(x)\text{,}&\text{otherwise}\\
+		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{newState}\}\ra}
 		\end{array}
-\label{eqn:def occurs}
+\end{array}
+\end{array}
+\label{req:def occurs}
 \end{equation}
@@
 \end{equation}
-	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
-	This difference represents a change.
+	Let us call $s_2\ominus s_1$ the difference ($\Delta$) between states $s_2$ and $s_1$.
 	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.
-	In a stateful history structure we can represent transactions by means of a function $\delta$,
-	where $\delta(e_1,e_2)$ represents the change of state from event $e_1$ to $e_2$.
-\begin{equation}
-	\delta(e_1,e_2)\ =\ \id{state}(e_2)\ominus\id{state}(e_1)
-\label{def:delta}
-\end{equation}
-	A transaction caused by event $e$ can be described by:
-\begin{eqnarray}
-&&	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
-		\id{state}(e)\ =\ s&\wedge\\
-		\forall e_1\in P: s\ =\ \id{state}(e_1)\oplus\delta(e_1,e)
-	\end{array}
-\label{req:transaction}
-\end{eqnarray}
+
 	We can define a stateful history structure,
@@
 	\id{occurByDelta}_{P,e_1,\Delta,e}(\la\Events, \id{prec}, \id{state}\ra)\\
-	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
+	\ \ =\ \begin{array}[t]{@{}l@{\hspace{.2em}}c@{\hspace{.2em}}ll}
 		\text{Let }s_1&=&\id{state}(e_1)\text{,}&\text{if }e_1\in P\\
 				&=&\mathcal I\text{,}&\text{otherwise}\\
-		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{state}\cup\{(e,s_1\oplus\Delta)\}\ra}
+		\text{Let }\id{newState}(x)&=&s_1\text{,}&\text{if }x=e\\
+		&=&\id{state}(x)\text{,}&\text{otherwise}\\
+		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{newState}\}\ra}
 		\end{array}
@@
 \item	$\la \Events, \id{prec}\ra$ is a history structure;
-\item	$\delta(e_1,e_2)$ represents a change, $\id{state}(e_2)\ominus\id{state}(e_1)$, for every pair $(e_1,e_2)\in\id{prec}$ (equation~\ref{def:delta});
+\item	$\delta(e_1,e_2)$ represents a change, $\id{state}(e_2)\ominus\id{state}(e_1)$, for every pair $(e_1,e_2)\in\id{prec}$;
 \item	$\iota(e)$ represents the change $\id{state}(e_1)\ominus\mathcal I$ for every event $e\in\Events$ that has no predecessor in $\id{prec}$.
@@
 
-\section{theorems}
-\label{sct:theorems}
-	An order theoretic meet-semilattice $\la S, \leq\ra$ gives rise to a binary operation $\oplus$ such that $\la S, \oplus\ra$ is an algebraic meet-semilattice.
-	With $\preceq$ being reflexive, antisymmetric, and transitive,
-	it follows immediately that $\oplus$ is idempotent, commutative, and associative.
+\section{Durability}
+\label{sct:Durability}
+	In the eyes of the user, durability means that history is not changed.
+	In the previous definitions durability has been secured by defining every change as an expansion of an existing system.
+	Events are added only. Once added they are never removed or replaced.
+	This corresponds to working practices in financial ledgers.
+	Transactions are never removed, only compensated.
+
 \section{Bibliography}

### 2019-03-29 2018 Transactions/articleTransactions.tex
The storyt starts with Foster

@@
 	{\em Durability} means that the contents of a database cannot change unintendedly.
-	To motivate this theory, let us formulate requirements to characterize a DIS:
 
@@
 \label{sct:Related work}
+	In 2017, Simon Foster et al. published a theory of time with generalised reactive processes~\cite{DBLP:journals/ipl/FosterCWZ18}.
+	It provides a generic foundation for denotational semantics of concurrent languages,
+	the proof obligation of which are satisfied by automated verification~\cite{Foster14}.
+	Their theory describes program behaviour as relational predicates whose variables correspond to observable quantities.
+	It draws on the seminal work of Hoare on communicating sequential processes~\cite{HoareHe98},
+	work on process algebras by Bergstra and Klop~\cite{DBLP:journals/iandc/BergstraK84}, and
+	Robin Milner's work on CCS~\cite{Milner1989}.
+	We chose to build our theory on the work of Foster et al. because it provides three healthiness conditions
+	that theories of time must adhere to,
+	ready for us to adopt.
+	In the same style we have tried to keep our theory down to the bare minimum, to make it as generally applicable as possible.
+	We too have used Isabelle to verify the required proofs,
+	to increase our confidence in this theory.
+
 \cite{Foster17c}
-
+	In the nineties, scientists saw that category theory can be used to formalize database schemas~\cite{Frederiks1997},
+	which was advocated as a means to further foundational theories of information systems.
 	An interesting read is~\cite{Foster17c}\footnote{\tt\tiny https://towardsdatascience.com/distributed-transactions-and-why-you-should-care-116b6da8d72}.
@@
 	so time is the first aspect we focus on.
+	Being distributed means that we cannot assume that the temporal order of events is known in all cases.
+	For this reason we draw on a theory of concurrency, which is generally applicable,
+	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
 	In this paper we simplify the notion of time to precedence of events,
@@
 	Let $\mathbb H$ be the set of all conceivable history structures.
+	We describe the history structure of a DIS as a partial order of events
+
 	A {\em history structure} is a tuple $\la \Events, \id{prec}\ra$ in which
@@
 \end{eqnarray}
-	Let set $\Events$ represent all events that the DIS has registered up to some point in time.
-	So for every event $e\in\Events$ we can say that it has happened (at some time in the past).
+	In a history structure $\la \Events, \id{prec}\ra$ set $\Events$ represents the events that have been registered up to some point in time.
+	For every event $e\in\Events$ we can say that it has happened (at some time in the past).
 	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
@@
 	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
-	Let us represent the fact that $h$ is a {\em historic substructure} of $h'$ by:
+	We call this a {\em historic substructure}, which is defined by:
 \begin{equation}

### 2019-04-03 2018 Transactions/articleTransactions.tex
WIP

@@
 	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
-	These so-called NoSQL databases often lack functionality for transactions.
-	For the purpose of doing transactions in a non-transactional situation, several architectural patterns exist.
+	These so-called NoSQL databases often lack functionality for transactions,
+	leaving the problems for the user to solve.
+	As a result, many architectural patterns exist for the purpose of doing transactions in a distributed setting.
 	Some of these patterns simply circumvent the problem.
@@
 	Still others introduce locking mechanisms to do the job.
-	Yet, in the world of event-driven distributed systems, a dire need is felt for a theory that brings what transactions do for shared databases.
+	In paper we will show that the ACID-properties of transactions follow quite naturally from the theory of time.
+	In informal terms: the merge of Git corresponds to the transactional commit in databases.
+	This result contributes to the body of foundational theory of information systems.
 
-	This paper presents such a theory.
-	It describes transactions without the assumption of a shared database.
-	In this theory we express the characteristic properties of transactions: Atomicity, Consistency, Isolation, and Durability.
+	This paper describes transactions without the assumption of a shared database.
+	In we express the characteristic properties of transactions: Atomicity, Consistency, Isolation, and Durability.
 	{\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
@@
 	{\em Durability} means that the contents of a database cannot change unintendedly.
+	For this purpose we build on a well-known theory of time~\cite{DBLP:journals/ipl/FosterCWZ18},
+	which is generally applicable,
+	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
 
@@
 	Being distributed means that we cannot assume that the temporal order of events is known in all cases.
-	For this reason we draw on a theory of concurrency, which is generally applicable,
-	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
 	In this paper we simplify the notion of time to precedence of events,
